---
title: "microbiome by habitat"
author: "Dani Blumstein"
date: "2025-12-30"
output: html_document
---

DB to change east west lable to region!
```{r}
#if (!requireNamespace("BiocManager", quietly=TRUE))
#    install.packages("BiocManager")
## BiocManager::install("BiocUpgrade") ## you may need this
#BiocManager::install("MicrobiotaProcess",force = TRUE)
#BiocManager::install("phyloseq",force = TRUE)
#BiocManager::install("ggtree", force = TRUE)
#BiocManager::install("tidytree", force = TRUE, type = "source")

#install.packages("remotes")
#remotes::install_github("YuLab-SMU/ggtree")
#remotes::install_github("YuLab-SMU/MicrobiotaProcess")



#install.packages(
#  "microViz",
#  repos = c(davidbarnett = "https://david-barnett.r-universe.dev", getOption("repos"))
#)

#if (!requireNamespace("devtools", quietly = TRUE)){install.packages("devtools")}
#devtools::install_github("jbisanz/qiime2R")
#devtools::install_github("zpneal/backbone", ref = "devel", build_vignettes = TRUE)

suppressPackageStartupMessages({
    library(MicrobiotaProcess) 
    library(phyloseq)
    library(ggplot2)
    library(tidyverse) 
    library(vegan) 
    library(coin) 
    library(reshape2) 
    library(ggnewscale)
    library(qiime2R)
    library(readxl)
    library(dplyr)
    library(microViz)
    library(UpSetR)
    library(patchwork)
    library(ggtree)
    library(tidytree)
    library(phylobase)
    library(ggtreeExtra)
    library(sf)
    library(phytools)
    library(SummarizedExperiment)
    library(remotes)
    library(terra)
    library(ggspatial)
    library(stringr)
    library(purrr)
    library(gridExtra)
})
```

```{r}
habitat_colors <- c(
  # East of Mississippi (cool / forested)
  "Deciduous Forest"     = "#0072B2",  # true blue
  "Mixed Forest"         = "#2CA25F",  # teal-green
  
  # West of Mississippi (warm / dry / managed)
  "Cultivated Crops"     = "#E69F00",  # orange (CVD-safe, high salience)
  "Evergreen Forest"     = "#A0522D",  # 
  "Grassland/Herbaceous" = "#F0E442",  # yellow (excellent contrast on maps)
  "Shrub/Scrub"          = "#D55E00"   # vermillion (distinct from crops)
)
  
east_west_colors <- c(
  "east" = "#6B8FB3",  # muted steel blue
  "west" = "#B07C59"   # muted warm tan
)

```

```{r}
mpse_samples <- mp_import_qiime2(otuqza="data/microbiome/rarefied-table.qza", 
                          taxaqza="data/microbiome/taxonomy.qza", 
                          mapfilename="data/microbiome/metadata_edit.tsv")
mpse_samples

#Set labtracking column as the sample names
new_names <- colData(mpse_samples)$labtracking

# Make sure they are unique
stopifnot(!anyDuplicated(new_names))

# Assign as sample names
colnames(mpse_samples) <- new_names
rownames(colData(mpse_samples)) <- new_names

#filter to only pema and drop AZ
mpse_samples_filtered <- mpse_samples %>%
  filter(specificEpithet == "maniculatus") %>%
  filter(stateProvince != "Arizona")

asv_table <- assay(mpse_samples_filtered, "Abundance")
```

```{r}
#use the gps points and USGS national land coverage data to assign habitat
point <- as.data.frame(colData(mpse_samples_filtered))
point$decimalLatitude <- as.numeric(point$decimalLatitude)
point$decimalLongitude <- as.numeric(point$decimalLongitude)

p1 <- vect(point, geom = c("decimalLongitude", "decimalLatitude"), crs = "+proj=longlat +datum=WGS84")

#Convert to sf
points_sf <- st_as_sf(p1)  

#raster data from https://www.mrlc.gov/data
raster_obj <- rast("Annual_NLCD_LndCov_2020_CU_C1V1/Annual_NLCD_LndCov_2020_CU_C1V1.tif")

#raster data from https://prism.oregonstate.edu/data/
rain_raster <- rast("prism_ppt_us_30s_2020/prism_ppt_us_30s_2020.tif")
mean_temp_raster <- rast("prism_tmean_us_30s_2020/prism_tmean_us_30s_2020.tif")

plot(raster_obj)
plot(rain_raster)
plot(mean_temp_raster)

pts_sf <- st_transform(points_sf, crs(raster_obj))
pst_sf_rain <- st_transform(points_sf, crs(rain_raster))
pts_sf_temp <- st_transform(points_sf, crs(mean_temp_raster))

# Convert sf → terra vector
pts_vect <- vect(pts_sf)
pts_vect_rain <- vect(pst_sf_rain)
pts_vect_temp <- vect(pts_sf_temp)

# Extract raster values
vals <- extract(raster_obj, pts_vect)
vals_rain <- extract(rain_raster, pts_vect_rain)
vals_temp <- extract(mean_temp_raster, pts_vect_temp)

# Combine with original data
results <- pts_sf |>
  bind_cols(
    land_use = vals[,2]
  )

results$land_use_location <- paste(results$land_use, results$Mississippi_River_Label, sep = " ")
#write_tsv(results, file = "metadata_edit2.tsv")

colData(mpse_samples_filtered)$land_use_location <- results$land_use_location
colData(mpse_samples_filtered)$land_use <- results$land_use

results <- pst_sf_rain |>
  bind_cols(
    prism_ppt_us_30s_2020 = vals_rain[,2]
  )
colData(mpse_samples_filtered)$rain <- results$prism_ppt_us_30s_2020

results <- pts_sf_temp |>
  bind_cols(
    prism_tmean_us_30s_2020 = vals_temp[,2]
  )
colData(mpse_samples_filtered)$temp <- results$prism_tmean_us_30s_2020

# Extract sample data from MPSE object
sample_data_df <- colData(mpse_samples_filtered)

# Export as tab-delimited text file
write.table(sample_data_df, 
            file = "MicrobiotaProcess_sample_data.txt", 
            sep = "\t", 
            quote = FALSE, 
            row.names = TRUE)

```

#sort the metadata and the otu table
metadata_filtered_sorted <- metadata_filtered[order(metadata_filtered$labtracking),]
otu_sorted <- otu[order(rownames(otu)),]

#check the data for order 
unique(rownames(otu_sorted) == metadata_filtered_sorted$labtracking) # OK

based off gabri code

### Calculate alpha diversity
metadata_filtered_sorted$Richness.rar<-specnumber(otu_sorted) # Counts otu number
metadata_filtered_sorted$Shannon.rar<-vegan::diversity(otu_sorted, index="shannon") ## calculate Shannon H

a = ggplot(metadata_filtered_sorted, aes(x = land_use_location, y = Richness.rar, color = stateProvince)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), alpha = 0.7) +
  theme_classic() +
  labs(y = "Number of otus", x = "Mississippi River Split") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
a



### Beta diversity 
##### Beta diversity ----------------
set.seed(12)
bac.dist<-vegdist(otu_sorted, method="bray") ### this calculates BC dissimilaries
bac.dist.sor<-vegdist(otu_sorted, method="bray", binary  = TRUE ) ### this calculates Sorensen dissimilaries

set.seed(12)
bac.nmds<-metaMDS(bac.dist, k=2, try=999)
bac.nmds$stress #0.21
stress= paste("stress = ", round(bac.nmds$stress, digits = 2))

metadata_filtered$Axis01<-bac.nmds$points[,1]
metadata_filtered$Axis02<-bac.nmds$points[,2]

# install.packages("viridis") if needed
bac_nmds <- ggplot(metadata_filtered, aes(x = Axis01, y = Axis02)) +
  geom_point(aes(color = stateProvince, shape = land_use_location), size = 2, alpha = 0.8) +
  stat_ellipse(aes(fill = land_use_location), geom = "polygon", alpha = 0.45, level = 0.68, show.legend = FALSE) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  scale_fill_viridis_d(option  = "turbo", begin = 0.05, end = 0.95) +
  theme_classic() +
  labs(x = "NMDS1", y = "NMDS2", shape = "land_use_location") +
  ggtitle("Bray Curtis dissimilarity")
bac_nmds

#perm = adonis2(bac.dist ~ land_use_location + stateProvince, data = metadata_filtered, by = "margin")
#perm

vignette used for code chunks below: https://bioconductor.org/packages/release/bioc/vignettes/MicrobiotaProcess/inst/doc//MicrobiotaProcess.html


Rarefied species richness
```{r}
mpse_samples_filtered %<>% mp_rrarefy()

mpse_samples_filtered %<>% 
    mp_cal_rarecurve(
        .abundance = RareAbundance,
        chunks = 400
    )
# default will display the confidence interval around smooth.
# se=TRUE
p1 <- mpse_samples_filtered %>% 
      mp_plot_rarecurve(
         .rare = RareAbundanceRarecurve, 
         .alpha = Observe,
      )

p2 <- mpse_samples_filtered %>% 
      mp_plot_rarecurve(
         .rare = RareAbundanceRarecurve, 
         .alpha = Observe, 
         .group = land_use_location)
      #) +
      #scale_color_manual(values=c("#00A087FF", "#3C5488FF")) +
      #scale_fill_manual(values=c("#00A087FF", "#3C5488FF"), guide="none")

# combine the samples belong to the same groups
p3 <- mpse_samples_filtered %>% 
      mp_plot_rarecurve(
         .rare = RareAbundanceRarecurve, 
         .alpha = "Observe", 
         .group = land_use_location, 
         plot.group = TRUE)
      #) +
       #scale_color_manual(values=c("#00A087FF", "#3C5488FF")) +
       #scale_fill_manual(values=c("#00A087FF", "#3C5488FF"),guide="none")  

p1 
p2
p3
```

calculate alpha index and visualization
```{r}
mpse_samples_filtered %<>% 
    mp_cal_alpha(.abundance=RareAbundance)

f1 <- mpse_samples_filtered %>% 
      mp_plot_alpha(
        .group=land_use_location, 
        .alpha=c(Observe, Chao1, ACE, Shannon, Simpson, Pielou))
      #) +
      #scale_fill_manual(values=c("#00A087FF", "#3C5488FF"), guide="none") +
      #scale_color_manual(values=c("#00A087FF", "#3C5488FF"), guide="none")

f2 <- mpse_samples_filtered %>%
      mp_plot_alpha(
        .alpha=c(Observe, Chao1, ACE, Shannon, Simpson, Pielou)
      )

f1 / f2
```

visualization of taxonomy abundance
```{r}
mpse_samples_filtered %<>%
    mp_cal_abundance( # for each samples
      .abundance = RareAbundance
    ) %>%
    mp_cal_abundance( # for each groups 
      .abundance=RareAbundance,
      .group=land_use_location
    )

# visualize the relative abundance of top 20 phyla for each sample.
p1 <- mpse_samples_filtered %>%
         mp_plot_abundance(
           .abundance=RareAbundance,
           .group=land_use_location, 
           taxa.class = Phylum, 
           topn = 20,
           relative = TRUE
         )
# visualize the abundance (rarefied) of top 20 phyla for each sample.
p2 <- mpse_samples_filtered %>%
          mp_plot_abundance(
            .abundance=RareAbundance,
            .group=land_use_location,
            taxa.class = Phylum,
            topn = 20,
            relative = FALSE
          )
p1
p2


h1 <- mpse_samples_filtered %>%
         mp_plot_abundance(
           .abundance = RareAbundance,
           .group = land_use_location,
           taxa.class = Phylum,
           relative = TRUE,
           topn = 20,
           geom = 'heatmap',
           features.dist = 'euclidean',
           features.hclust = 'average',
           sample.dist = 'bray',
           sample.hclust = 'average'
         )

h2 <- mpse_samples_filtered %>%
          mp_plot_abundance(
            .abundance = RareAbundance,
            .group = land_use_location,
            taxa.class = Phylum,
            relative = FALSE,
            topn = 20,
            geom = 'heatmap',
            features.dist = 'euclidean',
            features.hclust = 'average',
            sample.dist = 'bray',
            sample.hclust = 'average'
          )
# the character (scale or theme) of figure can be adjusted by set_scale_theme
# refer to the mp_plot_dist
#aplot::plot_list(gglist=list(h1, h2), tag_levels="A")
h1
h2

# visualize the relative abundance of top 20 phyla for each .group (time)
p3 <- mpse_samples_filtered %>%
         mp_plot_abundance(
            .abundance=RareAbundance, 
            .group=land_use_location,
            taxa.class = Phylum,
            topn = 20,
            plot.group = TRUE
          )

# visualize the abundance of top 20 phyla for each .group (time)
p4 <- mpse_samples_filtered %>%
          mp_plot_abundance(
             .abundance=RareAbundance,
             .group= land_use_location,
             taxa.class = Phylum,
             topn = 20,
             relative = FALSE,
             plot.group = TRUE
           )
p3
p4
```

Beta diversity analysis
```{r}
mpse_samples_filtered %<>% 
    mp_decostand(.abundance=Abundance)

mpse_samples_filtered %<>% mp_cal_dist(.abundance=hellinger, distmethod="bray")

# when .group is provided, the dot heatmap plot with group information will be return.

p2 <- mpse_samples_filtered %>% mp_plot_dist(.distmethod = bray, .group = land_use_location)
p2
# The scale or theme of dot heatmap plot can be adjusted using set_scale_theme function.
#p2 %>% set_scale_theme(
#          x = scale_fill_manual(
#                 values=c("orange", "deepskyblue"), 
#                 guide = guide_legend(
#                             keywidth = 1, 
#                             keyheight = 0.5, 
#                             title.theme = element_text(size=8),
#                             label.theme = element_text(size=6)
#                 )
#              ), 
#          aes_var = land_use_location # specific the name of variable 
#       ) %>%
#       set_scale_theme(
#          x = scale_color_gradient(
#                 guide = guide_legend(keywidth = 0.5, keyheight = 0.5)
#              ),
#         aes_var = bray
#      ) %>%
#       set_scale_theme(
#          x = scale_size_continuous(
#                 range = c(0.1, 3),
#                 guide = guide_legend(keywidth = 0.5, keyheight = 0.5)
#              ),
#          aes_var = bray
#       )

p3 <- mpse_samples_filtered %>% mp_plot_dist(.distmethod = bray, .group = land_use_location, group.test=T, textsize=2)
p3 
```

pcoa
```{r}
mpse_samples_filtered %<>% 
    mp_cal_pcoa(.abundance=hellinger, distmethod="bray")

#perform adonis or anosim to check whether it is significant to the dissimilarities of groups.
mpse_samples_filtered %<>%
     mp_adonis(.abundance=hellinger, .formula=~Mississippi_River_Label, distmethod="bray", permutations=9999, action="add")

mpse_samples_filtered$Mississippi_River_Label <- 
  factor(
    mpse_samples_filtered$Mississippi_River_Label,
    levels = c("east", "west")
  )

mpse_samples_filtered %>% mp_extract_internal_attr(name=adonis)

p1 <- mpse_samples_filtered %>%
        mp_plot_ord(
          .ord = pcoa, 
          .group = land_use, 
          .color = land_use, 
          .starshape = Mississippi_River_Label,
          .size = 1.2,
          .alpha = 1,
          ellipse = TRUE,
          show.legend = F # don't display the legend of stat_ellipse
        )+   # land-use colors (primary)
  scale_color_manual(
    name   = "Locataion",
    values = habitat_colors
  ) +
  scale_fill_manual(
    name   = "Land use",
    values = habitat_colors
  ) 

p1

# The size of point also can be mapped to other variables such as Observe, or Shannon 
# Then the alpha diversity and beta diversity will be displayed simultaneously.
p2 <- mpse_samples_filtered %>% 
        mp_plot_ord(
          .ord = pcoa, 
          .group = land_use, 
          .color = land_use, 
          .size = Observe, 
          .alpha = Shannon,
          .starshape = Mississippi_River_Label,
          ellipse = TRUE,
          show.legend = FALSE # don't display the legend of stat_ellipse 
        ) +
        scale_color_manual(
    name   = "Locataion",
    values = habitat_colors
  ) +
  scale_fill_manual(
    name   = "Land use",
    values = habitat_colors
  ) +
        scale_size_continuous(
          range=c(0.5, 3),
          guide = guide_legend(keywidth=0.6, keyheight=0.6, label.theme=element_text(size=6.5))
        )
p1 
p2
```

below this is whole microbiome community
___________________________________________________________________________
physical distance matrix using GPS
```{r}
gps <- as.data.frame(mpse_samples_filtered@colData)[, c("labtracking","stateProvince","land_use","decimalLatitude","decimalLongitude","rain","temp","Mississippi_River_Label")]

locations <- unique(gps$stateProvince)

####################
#   rain distance  #
####################
rain_dist <- dist(gps$rain)
# Set row/column names
rownames(rain_dist) <- gps$labtracking
colnames(rain_dist) <- gps$labtracking
rain_object <- as.dist(rain_dist)
# Perform hierarchical clustering 
rain_tree <- hclust(rain_object, method = "complete") 
# Plot the dendrogram
plot(rain_tree, main = "Rain Clustering Dendrogram")

####################
#   Temp distance  #
####################
temp_dist <- dist(gps$temp)
# Set row/column names
rownames(temp_dist) <- gps$labtracking
colnames(temp_dist) <- gps$labtracking
temp_object <- as.dist(temp_dist)
# Perform hierarchical clustering 
temp_tree <- hclust(temp_object, method = "complete") 
# Plot the dendrogram
plot(temp_tree, main = "Temp Clustering Dendrogram")


####################
# Physical distance#
####################
sf_points <- st_as_sf(
  gps,
  coords = c("decimalLongitude", "decimalLatitude"),
  crs = 4326  # WGS84
)

dist_matrix <- st_distance(sf_points)

# Set row/column names
rownames(dist_matrix) <- gps$labtracking
colnames(dist_matrix) <- gps$labtracking

dist_object <- as.dist(dist_matrix)

# Perform hierarchical clustering 
hc_tree <- hclust(dist_object, method = "complete") 
hc_phylo <- as.phylo(hc_tree)

# Create the expanded tree
expanded_tree <- hc_phylo
for(stateProvince in hc_phylo$tip.label) {
  # Find all samples for this stateProvince
  samples <- gps$labtracking[gps$stateProvince == stateProvince]
  
  if(length(samples) > 0) {
    # Find the tip position for this stateProvince
    tip_position <- which(expanded_tree$tip.label == stateProvince)
    
    if(length(samples) == 1) {
      # If only one sample, just rename the tip
      expanded_tree$tip.label[tip_position] <- samples[1]
    } else {
      # If multiple samples, we need to add them as a polytomy
      # First, rename the original tip to the first sample
      expanded_tree$tip.label[tip_position] <- samples[1]
      
      # Then add the rest as sister tips (with very small branch length)
      for(i in 2:length(samples)) {
        expanded_tree <- phytools::bind.tip(expanded_tree, 
                                            tip.label = samples[i],
                                            where = tip_position,
                                            position = 0.0001)  # tiny branch length
      }
    }
  }
}

plot(expanded_tree)

# Match tip labels to rows in gps
tip_index <- match(expanded_tree$tip.label, gps$labtracking)

# Get habitats for all tips (NA if no match)
tip_habitats <- as.character(gps$land_use[tip_index])

# Look up colors in habitat_colors
tip_colors <- habitat_colors[tip_habitats]

# Replace any missing colors with black
tip_colors[is.na(tip_colors)] <- "black"

#build tip color vector from your state/colors (must be same length as nrow(assoc))
gps$tip_colors <- habitat_colors[
  match(gps$land_use, names(habitat_colors))
]
names(gps$tip_colors) <- gps$labtracking

phy_tree <- as.phylo(expanded_tree)

# Plot the tree with colored tips
plot(phy_tree, 
     tip.color = tip_colors,
     cex = 0.6,
     no.margin = TRUE)


```

mantel test
```{r}
bc_dist   <- vegdist(t(asv_table), method = "bray") 
sor_dist  <- vegdist(t(asv_table), method = "bray", binary = TRUE) 

labs <- attr(bc_dist, "Labels")
bc_mat <- as.matrix(bc_dist)[labs, labs]

####################
#   temp distance  #
####################
sor_mat <- as.matrix(sor_dist)[labs, labs]
sor_mat <- sor_mat[rownames(temp_object), colnames(temp_object)]

unique(rownames(temp_object) == rownames(sor_mat)) # ok 
unique(colnames(temp_object) == colnames(sor_mat)) # ok

mt = mantel(sor_mat,temp_object, method = "spearman")
mt$statistic
mt$signif

#### remove same states 
dist_matrix_no_zero = temp_object
dist_matrix_no_zero[dist_matrix_no_zero < 1] = 0
mr = cultevo::mantel.test(dist(sor_mat),dist(dist_matrix_no_zero), omitzerodistances = TRUE, trials = 999)
mr


#### plot the mantel tests 
df <- data.frame(
  phy = as.vector(dist_matrix_no_zero),
  micro = as.vector(dist(sor_mat))
)

ggplot(df, aes(x = phy, y = micro)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  theme_classic() +
  labs(x = "temp distance", y = "Microbiome distance")

grid.arrange(p1, p2, ncol = 2)
####################
#   rain distance  #
####################
sor_mat <- as.matrix(sor_dist)[labs, labs]
sor_mat <- sor_mat[rownames(rain_object), colnames(rain_object)]

unique(rownames(rain_object) == rownames(sor_mat)) # ok 
unique(colnames(rain_object) == colnames(sor_mat)) # ok

mt = mantel(sor_mat,rain_object, method = "spearman")
mt$statistic
mt$signif

dist_object_no_zero = dist_object
mt = cultevo::mantel.test(dist(sor_mat),dist(dist_object_no_zero), omitzerodistances = TRUE, trials = 999)
mr


#### plot the mantel tests 
# Convert to long data frames for plotting
df_with_zeros <- data.frame(
  x = as.vector(dist_object),
  y = as.vector(sor_dist)
)

df_no_zeros <- data.frame(
  x = as.vector(dist_object_no_zero),
  y = as.vector(sor_dist)
)
df_no_zeros <- df_no_zeros[df_no_zeros$x > 0,]

# Plot with zeros
p1 <- ggplot(df_with_zeros, aes(x = x, y = y)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", col = "blue") +
  labs(
    x = "rain distance",
    y = "Microbiome distance")+
  theme_minimal()

# Plot without zeros
p2 <- ggplot(df_no_zeros, aes(x = x, y = y)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", col = "red") +
  labs(
    x = "rain distance no same state compairsons",
    y = "Microbiome distance"
  ) +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)

####################
# Physical distance#
####################
sor_mat <- as.matrix(sor_dist)[labs, labs]
sor_mat <- sor_mat[rownames(dist_object), colnames(dist_object)]

unique(rownames(dist_object) == rownames(sor_mat)) # ok 
unique(colnames(dist_object) == colnames(sor_mat)) # ok

mt = mantel(sor_mat,dist_object, method = "spearman")
mt$statistic
mt$signif

dist_object_no_zero = dist_object
mt = cultevo::mantel.test(dist(sor_mat),dist(dist_object_no_zero), omitzerodistances = TRUE, trials = 999)
mr


#### plot the mantel tests 
# Convert to long data frames for plotting
df_with_zeros <- data.frame(
  x = as.vector(dist_object),
  y = as.vector(sor_dist)
)

df_no_zeros <- data.frame(
  x = as.vector(dist_object_no_zero),
  y = as.vector(sor_dist)
)
df_no_zeros <- df_no_zeros[df_no_zeros$x > 0,]

# Plot with zeros
p1 <- ggplot(df_with_zeros, aes(x = x, y = y)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", col = "blue") +
  labs(
    x = "physical distance",
    y = "Microbiome distance")+
  theme_minimal()

# Plot without zeros
p2 <- ggplot(df_no_zeros, aes(x = x, y = y)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", col = "red") +
  labs(
    x = "physical distance no same state compairsons",
    y = "Microbiome distance"
  ) +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

#### Plot the tanglegram with the expanded tree --------------------------------
note: this phylosymbosis
```{r}
hc <- hclust(sor_dist, method = "average")
tree <- as.phylo(hc)

plot(tree,
     tip.color = tip_colors,
     cex = 0.6,
     no.margin = TRUE,
     #x.lim = c(0, max(node.depth.edgelength(tree)) * 0.1))
     use.edge.length = FALSE)


assoc <- as.matrix(data.frame(
  V1 = as.character(gps$labtracking),   # labels for tree_exp (tree1)
  V2 = as.character(gps$labtracking),   # labels for tree   (tree2)
  stringsAsFactors = FALSE
))

# 3) build link color vector from your state/colors (must be same length as nrow(assoc))
gps$link_cols <- habitat_colors[ match(gps$land_use, names(habitat_colors)) ]
link_cols <- as.character(gps$link_cols[ match(assoc[,1], gps$labtracking) ])

tree2 <- tree
tree2$edge.length <- NULL

### reate cophylo - object for tanglegram --------------------------------------
cophylo_obj <- cophylo(tree2, phy_tree,assoc)

plot(cophylo_obj,
     type = "phylogram",
     link.type = "curved",
     link.lty = "solid",
     link.lwd = 1.5,
     link.col = link_cols,
     ftype = "off",
     #cex = 0.6,
     mar = c(0.1, 0.1, 0.1, 0.1))

# assuming state_colors is defined as you pasted
df_legend <- data.frame(state = names(habitat_colors), color = unname(habitat_colors))
df_legend$state <- factor(df_legend$state, levels = names(habitat_colors))

ggplot(df_legend, aes(x = 1, y = state, fill = state)) +
  geom_tile(color = "white", height = 0.9, width = 0.6) +
  scale_fill_manual(values = habitat_colors) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  theme_minimal(base_size = 10) +
  theme(axis.title = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none") +
  labs(title = "habitat color legend")
```

```{r}
library(mapdata) 

smoll_map <- gps[,c("decimalLatitude", "decimalLongitude")]
smoll_map$decimalLatitude <- as.numeric(smoll_map$decimalLatitude)
smoll_map$decimalLongitude <- as.numeric(smoll_map$decimalLongitude)
smoll_map$link_cols <- as.factor(gps$link_cols)

# Create a palette
palette <- rainbow(length(levels(smoll_map$link_cols)))

# Map colors in tip order
tip_colors <- palette[as.numeric(smoll_map$link_cols[match(tree_clado$tip.label, gps$SampleID)])]

tree_clado <- tree
tree_clado$edge.length <- rep(1, length(tree$edge.length))

phymap <- phylo.to.map(
  tree_clado,
  smoll_map,
  plot = FALSE,
  database="worldHires",regions="USA",
  
)

plot(phymap,
     ftype="off",
     fsize=0.8, 
  lty="solid",
  map.bg="grey",
  colors=tip_colors,
  xlim=c(-130, -65),
     ylim=c(24, 50))

```

this is based on an aggregated speices otu table
___________________________________________________________________________
```{r}
Taxtab<-read_qza("data/microbiome/taxonomy.qza")$data %>% parse_taxonomy()

### Let's try aggrergating to species ---------------------
Taxtab$lowest_taxon <- apply(Taxtab, 1, function(x) tail(na.omit(x), 1))
otu_by_taxon <- as.data.frame(asv_table)
otu_by_taxon$taxon <- Taxtab[rownames(asv_table), "lowest_taxon"]
otu_by_taxon <- aggregate(. ~ taxon, data = otu_by_taxon, sum)
otu_by_taxon <- otu_by_taxon %>%
  column_to_rownames(var = "taxon")
nrow(otu_by_taxon) # ok
unique(colnames(otu_by_taxon) == gps$labtracking) # ok
otu_by_taxon = as.data.frame(otu_by_taxon)

summary(rownames(otu_by_taxon))
```

```{r}
### repeat tests
bc_dist_tax  <- vegdist(t(otu_by_taxon), method = "bray") 
sor_dist_tax  <- vegdist(t(otu_by_taxon), method = "bray", binary = TRUE) 

##### Mantel tests 
labs <- attr(bc_dist_tax, "Labels")
bc_mat <- as.matrix(bc_dist_tax)[labs, labs]
bc_mat <- bc_mat[rownames(dist_object), colnames(dist_object)]

sor_mat <- as.matrix(sor_dist_tax)[labs, labs]
sor_mat <- sor_mat[rownames(dist_object), colnames(dist_object)]

unique(rownames(dist_object) == rownames(sor_mat)) # ok 
unique(colnames(dist_object) == colnames(sor_mat)) # ok

mt_tax = mantel(sor_mat,dist_object, method = "spearman")
mt_tax$statistic
mt_tax$signif

#### remove same species 
dist_matrix_no_zero = dist_object
dist_matrix_no_zero[dist_matrix_no_zero < 1] = 0
mr = cultevo::mantel.test(dist(sor_mat),dist(dist_matrix_no_zero), omitzerodistances = TRUE, trials = 999)
mr

#### plot the mantel tests 
df <- data.frame(
  phy = as.vector(dist_matrix_no_zero),
  micro = as.vector(dist(sor_mat))
)

ggplot(df, aes(x = phy, y = micro)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  theme_classic() +
  labs(x = "Physical distance", y = "Microbiome distance")
```

codiversity attempt 
```{r}
#---------------------------
# Import data
#---------------------------
physeq <- qza_to_phyloseq(
  features  = "data/microbiome/rarefied-table.qza",
  tree      = "data/microbiome/rooted-tree.qza",
  taxonomy  = "data/microbiome/taxonomy.qza",
  metadata  = "MicrobiotaProcess_sample_data.tsv"
)

#---------------------------
# Sample filtering
#---------------------------
ps <- physeq %>%
  ps_filter(scientificName == "Peromyscus maniculatus") %>%
  ps_filter(stateProvince != "Arizona")

sample_names(ps) <- sample_data(ps)$labtracking

#---------------------------
# Taxonomic aggregation
#---------------------------
ps_Family <- tax_glom(ps, taxrank = "Genus", NArm = TRUE)

#---------------------------
# Clean taxa & samples
#---------------------------
ps_clean <- ps_Family %>%
  prune_taxa(taxa_sums(.) > 0, .) %>%
  prune_samples(sample_sums(.) > 0, .)

taxa_names(ps_clean) <- make.unique(taxa_names(ps_clean))


#---------------------------
# Prevalence filtering (≥5%)
#---------------------------
asv_mat <- as(otu_table(ps_clean), "matrix")
prev <- rowSums(asv_mat > 0)
keep <- prev >= 0.05 * ncol(asv_mat)
ps_clean <- prune_taxa(keep, ps_clean)

#---------------------------
# Split asvs by Phylum
#---------------------------
#tax_df <- as(tax_table(ps_clean), "matrix") %>%
#  as.data.frame() %>%
#  rownames_to_column("asv") %>%
#  select(asv, Phylum)

#asv_by_phylum <- split(tax_df$asv, tax_df$Phylum)

#---------------------------
# Split asvs by Family
#---------------------------
tax_df <- as(tax_table(ps_clean), "matrix") %>%
  as.data.frame() %>%
  rownames_to_column("asv") %>%
  select(asv, Family)

asv_by_Family <- split(tax_df$asv, tax_df$Family)

#---------------------------
# Build Left Trees (per phylum)
#---------------------------
#scale_left <- 1.5

#left_tree_plot_df <- map_dfr(names(asv_by_phylum), function(ph) {
#  asvs <- asv_by_phylum[[ph]]
#  ps_sub <- prune_taxa(asvs, ps_clean)
  
#  if (ntaxa(ps_sub) < 2) return(NULL) # skip empty / singleton trees
  
#  p <- ggtree(phy_tree(ps_sub)) + theme_tree2()
  
#  p$data %>%
#    mutate(
#      Phylum = ph,
#      x = x * scale_left  # scale left tree
#    )
#})


# Only keep phyla with trees
#valid_phyla <- unique(left_tree_plot_df$Phylum)

#left_tree_plot_df <- left_tree_plot_df %>%
#  filter(Family %in% valid_phyla)

#left_tip_df <- left_tree_plot_df %>%
#  filter(isTip) %>%
#  transmute(
#    asv = label,
#    x_left = x,
#    y_left = y,
#    Phylum
#  )


#---------------------------
# Build Left Trees (per Family)
#---------------------------
scale_left <- 1.5

left_tree_plot_df <- purrr::map_dfr(names(asv_by_Family), function(gen) {
  asvs <- asv_by_Family[[gen]]
  ps_sub <- prune_taxa(asvs, ps_clean)

  if (ntaxa(ps_sub) < 2) return(NULL)

  p <- ggtree(phy_tree(ps_sub)) + theme_tree2()

  p$data %>%
    mutate(
      Family = gen,
      x = x * scale_left
    )
})

valid_genera <- unique(left_tree_plot_df$Family)

left_tree_plot_df <- left_tree_plot_df %>%
  filter(Family %in% valid_genera)

left_tip_df <- left_tree_plot_df %>%
  filter(isTip) %>%
  transmute(
    asv = label,
    x_left = x,
    y_left = y,
    Family
  )


#---------------------------
# Build Right Tree (all samples)
#---------------------------
scale_right <- 1.0 # right tree width

p_right <- ggtree(phy_tree) +
  scale_x_reverse() +
  theme_tree2()

right_df <- p_right$data %>%
  filter(isTip) %>%
  transmute(
    labtracking = label,
    x_right = x,
    y_right = y
  )

right_tree_df <- p_right$data
max_x <- max(right_tree_df$x, na.rm = TRUE)
# flip tips to face left
right_tree_df <- right_tree_df %>%
  mutate(
    x = max(x, na.rm = TRUE) - x # mirror
  )

# normalize and scale
right_tree_df <- right_tree_df %>%
  mutate(
    x = (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE)),
    x = x * scale_right
  )

right_tip_df <- right_tree_df %>%
  filter(isTip) %>%
  transmute(
    labtracking = label,
    x_base = x,
    y_base = y
  )

#---------------------------
# Shift right trees per phylum
#---------------------------
#max_left_x_df <- left_tree_plot_df %>%
#  group_by(Phylum) %>%
#  summarize(max_left_x = max(x, na.rm = TRUE), .groups = "drop")

# replicate right tree for each valid phylum
#right_tree_df <- right_tree_df %>%
#  crossing(v = valid_phyla) %>%
#  left_join(max_left_x_df, by = "Phylum") %>%
#  mutate(x = x + max_left_x + 1) # shift right tree after left tree

#right_tip_df <- right_tip_df %>%
#  crossing(Phylum = valid_phyla) %>%
#  left_join(max_left_x_df, by = "Phylum") %>%
#  mutate(x_right = x_base + max_left_x + 1,
#         y_right = y_base) %>%
#  select(labtracking, x_right, y_right, Phylum)

#---------------------------
# Shift right trees per Family
#---------------------------
max_left_x_df <- left_tree_plot_df %>%
  group_by(Family) %>%
  summarize(max_left_x = max(x, na.rm = TRUE), .groups = "drop")

right_tree_df <- right_tree_df %>%
  crossing(Family = valid_genera) %>%
  left_join(max_left_x_df, by = "Family") %>%
  mutate(x = x + max_left_x + 1)

right_tip_df <- right_tip_df %>%
  crossing(Family = valid_genera) %>%
  left_join(max_left_x_df, by = "Family") %>%
  mutate(
    x_right = x_base + max_left_x + 1,
    y_right = y_base
  ) %>%
  select(labtracking, x_right, y_right, Family)

#---------------------------
# Build links
#---------------------------
asv_mat <- as(otu_table(ps_clean), "matrix")

links <- asv_mat %>%
  as.data.frame() %>%
  rownames_to_column("asv") %>%
  pivot_longer(
    -asv,
    names_to = "labtracking",
    values_to = "Abundance"
  ) %>%
  filter(Abundance > 0) %>%
  left_join(tax_df, by = "asv") %>%
  filter(Family %in% valid_genera) %>% # only valid phyla
  left_join(as(sample_data(ps_clean), "data.frame"), by = "labtracking") %>%
  mutate(
    w = scales::rescale(log10(Abundance + 1), c(0.2, 1))
  )

links_plot <- links %>%
  left_join(left_tip_df,  by = c("asv", "Family")) %>%
  left_join(right_tip_df, by = c("labtracking", "Family")) %>%
  filter(
    !is.na(x_left) & !is.na(y_left) &
    !is.na(x_right) & !is.na(y_right)
  ) %>%
  mutate(
    x_right = ifelse(x_left == x_right, x_right + 1e-6, x_right),
    y_right = ifelse(y_left == y_right, y_right + 1e-6, y_right)
  )



#---------------------------
# Plot
#---------------------------
ggplot() +
  # LEFT trees
  geom_tree(
    data = left_tree_plot_df,
    aes(x = x, y = y),
    inherit.aes = FALSE
  ) +
  # RIGHT trees
  geom_tree(
    data = right_tree_df,
    aes(x = x, y = y),
    inherit.aes = FALSE
  ) +
  # LINKS
  geom_curve(
    data = links_plot,
    aes(
      x = x_left,
      y = y_left,
      xend = x_right,
      yend = y_right,
      color = land_use_location,
      linewidth = w
    ),
    curvature = 0.1,
    alpha = 0.3
  ) +
  scale_color_brewer(palette = "Set2", name = "Habitat") +
  scale_linewidth(range = c(0.2, 1), guide = "none") +
  facet_wrap(~Family, scales = "free_y") +
  theme_void() +
  coord_cartesian(clip = "off")

```


```{r}
phyla <- unique(as.character(tax_table(ps_clean)[, "Phylum"]))

phylum.sum = tapply(taxa_sums(ps_clean), tax_table(ps_clean)[, "Phylum"], sum, na.rm=TRUE)
top5phyla = names(sort(phylum.sum, TRUE))[1:9]
ps_clean = prune_taxa((tax_table(ps_clean)[, "Phylum"] %in% top5phyla), ps_clean)

GP.ord <- ordinate(ps_clean, "PCoA", "bray")
p1 = plot_ordination(ps_clean, GP.ord, type="taxa", color="Phylum", title="taxa")
print(p1)

p1 + facet_wrap(~Phylum, 3)

```

```{r}
#Step 1: Get all phyla
phyla <- unique(tax_table(ps)[, "Phylum"])

# Step 2: Prepare lists
pcoa_list <- list()
axis_labels <- data.frame()

for (ph in phyla) {
  
  # Subset by phylum
  ps_sub <- subset_taxa(ps, Phylum == ph)
  
  # Remove samples with zero counts
  ps_sub <- prune_samples(sample_sums(ps_sub) > 0, ps_sub)
  
  # Skip if fewer than 2 samples
  if (nsamples(ps_sub) < 2) next
  
  # Distance matrix
  dist_mat <- phyloseq::distance(ps_sub, method = "bray")
  if (any(is.na(dist_mat)) || any(is.infinite(dist_mat))) next
  
  # PCoA
  pcoa_res <- ape::pcoa(dist_mat)
  
  # Percent variance
  var_exp <- round(100 * pcoa_res$values$Relative_eig[1:2], 1)
  
  # Sample data
  sample_df <- data.frame(sample_data(ps_sub))
  
  # PCoA dataframe
  pcoa_df <- data.frame(
    Sample = rownames(sample_df),
    PC1 = pcoa_res$vectors[,1],
    PC2 = pcoa_res$vectors[,2],
    Habitat = sample_df$land_use,
    Region = sample_df$Region,
    Phylum = ph
  )
  
  # Store title with phylum + PC labels
  axis_labels <- rbind(axis_labels,
                       data.frame(
                         Phylum = ph,
                         title = paste0(ph, "\nPC1 (", var_exp[1], "%) | PC2 (", var_exp[2], "%)")
                       ))
  
  # Add to list
  pcoa_list[[ph]] <- pcoa_df
}

# Combine all phyla
combined_pcoa <- bind_rows(pcoa_list)

# Join titles
combined_pcoa <- combined_pcoa %>%
  left_join(axis_labels, by = "Phylum")

# Step 3: Plot
ggplot(combined_pcoa, aes(x = PC1, y = PC2, color = Habitat, shape = Region)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(aes(group = interaction(Habitat, Region)), level = 0.95, linetype = 2) +
  facet_wrap(~title, scales = "free") +  # facet titles = Phylum + PC labels
  theme_bw() +
  scale_color_manual(values = habitat_colors) +
  theme(strip.text = element_text(face = "bold", size = 12)) +
  labs(x = "PC1", y = "PC2")

```

