---
title: "NEON microbiome"
author: "Dani Blumstein"
date: "2025-12-30"
output: html_document
---

mostly packages 
```{r}
#if (!requireNamespace("BiocManager", quietly=TRUE))
#    install.packages("BiocManager")
## BiocManager::install("BiocUpgrade") ## you may need this
#BiocManager::install("MicrobiotaProcess",force = TRUE)
#BiocManager::install("phyloseq",force = TRUE)
#BiocManager::install("ggtree", force = TRUE)
#BiocManager::install("tidytree", force = TRUE, type = "source")

#install.packages("remotes")
#remotes::install_github("YuLab-SMU/ggtree")
#remotes::install_github("YuLab-SMU/MicrobiotaProcess")
#devtools::install_github("FrederickHuangLin/ANCOMBC")


#install.packages(
#  "microViz",
#  repos = c(davidbarnett = "https://david-barnett.r-universe.dev", getOption("repos"))
#)

#if (!requireNamespace("devtools", quietly = TRUE)){install.packages("devtools")}
#devtools::install_github("jbisanz/qiime2R")
#devtools::install_github("zpneal/backbone", ref = "devel", build_vignettes = TRUE)

#if (!require("ANCOMBC", quietly = TRUE)){
       #BiocManager::install("ANCOMBC")}

#if (!require("microbiome", quietly = TRUE)){
       #BiocManager::install("microbiome")}

setwd("~/Library/CloudStorage/OneDrive-ArizonaStateUniversity/pdoc_ASU/projects/NEON/Neon_Rproject")

suppressPackageStartupMessages({
    library(MicrobiotaProcess) 
    library(brms)
    library(phyloseq)
    library(ggplot2)
    library(tidyverse) 
    library(vegan) 
    library(coin) 
    library(reshape2) 
    library(ggnewscale)
    library(qiime2R)
    library(readxl)
    library(dplyr)
    library(microViz)
    library(UpSetR)
    library(patchwork)
    library(ggtree)
    library(tidytree)
    library(phylobase)
    library(ggtreeExtra)
    library(sf)
    library(phytools)
    library(SummarizedExperiment)
    library(remotes)
    library(terra)
    library(ggspatial)
    library(stringr)
    library(purrr)
    library(gridExtra)
    library(geosphere)
    library(ape)
    library(phangorn)
    library(ggrepel)
    library(ANCOMBC)
    library(microbiome)
    library(jsonlite)
    library(colorspace)
})
```

colors
```{r}
habitat_colors <- c(
  "forest"         = "lightgreen", 
  "prairie"     = "#9271A7"
)
  
east_west_colors <- c(
  "east" = "#5ab4ac",  # muted steel blue
  "west" = "#d8b365"   # muted warm tan
)

state_colors <- c(
  # WEST (warm tones)
  "California"     = "yellow",  # electric magenta (highly distinct)
  "Washington"     = "#D55E00",
  "Kansas"         = "#F39B7F",
  "North Dakota"   = "#C44E52",
  "Utah"           = "#F28E2B",
  "Colorado"       = "#FF9DA7",
  "Oklahoma"       = "#B5651D",
  "Wyoming"        = "#E17C05",

  # EAST (cool tones)
  "New Hampshire"  = "#4DBBD5",
  "Wisconsin"      = "#3C5488",
  "Virginia"       = "#00A087",
  "Tennessee"      = "#8491B4"
)

base_colors <- c(
  "within east" = "#5ab4ac",
  "within west" = "#d8b365",
  "between regions (east vs west)" = "#D85F64"
)

line_colors <- darken(
  base_colors,
  0.3
)
```

MicrobiotaProcess import
```{r}
mpse_samples <- mp_import_qiime2(otuqza="data/microbiome/rarefied-table.qza", 
                          taxaqza="data/microbiome/taxonomy.qza", 
                          mapfilename="data/microbiome/metadata_edit.tsv")
mpse_samples

#Set labtracking column as the sample names
new_names <- colData(mpse_samples)$labtracking

# Make sure they are unique
stopifnot(!anyDuplicated(new_names))

# Assign as sample names
colnames(mpse_samples) <- new_names
rownames(colData(mpse_samples)) <- new_names

#filter to only pema and drop AZ
mpse_samples_filtered <- mpse_samples %>%
  filter(specificEpithet == "maniculatus") %>%
  filter(stateProvince != "Arizona")

mpse_asv_table <- assay(mpse_samples_filtered, "Abundance")

mpse_bc_dist   <- vegdist(t(mpse_asv_table), method = "bray") 
mpse_sor_dist  <- vegdist(t(mpse_asv_table), method = "bray", binary = TRUE) 

labs <- attr(mpse_bc_dist, "Labels")
```

phyloseq import
```{r}
physeq <- qza_to_phyloseq(
  features  = "data/microbiome/rarefied-table.qza",
  tree      = "data/microbiome/rooted-tree.qza",
  taxonomy  = "data/microbiome/taxonomy.qza",
  metadata  = "MicrobiotaProcess_sample_data2.tsv"
)

#---------------------------
# Sample filtering
#---------------------------
ps <- physeq %>%
  ps_filter(scientificName == "Peromyscus maniculatus") %>%
  ps_filter(stateProvince != "Arizona")

sample_names(ps) <- sample_data(ps)$labtracking

ps_otu <- otu_table(ps)
sort(rowSums(t(ps_otu)))

ps_bc_dist   <- vegdist(t(ps_otu), method = "bray") 
ps_sor_dist  <- vegdist(t(ps_otu), method = "bray", binary = TRUE)

meta <- data.frame(sample_data(ps))

# Assign taxa names
taxa_names(ps3) <- TAXtab$ASVname

meta$decimalLatitude  <- as.numeric(meta$decimalLatitude)
meta$decimalLongitude <- as.numeric(meta$decimalLongitude)

# optional export
write.table(
  meta,
  file = "meta.tsv",
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)
```

geographic distance
```{r}
# -----------------------------
# Geographic distance
# -----------------------------
#distance from a centroid
coords <- as.matrix(meta[, c("decimalLongitude", "decimalLatitude")])
centroid <- colMeans(coords, na.rm = TRUE)
meta$geo_km <- distVincentyEllipsoid(coords, centroid) / 1000

#or 

#distance in a matrix
coords2 <- meta[, c("decimalLongitude", "decimalLatitude")]
geo_dist <- distm(coords2, fun = distHaversine) / 1000 
rownames(geo_dist) <- rownames(meta) 
colnames(geo_dist) <- rownames(meta)
```

ANCOMBC - region
```{r}
# Remove uncultured / unclassified / unidentified taxa
ps2 <- subset_taxa(
  ps,
  !grepl(
    "uncultured|unclassified|unidentified|unknown|uncultured|uncultured_bacterium",
    Genus,
    ignore.case = TRUE
  )
)

out = ancombc(data = ps2, formula = "Region", p_adj_method = "fdr")

results_ancom_bc = data.frame(ASV = out$res$lfc$taxon,
                              lcf = out$res$lfc$Regionwest, 
                              W = out$res$W$Regionwest, 
                              p_val = out$res$p_val$Regionwest, 
                              q_value = out$res$q_val$Regionwest, 
                              Diff_ab =  out$res$diff_abn$Regionwest)
results_ancom_bc$lcf = results_ancom_bc$lcf * -1

results_ancom_bc$association = ifelse(
  results_ancom_bc$q_value < 0.05 & results_ancom_bc$lcf > 0,
  "east",
  "Not Significant"
)

results_ancom_bc$association = ifelse(results_ancom_bc$q_value < 0.05 & results_ancom_bc$lcf < 0, "west",results_ancom_bc$association)

TAXtab =as.data.frame(tax_table(ps2))

name_tmp = sprintf("ASV%s",seq(1:nrow(TAXtab))) 
#now let's add some information to this. For example, we want to add the genus, and if not available, the last taxonomic information available.
TAXtab$name =  TAXtab$Species
TAXtab$name = ifelse(is.na(TAXtab$name), TAXtab$Genus, TAXtab$name)
TAXtab$name = ifelse(is.na(TAXtab$name), TAXtab$Family, TAXtab$name)
TAXtab$name = ifelse(is.na(TAXtab$name), TAXtab$Order, TAXtab$name)
TAXtab$name = ifelse(is.na(TAXtab$name), TAXtab$Class, TAXtab$name)
TAXtab$name = ifelse(is.na(TAXtab$name), TAXtab$Phylum, TAXtab$name)
unique(is.na(TAXtab$name)) # Ok, we have all names

TAXtab$ASVname = paste(name_tmp, TAXtab$name, sep = "-")
TAXtab$name[1:10] # ok it looks good.

write.table(
  TAXtab,
  file = "TAXtab.tsv",
  sep = "\t",
  quote = FALSE
)

results_ancom_bc$ASVname = TAXtab$ASVname[match(results_ancom_bc$ASV ,rownames(TAXtab))]

results_ancom_bc$Genus = TAXtab$Genus[match(results_ancom_bc$ASV ,rownames(TAXtab))]
results_sig = results_ancom_bc[results_ancom_bc$q_value < 0.05,]
table(results_sig$Diff_ab)
table(results_sig$ASVname,results_sig$Diff_ab, results_sig$Genus)

Sum_abundance= colSums(decostand(t(ps_otu), method = "total"))
results_sig$sum_abundance = Sum_abundance[match(results_sig$ASV, names(Sum_abundance))]
results_sig = results_sig[order(results_sig$sum_abundance, decreasing = TRUE),]
results_sig = results_sig[1:14,]
results_sig$ASVname = factor(results_sig$ASVname, levels = rev(results_sig$ASVname))


# Extract OTU table
otu_mat <- as.data.frame(otu_table(ps2))

# Make sure taxa are rows
if (!taxa_are_rows(ps2)) {
  otu_mat <- t(otu_mat)
}

otu_mat <- as.data.frame(otu_mat)

# Sample metadata
meta <- as.data.frame(sample_data(ps2))

# Add Region
meta$SampleID <- rownames(meta)

# Convert OTU matrix to long format
otu_long <- otu_mat %>%
  tibble::rownames_to_column("ASV") %>%
  tidyr::pivot_longer(
    cols = -ASV,
    names_to = "SampleID",
    values_to = "Abundance"
  ) %>%
  left_join(meta[, c("SampleID", "Region")],
            by = "SampleID")

# Get dominant state for each ASV
asv_state <- otu_long %>%
  group_by(ASV, Region) %>%
  summarise(total_abundance = sum(Abundance), .groups = "drop") %>%
  group_by(ASV) %>%
  slice_max(total_abundance, n = 1, with_ties = FALSE) %>%
  ungroup()

# Merge into ANCOMBC results
results_ancom_bc <- left_join(
  results_ancom_bc,
  asv_state[, c("ASV", "Region")],
  by = "ASV"
)

ggplot(results_ancom_bc,
       aes(x = as.numeric(lcf),
           y = -log10(as.numeric(q_value)),
           color = Region)) +
  
  geom_point(size = 3, alpha = 0.9) +
  
  geom_text_repel(
    data = subset(results_ancom_bc, q_value < 0.05),
    aes(label = ASVname),
    size = 4,
    max.overlaps = 20
  ) +
  
  geom_vline(xintercept = 0,
             linetype = "dashed",
             color = "gray50",
             linewidth = 0.8) +
  
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed",
             color = "gray50",
             linewidth = 0.8) +
  
  labs(
    x = "Log2 Fold Change",
    y = expression(-log[10](q-value)),
    title = "Differentially Abundant Taxa",
    color = "Region"
  ) +
  
  theme_bw(base_size = 18) +
  
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    panel.grid = element_blank()
  )

ASVtab_rel = decostand(t(ps_otu), method = "total")
AASVtab_sig = data.frame(ASVtab_rel[,as.numeric(rownames(results_sig))])
colnames(AASVtab_sig) = results_sig$ASVname
AASVtab_sig$Region = meta$Region
melted = reshape2::melt(AASVtab_sig, values = "Region")
melted$variable = factor(melted$variable, levels = rev(results_sig$ASVname))
```

ANCOM-BC - Habitat
```{r}
# ============================================================
# 1. Remove uncultured / unclassified / unidentified taxa
# ============================================================

ps2 <- subset_taxa(
  ps,
  !grepl(
    "uncultured|unclassified|unidentified|unknown|uncultured_bacterium",
    Genus,
    ignore.case = TRUE
  )
)

# ============================================================
# 2. Make sure habitat is a factor
# ============================================================

meta <- data.frame(sample_data(ps2))
meta$SampleID <- rownames(meta)

meta$habitat <- factor(meta$habitat)

# Check the two habitat groups
table(meta$habitat)

# See the order of the levels
levels(meta$habitat)

# ============================================================
# 3. Make sure habitat is in the phyloseq metadata
# ============================================================

sample_data(ps2)$habitat <- meta$habitat[
  match(sample_names(ps2), meta$SampleID)
]

# ============================================================
# 4. Run ANCOM-BC
# ============================================================

out <- ancombc(
  data = ps2,
  formula = "habitat",
  p_adj_method = "fdr"
)

# ============================================================
# 5. Identify the habitat coefficient
# ============================================================

# With a binary variable, ANCOM-BC will produce one
# coefficient comparing the second level to the first.
#
# Check the available coefficient names:
colnames(out$res$lfc)

# ============================================================
# 6. Extract habitat results
# ============================================================

habitat_coef <- setdiff(
  colnames(out$res$lfc),
  "taxon"
)

habitat_coef

results_ancom_bc <- data.frame(
  ASV = out$res$lfc$taxon,
  lcf = out$res$lfc$habitatprairie,
  W = out$res$W$habitatprairie,
  p_val = out$res$p_val$habitatprairie,
  q_value = out$res$q_val$habitatprairie,
  Diff_ab = out$res$diff_abn$habitatprairie
)

# ============================================================
# 7. Add interpretation of direction
# ============================================================

# The coefficient represents:
# second habitat level vs first habitat level

habitat_levels <- levels(meta$habitat)

results_ancom_bc$association <- ifelse(
  results_ancom_bc$q_value < 0.05 &
    results_ancom_bc$lcf > 0,
  habitat_levels[2],
  "Not Significant"
)

results_ancom_bc$association <- ifelse(
  results_ancom_bc$q_value < 0.05 &
    results_ancom_bc$lcf < 0,
  habitat_levels[1],
  results_ancom_bc$association
)

# Add contrast name
results_ancom_bc$Contrast <- paste(
  habitat_levels[2],
  "vs",
  habitat_levels[1]
)

# ============================================================
# 8. Add taxonomy
# ============================================================

TAXtab <- as.data.frame(tax_table(ps2))

name_tmp <- sprintf(
  "ASV%s",
  seq(1, nrow(TAXtab))
)

TAXtab$name <- TAXtab$Species

TAXtab$name <- ifelse(
  is.na(TAXtab$name),
  TAXtab$Genus,
  TAXtab$name
)

TAXtab$name <- ifelse(
  is.na(TAXtab$name),
  TAXtab$Family,
  TAXtab$name
)

TAXtab$name <- ifelse(
  is.na(TAXtab$name),
  TAXtab$Order,
  TAXtab$name
)

TAXtab$name <- ifelse(
  is.na(TAXtab$name),
  TAXtab$Class,
  TAXtab$name
)

TAXtab$name <- ifelse(
  is.na(TAXtab$name),
  TAXtab$Phylum,
  TAXtab$name
)

TAXtab$ASVname <- paste(
  name_tmp,
  TAXtab$name,
  sep = "-"
)

# ============================================================
# 9. Save taxonomy table
# ============================================================

write.table(
  TAXtab,
  file = "TAXtab.tsv",
  sep = "\t",
  quote = FALSE
)

# ============================================================
# 10. Add ASV names and genus
# ============================================================

results_ancom_bc$ASVname <- TAXtab$ASVname[
  match(
    results_ancom_bc$ASV,
    rownames(TAXtab)
  )
]

results_ancom_bc$Genus <- TAXtab$Genus[
  match(
    results_ancom_bc$ASV,
    rownames(TAXtab)
  )
]

# ============================================================
# 11. Significant ASVs
# ============================================================

results_sig <- results_ancom_bc %>%
  filter(q_value < 0.05)

table(results_sig$Diff_ab)
table(
  results_sig$association
)

# ============================================================
# 12. Calculate ASV abundance
# ============================================================

otu_mat <- as.data.frame(otu_table(ps2))

# Make sure taxa are rows
if (!taxa_are_rows(ps2)) {
  otu_mat <- t(otu_mat)
}

otu_mat <- as.data.frame(otu_mat)

ASV_abundance <- rowSums(otu_mat)

results_sig$sum_abundance <- ASV_abundance[
  match(
    results_sig$ASV,
    names(ASV_abundance)
  )
]

results_sig <- results_sig %>%
  arrange(desc(sum_abundance))

# ============================================================
# 13. Keep top 14 significant ASVs
# ============================================================

results_sig <- results_sig %>%
  slice_head(n = 14)

results_sig$ASVname <- factor(
  results_sig$ASVname,
  levels = rev(results_sig$ASVname)
)

# ============================================================
# 14. Volcano plot
# ============================================================

ggplot(
  results_ancom_bc,
  aes(
    x = as.numeric(lcf),
    y = -log10(as.numeric(q_value))
  )
) +

  geom_point(
    size = 3,
    alpha = 0.9
  ) +

  geom_text_repel(
    data = subset(
      results_ancom_bc,
      q_value < 0.05
    ),
    aes(label = ASVname),
    size = 4,
    max.overlaps = 20
  ) +

  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "gray50",
    linewidth = 0.8
  ) +

  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed",
    color = "gray50",
    linewidth = 0.8
  ) +

  labs(
    x = "Log2 Fold Change",
    y = expression(-log[10](q-value)),
    title = paste(
      "Differentially Abundant Taxa:",
      habitat_levels[2],
      "vs",
      habitat_levels[1]
    )
  ) +

  theme_bw(base_size = 18) +

  theme(
    plot.title = element_text(
      size = 22,
      face = "bold",
      hjust = 0.5
    ),
    axis.title = element_text(
      size = 16
    ),
    axis.text = element_text(
      size = 16
    ),
    panel.grid = element_blank()
  )

# ============================================================
# 15. Relative abundance table
# ============================================================

ASVtab_rel <- decostand(
  t(otu_mat),
  method = "total"
)

# Significant ASVs
sig_ASVs <- unique(
  results_sig$ASV
)

ASVtab_sig <- ASVtab_rel[
  ,
  colnames(ASVtab_rel) %in% sig_ASVs,
  drop = FALSE
]

# Rename ASVs
asv_names <- results_sig %>%
  distinct(ASV, .keep_all = TRUE)

colnames(ASVtab_sig) <- asv_names$ASVname[
  match(
    colnames(ASVtab_sig),
    asv_names$ASV
  )
]

# ============================================================
# 16. Add habitat
# ============================================================

meta <- as.data.frame(sample_data(ps2))
meta$SampleID <- rownames(meta)

ASVtab_sig <- as.data.frame(ASVtab_sig)

ASVtab_sig$habitat <- meta$habitat[
  match(
    rownames(ASVtab_sig),
    meta$SampleID
  )
]

# ============================================================
# 17. Melt for plotting
# ============================================================

melted <- reshape2::melt(
  ASVtab_sig,
  id.vars = "habitat"
)

melted$variable <- factor(
  melted$variable,
  levels = rev(
    colnames(ASVtab_sig)[
      colnames(ASVtab_sig) != "habitat"
    ]
  )
)
```

ANCOM-BC - Continuous Temperature
```{r}
# ============================================================
# 1. Remove uncultured / unclassified / unidentified taxa
# ============================================================

ps2 <- subset_taxa(
  ps,
  !grepl(
    "uncultured|unclassified|unidentified|unknown|uncultured_bacterium",
    Genus,
    ignore.case = TRUE
  )
)

# ============================================================
# 2. Make sure temp is numeric
# ============================================================

sample_data(ps2)$temp <- as.numeric(
  as.character(sample_data(ps2)$temp)
)

# ============================================================
# 3. Remove samples with missing temperature
# ============================================================

keep_samples <- !is.na(
  sample_data(ps2)$temp
)

ps2 <- prune_samples(
  keep_samples,
  ps2
)

# Remove taxa that are now completely absent
ps2 <- prune_taxa(
  taxa_sums(ps2) > 0,
  ps2
)

# Check temperature
cat("\nTemperature summary:\n")
print(summary(sample_data(ps2)$temp))

cat("\nNumber of samples:\n")
print(nsamples(ps2))

# ============================================================
# 4. Run ANCOM-BC
# ============================================================

out <- ancombc(
  data = ps2,
  formula = "temp",
  p_adj_method = "fdr"
)

# Check coefficient names
cat("\nANCOM-BC coefficient names:\n")
print(colnames(out$res$lfc))

# ============================================================
# 5. Extract temperature results
# ============================================================

results_ancom_bc <- data.frame(
  ASV = out$res$lfc$taxon,
  lfc = out$res$lfc$temp,
  W = out$res$W$temp,
  p_val = out$res$p_val$temp,
  q_value = out$res$q_val$temp,
  Diff_ab = out$res$diff_abn$temp,
  stringsAsFactors = FALSE
)

# ============================================================
# 6. Remove rows with missing statistics
# ============================================================

results_ancom_bc <- results_ancom_bc %>%
  filter(
    !is.na(lfc),
    !is.na(q_value)
  )

# ============================================================
# 7. Define direction of temperature association
# ============================================================

results_ancom_bc$association <- "Not Significant"

results_ancom_bc$association[
  results_ancom_bc$q_value < 0.05 &
    results_ancom_bc$lfc > 0
] <- "Increases with temperature"

results_ancom_bc$association[
  results_ancom_bc$q_value < 0.05 &
    results_ancom_bc$lfc < 0
] <- "Decreases with temperature"

# ============================================================
# 8. Taxonomy table
# ============================================================

TAXtab <- as.data.frame(
  tax_table(ps2)
)

TAXtab$ASV <- rownames(TAXtab)

# ------------------------------------------------------------
# Make a taxonomic name using the most specific available
# ------------------------------------------------------------

TAXtab$name <- NA_character_

tax_levels <- c(
  "Species",
  "Genus",
  "Family",
  "Order",
  "Class",
  "Phylum"
)

for (tax_level in tax_levels) {

  if (tax_level %in% colnames(TAXtab)) {

    TAXtab$name[
      is.na(TAXtab$name) |
        TAXtab$name == ""
    ] <- as.character(
      TAXtab[[tax_level]][
        is.na(TAXtab$name) |
          TAXtab$name == ""
      ]
    )
  }
}

# Replace any remaining missing names
TAXtab$name[
  is.na(TAXtab$name) |
    TAXtab$name == ""
] <- "Unclassified"

# ============================================================
# 9. Create ASV names
# ============================================================

TAXtab$ASVname <- paste(
  paste0(
    "ASV",
    seq_len(nrow(TAXtab))
  ),
  TAXtab$name,
  sep = "-"
)

# ============================================================
# 10. Save taxonomy table
# ============================================================

write.table(
  TAXtab,
  file = "TAXtab.tsv",
  sep = "\t",
  quote = FALSE,
  row.names = TRUE
)

# ============================================================
# 11. Add taxonomy to ANCOM-BC results
# ============================================================

results_ancom_bc$ASVname <- TAXtab$ASVname[
  match(
    results_ancom_bc$ASV,
    TAXtab$ASV
  )
]

# Add Genus if available
if ("Genus" %in% colnames(TAXtab)) {

  results_ancom_bc$Genus <- TAXtab$Genus[
    match(
      results_ancom_bc$ASV,
      TAXtab$ASV
    )
  ]

} else {

  results_ancom_bc$Genus <- NA_character_

}

# ============================================================
# 12. Significant ASVs
# ============================================================

results_sig <- results_ancom_bc %>%
  filter(q_value < 0.05)

cat(
  "\nNumber of significant ASVs (q < 0.05): ",
  nrow(results_sig),
  "\n",
  sep = ""
)

cat("\nDirection of significant associations:\n")
print(table(results_sig$association))

# ============================================================
# 13. Get OTU table
# ============================================================

otu_mat <- as.data.frame(
  otu_table(ps2)
)

# Make sure taxa are rows
if (!taxa_are_rows(ps2)) {
  otu_mat <- t(otu_mat)
}

otu_mat <- as.data.frame(otu_mat)

# ============================================================
# 14. Total abundance per ASV
# ============================================================

ASV_abundance <- rowSums(
  otu_mat
)

results_ancom_bc$sum_abundance <- ASV_abundance[
  match(
    results_ancom_bc$ASV,
    names(ASV_abundance)
  )
]

results_sig$sum_abundance <- ASV_abundance[
  match(
    results_sig$ASV,
    names(ASV_abundance)
  )
]

# ============================================================
# 15. Sort significant ASVs by abundance
# ============================================================

results_sig <- results_sig %>%
  arrange(
    desc(sum_abundance)
  )

# ============================================================
# 16. Top 14 significant ASVs
# ============================================================

results_sig_top <- results_sig %>%
  slice_head(n = 14)

if (nrow(results_sig_top) > 0) {

  results_sig_top$ASVname <- factor(
    results_sig_top$ASVname,
    levels = rev(
      results_sig_top$ASVname
    )
  )

}

# ============================================================
# 17. Volcano plot
# ============================================================

p_volcano <- ggplot(
  results_ancom_bc,
  aes(
    x = lfc,
    y = -log10(q_value),
    color = association
  )
) +

  geom_point(
    size = 3,
    alpha = 0.9
  ) +

  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "gray50",
    linewidth = 0.8
  ) +

  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed",
    color = "gray50",
    linewidth = 0.8
  ) +

  scale_color_manual(
    values = c(
      "Increases with temperature" = "#D73027",
      "Decreases with temperature" = "#4575B4",
      "Not Significant" = "gray70"
    )
  ) +

  labs(
    x = "Temperature Log2 Fold Change",
    y = expression(-log[10](q-value)),
    title = "Differentially Abundant Taxa Along Temperature Gradient",
    color = "Temperature association"
  ) +

  theme_bw(
    base_size = 18
  ) +

  theme(
    plot.title = element_text(
      size = 22,
      face = "bold",
      hjust = 0.5
    ),
    axis.title = element_text(
      size = 16
    ),
    axis.text = element_text(
      size = 16
    ),
    legend.title = element_text(
      size = 16,
      face = "bold"
    ),
    legend.text = element_text(
      size = 14
    ),
    panel.grid = element_blank()
  )

# ------------------------------------------------------------
# Add labels ONLY if significant ASVs exist
# ------------------------------------------------------------

if (nrow(results_sig) > 0) {

  p_volcano <- p_volcano +
    geom_text_repel(
      data = results_sig,
      aes(
        label = ASVname
      ),
      size = 4,
      max.overlaps = 20
    )

}

# Display volcano plot
print(p_volcano)

# ============================================================
# 18. Relative abundance table
# ============================================================

ASVtab_rel <- decostand(
  t(otu_mat),
  method = "total"
)

# ============================================================
# 19. Keep significant ASVs
# ============================================================

sig_ASVs <- unique(
  results_sig$ASV
)

if (length(sig_ASVs) > 0) {

  ASVtab_sig <- ASVtab_rel[
    ,
    colnames(ASVtab_rel) %in% sig_ASVs,
    drop = FALSE
  ]

} else {

  ASVtab_sig <- ASVtab_rel[
    ,
    FALSE,
    drop = FALSE
  ]

}

cat(
  "\nNumber of significant ASV columns in abundance table: ",
  ncol(ASVtab_sig),
  "\n",
  sep = ""
)

# ============================================================
# 20. Rename significant ASVs
# ============================================================

if (ncol(ASVtab_sig) > 0) {

  asv_names <- results_sig %>%
    distinct(
      ASV,
      .keep_all = TRUE
    )

  new_names <- asv_names$ASVname[
    match(
      colnames(ASVtab_sig),
      asv_names$ASV
    )
  ]

  # Keep ASV ID if a name could not be matched
  new_names[
    is.na(new_names)
  ] <- colnames(
    ASVtab_sig
  )[
    is.na(new_names)
  ]

  colnames(ASVtab_sig) <- new_names

}

# ============================================================
# 21. Add temperature to abundance table
# ============================================================

meta <- as.data.frame(
  sample_data(ps2)
)

meta$SampleID <- rownames(meta)

ASVtab_sig$temp <- meta$temp[
  match(
    rownames(ASVtab_sig),
    meta$SampleID
  )
]

# ============================================================
# 22. Melt abundance table
# ============================================================

if (ncol(ASVtab_sig) > 1) {

  melted <- reshape2::melt(
    ASVtab_sig,
    id.vars = "temp",
    variable.name = "ASVname",
    value.name = "RelativeAbundance"
  )

  melted$ASVname <- factor(
    melted$ASVname,
    levels = rev(
      unique(melted$ASVname)
    )
  )

} else {

  melted <- NULL

  message(
    "\nNo ASVs were significant at q < 0.05.",
    "\n'melted' was set to NULL."
  )

}

# ============================================================
# 23. Results tables
# ============================================================

# All ANCOM-BC results
results_ancom_bc <- results_ancom_bc %>%
  arrange(
    q_value
  )

# Significant results only
results_sig <- results_ancom_bc %>%
  filter(
    q_value < 0.05
  )

# ============================================================
# 24. Print the most important results
# ============================================================

cat("\n============================================\n")
cat("ANCOM-BC CONTINUOUS TEMPERATURE RESULTS\n")
cat("============================================\n\n")

cat(
  "Total ASVs tested: ",
  nrow(results_ancom_bc),
  "\n",
  sep = ""
)

cat(
  "Significant ASVs (q < 0.05): ",
  nrow(results_sig),
  "\n",
  sep = ""
)

if (nrow(results_sig) > 0) {

  cat("\nSignificant taxa:\n")

  print(
    results_sig %>%
      select(
        ASV,
        ASVname,
        Genus,
        lfc,
        p_val,
        q_value,
        association
      )
  )

} else {

  cat(
    "\nNo ASVs were significant at q < 0.05.\n"
  )

}
```

ASV by genus trees loop and multipannel
```{r}
genus_list <- unique(tax_table(ps_top10)[, "Genus"])
genus_list <- genus_list[!is.na(genus_list)]

plot_list <- list()

for (g in genus_list) {

  ps_g <- subset_taxa(ps2, Genus == g)

  # skip if no taxa
  if (ntaxa(ps_g) == 0) next

  # skip if no tree
  if (is.null(phy_tree(ps_g, errorIfNULL = FALSE))) next

  tree <- phy_tree(ps_g)

  # sometimes pruning breaks tree validity
  if (is.null(tree)) next

  tree_plot <- ggtree(tree)
  tree_plot <- revts(tree_plot)

  tree_df <- psmelt(ps_g) %>%
    select(OTU, Region) %>%
    filter(!is.na(Region)) %>%
    distinct()

  tree_df$name <- TAXtab[match(tree_df$OTU, TAXtab$ASVname), "name"]
  tree_df$label_name <- ifelse(!duplicated(tree_df$name), tree_df$name, NA)

  p <- tree_plot %<+% tree_df +
    geom_tippoint(aes(color = Region),
                  alpha = 0.6,
                  size = 0.7) +
    geom_tiplab(
      aes(label = label_name),
      align = FALSE,
      linetype = NULL,
      size = 3,
      offset = 0.002,
      hjust = 0,
      na.rm = TRUE
    ) +
    scale_color_manual(values = east_west_colors) +
    ggtitle(g) +
    theme_tree2() +
    theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      legend.position = "none"
    ) +
    expand_limits(x = max(tree_plot$data$x) + 0.02)

  plot_list[[g]] <- p
}

wrap_plots(plot_list, ncol = 3)
```

ASV by genus trees
```{r}
# 1. Agglomerate taxa to Genus level
ps_genus <- tax_glom(ps2, taxrank = "Genus")

# 2. Get names of top 10 genera by total abundance
top10_names <- names(sort(taxa_sums(ps_genus), decreasing = TRUE))[1:10]
#bottom10_names <- names(sort(taxa_sums(ps_genus), decreasing = FALSE)[1:10])

# 3. Prune the object to keep only these 10 (optional)
#ps_bottom10 <- prune_taxa(bottom10_names, ps_genus)

# 4. View the Taxa Table for these bottom 10
#tax_table(ps_bottom10)

# 3. Prune the agglomerated object to only these 10
ps_top10 <- prune_taxa(top10_names, ps_genus)
tax_table(ps_top10)

ps3 <- ps2

# Assign taxa names
taxa_names(ps3) <- TAXtab$ASVname


# Select genus of interest
g <- "Lactobacillus"
#Ruminococcus"
#Desulfovibrio"
#Helicobacter"
#Colidextribacter"
#Lachnospiraceae_UCG-006"
#Roseburia"
#Muribaculaceae"
#Lachnospiraceae_UCG-001"
#Lachnospiraceae_NK4A136_group"
#Lactobacillus"

# Subset phyloseq object for that genus
ps_g <- subset_taxa(ps3, Genus == g)
phy_tree <- phy_tree(ps_g)

# Create tree plot
tree_plot <- ggtree(phy_tree)

asvtree<- phy_tree(ps_g)
plotTree(phy_tree(ps_g),fsize=0.8,ftype="i",
    lwd=1)

# Extract OTU and Region information
tree_df <- psmelt(ps_g) %>%
  dplyr::select(OTU, stateProvince,Abundance, Region) %>%
  filter(!is.na(Region)) %>%
  distinct()

# Add taxonomy name
tree_df$name <- TAXtab[match(tree_df$OTU, TAXtab$ASVname), "name"]

# Keep one label per unique name
tree_df$label_name <- ifelse(!duplicated(tree_df$name), tree_df$name, NA)

# Keep only tips with names
tip_positions <- tree_plot$data %>%
  filter(isTip) %>%
  filter(!is.na(label))

# Compute clade center for labeling (optional)
clade_labels <- tip_positions %>%
  group_by(label) %>%
  summarise(
    x = mean(x),
    y = mean(y),
    .groups = "drop"
  )

# Plot tree with tip points and labels
tree_plot_flipped <- revts(tree_plot)

tree_plot %<+% tree_df +
  geom_tippoint(aes(color = Region),
                alpha = 0.5,
                size = 2.5) +

  geom_tiplab(
    aes(label = label_name),
    align = FALSE,     # <-- removes vertical guide lines
    linetype = NULL,   # <-- extra safety (no lines)
    size = 3,
    offset = 0.01,    # <-- very small = right at tip
    hjust = 0,
    na.rm = TRUE
  ) +

  #scale_color_manual(values = east_west_colors) +

  ggtitle(g) +

  theme_tree2() +

  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  ) +

  expand_limits(x = max(tree_plot$data$x) + 0.02)
```

genus phylo and geo matrix
```{r}
# Subset phyloseq object for that genus
ps_g <- subset_taxa(ps2, Genus == g)
phy_tree <- phy_tree(ps_g)
phy_matrix <- cophenetic(phy_tree)
print(phy_matrix)

# setting
individual_col <- "labtracking"
lat_col <- "decimalLatitude"
lon_col <- "decimalLongitude"

summary_fun <- "mean"   # "mean", "min", or "median"
output_units <- "km"    # "km" or "m"

# -----------------------------
# Extract ASV table and metadata
# -----------------------------

asv_tab <- as(otu_table(ps_g), "matrix")

if (!taxa_are_rows(ps_g)) {
  asv_tab <- t(asv_tab)
}

meta <- data.frame(sample_data(ps_g), check.names = FALSE)

# Align metadata to samples
sample_ids <- colnames(asv_tab)
meta <- meta[sample_ids, , drop = FALSE]

# -----------------------------
# Remove duplicate ASV names if present
# -----------------------------

if (any(duplicated(rownames(asv_tab)))) {
  stop("Duplicate ASV names found in rownames(asv_tab). Fix ASV names before making the matrix.")
}

# -----------------------------
# Make ASV x individual presence/absence matrix
# -----------------------------

asv_pa_sample <- asv_tab > 0

individuals <- unique(meta[[individual_col]])

asv_by_ind <- sapply(individuals, function(ind) {
  samples_for_ind <- rownames(meta)[meta[[individual_col]] == ind]
  rowSums(asv_pa_sample[, samples_for_ind, drop = FALSE]) > 0
})

asv_by_ind <- as.matrix(asv_by_ind)
mode(asv_by_ind) <- "numeric"

rownames(asv_by_ind) <- rownames(asv_tab)
colnames(asv_by_ind) <- individuals

# -----------------------------
# Remove ASVs absent from all individuals
# -----------------------------

#asv_by_ind <- asv_by_ind[rowSums(asv_by_ind) > 0, , drop = FALSE]

# -----------------------------
# Get one lat/long per individual
# -----------------------------

individual_coords <- meta[
  !duplicated(meta[[individual_col]]),
  c(individual_col, lon_col, lat_col),
  drop = FALSE
]

individual_coords <- individual_coords[
  complete.cases(individual_coords[, c(lon_col, lat_col)]),
  ,
  drop = FALSE
]

rownames(individual_coords) <- individual_coords[[individual_col]]

# Keep only individuals with coordinates
common_inds <- intersect(colnames(asv_by_ind), rownames(individual_coords))

#asv_by_ind <- asv_by_ind[, common_inds, drop = FALSE]
individual_coords <- individual_coords[common_inds, , drop = FALSE]

# -----------------------------
# Calculate individual geographic distances ONCE
# -----------------------------

ind_coords_mat <- as.matrix(individual_coords[, c(lon_col, lat_col)])

ind_geo_dist <- geosphere::distm(
  ind_coords_mat,
  ind_coords_mat,
  fun = geosphere::distHaversine
)

rownames(ind_geo_dist) <- rownames(individual_coords)
colnames(ind_geo_dist) <- rownames(individual_coords)

# -----------------------------
# Build ASV x ASV geographic distance matrix
# -----------------------------

asvs <- rownames(asv_by_ind)

asv_geo_dist <- matrix(
  NA_real_,
  nrow = length(asvs),
  ncol = length(asvs),
  dimnames = list(asvs, asvs)
)

summarize_distance <- switch(
  summary_fun,
  mean = mean,
  min = min,
  median = median
)

for (i in seq_along(asvs)) {
  inds_i <- colnames(asv_by_ind)[asv_by_ind[i, ] > 0]

  for (j in i:length(asvs)) {
    inds_j <- colnames(asv_by_ind)[asv_by_ind[j, ] > 0]

    pairwise_distances <- ind_geo_dist[inds_i, inds_j, drop = FALSE]

    d <- summarize_distance(pairwise_distances, na.rm = TRUE)

    asv_geo_dist[i, j] <- d
    asv_geo_dist[j, i] <- d
  }
}

diag(asv_geo_dist) <- 0

if (output_units == "km") {
  asv_geo_dist <- asv_geo_dist / 1000
}
```

plots
```{r}
# phy_matrix = ASV x ASV phylogenetic distance matrix
# asv_geo_dist   = ASV x ASV geographic distance matrix

# Make sure both matrices are in the same ASV order
common_asvs <- intersect(rownames(phy_matrix), rownames(asv_geo_dist))

phy_matrix <- phy_matrix[common_asvs, common_asvs]
asv_geo_dist <- asv_geo_dist[common_asvs, common_asvs]

# Pull out only the lower triangle
lower_tri <- lower.tri(phy_matrix)

plot_df <- data.frame(
  ASV1 = rownames(phy_matrix)[row(phy_matrix)[lower_tri]],
  ASV2 = colnames(phy_matrix)[col(phy_matrix)[lower_tri]],
  phy_matrix = phy_matrix[lower_tri],
  asv_geo_dist = asv_geo_dist[lower_tri]
)

# Remove missing / infinite values
plot_df <- plot_df %>%
  filter(
    !is.na(phy_matrix),
    !is.na(asv_geo_dist),
    is.finite(phy_matrix),
    is.finite(asv_geo_dist)
  )

library(ggExtra)
library(cowplot)

p1 <- ggplot(plot_df, aes(x = asv_geo_dist, y = phy_matrix)) +
  geom_bin2d(bins = 80) +
  geom_point(alpha = 0.001, size = 0.1) +
  scale_fill_viridis_c(option = "magma", trans = "log10") +
  theme_classic() +
  labs(
    x = "Geographic distance between ASVs (km)",
    y = "Phylogenetic distance between ASVs",
    fill = "ASV pairs",
    title = g
  ) +
  theme(
    legend.position = "right"
  )

# Extract the legend
asv_legend <- cowplot::get_legend(p1)

# Make the marginal plot without the legend
p1_no_legend <- p1 +
  theme(legend.position = "none")

p1_marginal <- ggMarginal(
  p1_no_legend,
  type = "histogram",
  bins = 60,
  fill = "orange",
  color = "white"
)

# Put legend outside the marginal histograms
final_plot <- cowplot::plot_grid(
  p1_marginal,
  asv_legend,
  nrow = 1,
  rel_widths = c(1, 0.18)
)

final_plot

p2 <- ggplot(plot_df, aes(x = asv_geo_dist, y = phy_matrix)) +
  geom_point(
    alpha = 0.25,
    size = 1.2,
    color = "darkorange"
  ) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "black",
    fill = "gray70",
    linewidth = 1
  ) +
  theme_classic() +
  labs(
    title = g,
    x = "Geographic distance between ASVs (km)",
    y = "Phylogenetic distance between ASVs"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

p2 

mantel_result <- mantel(
  as.dist(phy_matrix),
  as.dist(asv_geo_dist),
  method = "spearman",
  permutations = 999
)

mantel_result
```

asv phylo and geo matrix
```{r}
# Use your original full phyloseq object, not the genus-subsetted one
# Example:
ps_full <- ps

ambiguous_pattern <- paste(
  c(
    "uncultured",
    "unidentified",
    "metagenome",
    "bacterium",
    "organism",
    "sp\\.",
    "spp\\.",
    "cf\\.",
    "aff\\."
  ),
  collapse = "|"
)

# -----------------------------
# Find top 10 named species
# -----------------------------

tax_df <- as.data.frame(tax_table(ps_full), check.names = FALSE)

otu_mat <- as(otu_table(ps_full), "matrix")

if (!taxa_are_rows(ps_full)) {
  otu_mat <- t(otu_mat)
}

tax_df$ASV <- rownames(tax_df)
tax_df$total_abundance <- rowSums(otu_mat)

tax_named <- tax_df %>%
  filter(
    !is.na(.data[[species_col]]),
    .data[[species_col]] != "",
    !grepl(ambiguous_pattern, .data[[species_col]], ignore.case = TRUE)
  )

top10_species <- tax_named %>%
  group_by(.data[[species_col]]) %>%
  summarize(
    species_abundance = sum(total_abundance),
    n_asvs = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(species_abundance)) %>%
  slice_head(n = 10)

print(top10_species)

top10_asvs <- tax_named %>%
  filter(.data[[species_col]] %in% top10_species[[species_col]]) %>%
  pull(ASV)

ps_top10_species <- prune_taxa(top10_asvs, ps_full)

# -----------------------------
# Remake phylogenetic distance matrix
# -----------------------------

phy_matrix <- cophenetic(phy_tree(ps_top10_species))

# -----------------------------
# Remake geographic ASV distance matrix
# -----------------------------

asv_tab <- as(otu_table(ps_top10_species), "matrix")

if (!taxa_are_rows(ps_top10_species)) {
  asv_tab <- t(asv_tab)
}

meta <- data.frame(sample_data(ps_top10_species), check.names = FALSE)

needed_cols <- c(individual_col, lat_col, lon_col)
missing_cols <- setdiff(needed_cols, colnames(meta))

if (length(missing_cols) > 0) {
  stop("Missing metadata columns: ", paste(missing_cols, collapse = ", "))
}

sample_ids <- colnames(asv_tab)

if (!all(sample_ids %in% rownames(meta))) {
  missing_samples <- setdiff(sample_ids, rownames(meta))
  stop(
    "Some OTU-table sample names are missing from sample_data rownames. Examples: ",
    paste(head(missing_samples), collapse = ", ")
  )
}

meta <- meta[sample_ids, , drop = FALSE]

asv_pa_sample <- asv_tab > 0

individuals <- unique(meta[[individual_col]])

asv_by_ind <- sapply(individuals, function(ind) {
  samples_for_ind <- rownames(meta)[meta[[individual_col]] == ind]
  rowSums(asv_pa_sample[, samples_for_ind, drop = FALSE]) > 0
})

asv_by_ind <- as.matrix(asv_by_ind)
mode(asv_by_ind) <- "numeric"

rownames(asv_by_ind) <- rownames(asv_tab)
colnames(asv_by_ind) <- individuals

#asv_by_ind <- asv_by_ind[rowSums(asv_by_ind) > 0, , drop = FALSE]

individual_coords <- meta[
  !duplicated(meta[[individual_col]]),
  c(individual_col, lon_col, lat_col),
  drop = FALSE
]

individual_coords <- individual_coords[
  complete.cases(individual_coords[, c(lon_col, lat_col)]),
  ,
  drop = FALSE
]

rownames(individual_coords) <- individual_coords[[individual_col]]

common_inds <- intersect(colnames(asv_by_ind), rownames(individual_coords))

#asv_by_ind <- asv_by_ind[, common_inds, drop = FALSE]
individual_coords <- individual_coords[common_inds, , drop = FALSE]

ind_coords_mat <- as.matrix(individual_coords[, c(lon_col, lat_col)])

ind_geo_dist <- geosphere::distm(
  ind_coords_mat,
  ind_coords_mat,
  fun = geosphere::distHaversine
)

rownames(ind_geo_dist) <- rownames(individual_coords)
colnames(ind_geo_dist) <- rownames(individual_coords)

asvs <- rownames(asv_by_ind)

asv_geo_dist <- matrix(
  NA_real_,
  nrow = length(asvs),
  ncol = length(asvs),
  dimnames = list(asvs, asvs)
)

summarize_distance <- switch(
  summary_fun,
  mean = mean,
  min = min,
  median = median
)

for (i in seq_along(asvs)) {
  inds_i <- colnames(asv_by_ind)[asv_by_ind[i, ] > 0]

  for (j in i:length(asvs)) {
    inds_j <- colnames(asv_by_ind)[asv_by_ind[j, ] > 0]

    pairwise_distances <- ind_geo_dist[inds_i, inds_j, drop = FALSE]

    d <- summarize_distance(pairwise_distances, na.rm = TRUE)

    asv_geo_dist[i, j] <- d
    asv_geo_dist[j, i] <- d
  }
}

diag(asv_geo_dist) <- 0

if (output_units == "km") {
  asv_geo_dist <- asv_geo_dist / 1000
}

# -----------------------------
# Plot one graph per top species
# -----------------------------

tax_top10 <- as.data.frame(tax_table(ps_top10_species), check.names = FALSE)
tax_top10$ASV <- rownames(tax_top10)

plot_list <- list()
diagnostic_df <- data.frame()

for (g in top10_species[[species_col]]) {

  species_asvs <- tax_top10$ASV[tax_top10[[species_col]] == g]

  species_asvs <- Reduce(
    intersect,
    list(
      species_asvs,
      rownames(phy_matrix),
      rownames(asv_geo_dist)
    )
  )

  n_pairwise <- ifelse(
    length(species_asvs) >= 2,
    choose(length(species_asvs), 2),
    0
  )

  diagnostic_df <- rbind(
    diagnostic_df,
    data.frame(
      species = g,
      asvs_used = length(species_asvs),
      pairwise_comparisons = n_pairwise
    )
  )

  if (length(species_asvs) < 2) {
    next
  }

  phy_sub <- phy_matrix[species_asvs, species_asvs, drop = FALSE]
  geo_sub <- asv_geo_dist[species_asvs, species_asvs, drop = FALSE]

  lower_tri <- lower.tri(phy_sub)

  plot_df <- data.frame(
    ASV1 = rownames(phy_sub)[row(phy_sub)[lower_tri]],
    ASV2 = colnames(phy_sub)[col(phy_sub)[lower_tri]],
    phy_distance = phy_sub[lower_tri],
    geo_distance = geo_sub[lower_tri]
  ) %>%
    filter(
      !is.na(phy_distance),
      !is.na(geo_distance),
      is.finite(phy_distance),
      is.finite(geo_distance)
    )

  if (nrow(plot_df) == 0) {
    next
  }

  p <- ggplot(plot_df, aes(x = geo_distance, y = phy_distance)) +
    geom_point(
      alpha = 0.35,
      size = 1.1,
      color = "darkorange"
    ) +
    theme_classic() +
    labs(
      title = g,
      x = "Geo distance (km)",
      y = "Phylo distance"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "italic")
    )

  if (nrow(plot_df) >= 3 && length(unique(plot_df$geo_distance)) >= 2) {
    p <- p +
      geom_smooth(
        method = "lm",
        se = TRUE,
        color = "black",
        fill = "gray70",
        linewidth = 0.8
      )
  }

  plot_list[[g]] <- p
}

print(diagnostic_df)

combined_plot <- wrap_plots(plot_list, ncol = 2)

combined_plot
```

asv phylo and geo matrix v2: rank them by number of ASVs/tips in the phylogeny
```{r}
# setting
individual_col <- "labtracking"
lat_col <- "decimalLatitude"
lon_col <- "decimalLongitude"
species_col <- "Species"
summary_fun <- "mean"   # "mean", "min", or "median"
output_units <- "km"    # "km" or "m"

point_color <- lighten("grey65", 0.3)
line_color <- "grey35"

ambiguous_pattern <- paste(
  c(
    "uncultured",
    "unidentified",
    "metagenome",
    "bacterium",
    "organism",
    "sp\\.",
    "spp\\.",
    "cf\\.",
    "aff\\."
  ),
  collapse = "|"
)

# -----------------------------
# Extract taxonomy and tree tips
# -----------------------------

tax_df <- as.data.frame(tax_table(ps_full), check.names = FALSE)
tax_df$ASV <- rownames(tax_df)

if (is.null(phy_tree(ps_full))) {
  stop("ps_full does not contain a phylogenetic tree.")
}

tree_tips <- phy_tree(ps_full)$tip.label

# -----------------------------
# Keep only named species with ASVs present in the phylogeny
# -----------------------------

tax_named_phylo <- tax_df %>%
  filter(
    !is.na(.data[[species_col]]),
    .data[[species_col]] != "",
    !grepl(ambiguous_pattern, .data[[species_col]], ignore.case = TRUE),
    ASV %in% tree_tips
  )

# -----------------------------
# Rank species by number of ASVs/tips in the phylogeny
# -----------------------------

top10_species <- tax_named_phylo %>%
  group_by(.data[[species_col]]) %>%
  summarize(
    n_tips = n_distinct(ASV),
    .groups = "drop"
  ) %>%
  arrange(desc(n_tips)) %>%
  slice_head(n = 10)

print(top10_species)

# -----------------------------
# Subset phyloseq object to ASVs from top 10 species
# -----------------------------

top10_asvs <- tax_named_phylo %>%
  filter(.data[[species_col]] %in% top10_species[[species_col]]) %>%
  pull(ASV)

ps_top10_species <- prune_taxa(top10_asvs, ps_full)

# -----------------------------
# Remake phylogenetic distance matrix
# -----------------------------

phy_matrix <- cophenetic(phy_tree(ps_top10_species))

# -----------------------------
# Remake geographic ASV distance matrix
# -----------------------------

asv_tab <- as(otu_table(ps_top10_species), "matrix")

if (!taxa_are_rows(ps_top10_species)) {
  asv_tab <- t(asv_tab)
}

meta <- data.frame(sample_data(ps_top10_species), check.names = FALSE)

needed_cols <- c(individual_col, lat_col, lon_col)
missing_cols <- setdiff(needed_cols, colnames(meta))

if (length(missing_cols) > 0) {
  stop("Missing metadata columns: ", paste(missing_cols, collapse = ", "))
}

sample_ids <- colnames(asv_tab)

if (!all(sample_ids %in% rownames(meta))) {
  missing_samples <- setdiff(sample_ids, rownames(meta))
  stop(
    "Some OTU-table sample names are missing from sample_data rownames. Examples: ",
    paste(head(missing_samples), collapse = ", ")
  )
}

meta <- meta[sample_ids, , drop = FALSE]

asv_pa_sample <- asv_tab > 0

individuals <- unique(meta[[individual_col]])

asv_by_ind <- sapply(individuals, function(ind) {
  samples_for_ind <- rownames(meta)[meta[[individual_col]] == ind]
  rowSums(asv_pa_sample[, samples_for_ind, drop = FALSE]) > 0
})

asv_by_ind <- as.matrix(asv_by_ind)
mode(asv_by_ind) <- "numeric"

rownames(asv_by_ind) <- rownames(asv_tab)
colnames(asv_by_ind) <- individuals

asv_by_ind <- asv_by_ind[rowSums(asv_by_ind) > 0, , drop = FALSE]

individual_coords <- meta[
  !duplicated(meta[[individual_col]]),
  c(individual_col, lon_col, lat_col),
  drop = FALSE
]

individual_coords <- individual_coords[
  complete.cases(individual_coords[, c(lon_col, lat_col)]),
  ,
  drop = FALSE
]

rownames(individual_coords) <- individual_coords[[individual_col]]

common_inds <- intersect(colnames(asv_by_ind), rownames(individual_coords))

asv_by_ind <- asv_by_ind[, common_inds, drop = FALSE]
individual_coords <- individual_coords[common_inds, , drop = FALSE]

ind_coords_mat <- as.matrix(individual_coords[, c(lon_col, lat_col)])

ind_geo_dist <- geosphere::distm(
  ind_coords_mat,
  ind_coords_mat,
  fun = geosphere::distHaversine
)

rownames(ind_geo_dist) <- rownames(individual_coords)
colnames(ind_geo_dist) <- rownames(individual_coords)

asvs <- rownames(asv_by_ind)

asv_geo_dist <- matrix(
  NA_real_,
  nrow = length(asvs),
  ncol = length(asvs),
  dimnames = list(asvs, asvs)
)

summarize_distance <- switch(
  summary_fun,
  mean = mean,
  min = min,
  median = median
)

for (i in seq_along(asvs)) {
  inds_i <- colnames(asv_by_ind)[asv_by_ind[i, ] > 0]

  for (j in seq(i, length(asvs))) {
    inds_j <- colnames(asv_by_ind)[asv_by_ind[j, ] > 0]

    pairwise_distances <- ind_geo_dist[inds_i, inds_j, drop = FALSE]

    d <- summarize_distance(pairwise_distances, na.rm = TRUE)

    asv_geo_dist[i, j] <- d
    asv_geo_dist[j, i] <- d
  }
}

diag(asv_geo_dist) <- 0

if (output_units == "km") {
  asv_geo_dist <- asv_geo_dist / 1000
}

# -----------------------------
# Plot one graph per top species and run Mantel tests
# -----------------------------

tax_top10 <- as.data.frame(tax_table(ps_top10_species), check.names = FALSE)
tax_top10$ASV <- rownames(tax_top10)

plot_list <- list()
diagnostic_df <- data.frame()
mantel_results <- data.frame()

for (g in top10_species[[species_col]]) {

  species_asvs <- tax_top10$ASV[tax_top10[[species_col]] == g]

  species_asvs <- Reduce(
    intersect,
    list(
      species_asvs,
      rownames(phy_matrix),
      rownames(asv_geo_dist)
    )
  )

  n_pairwise <- ifelse(
    length(species_asvs) >= 2,
    choose(length(species_asvs), 2),
    0
  )

  diagnostic_df <- rbind(
    diagnostic_df,
    data.frame(
      species = g,
      tips_used = length(species_asvs),
      pairwise_comparisons = n_pairwise
    )
  )

  if (length(species_asvs) < 2) {
    mantel_results <- rbind(
      mantel_results,
      data.frame(
        species = g,
        tips_used = length(species_asvs),
        pairwise_comparisons = n_pairwise,
        mantel_r = NA_real_,
        mantel_p = NA_real_
      )
    )
    next
  }

  phy_sub <- phy_matrix[species_asvs, species_asvs, drop = FALSE]
  geo_sub <- asv_geo_dist[species_asvs, species_asvs, drop = FALSE]

  lower_tri <- lower.tri(phy_sub)

  plot_df <- data.frame(
    ASV1 = rownames(phy_sub)[row(phy_sub)[lower_tri]],
    ASV2 = colnames(phy_sub)[col(phy_sub)[lower_tri]],
    phy_distance = phy_sub[lower_tri],
    geo_distance = geo_sub[lower_tri]
  ) %>%
    filter(
      !is.na(phy_distance),
      !is.na(geo_distance),
      is.finite(phy_distance),
      is.finite(geo_distance)
    )

  if (nrow(plot_df) == 0) {
    mantel_results <- rbind(
      mantel_results,
      data.frame(
        species = g,
        tips_used = length(species_asvs),
        pairwise_comparisons = n_pairwise,
        mantel_r = NA_real_,
        mantel_p = NA_real_
      )
    )
    next
  }

  if (length(species_asvs) >= 3) {
    mantel_out <- vegan::mantel(
      as.dist(phy_sub),
      as.dist(geo_sub),
      method = "spearman",
      permutations = 999
    )

    mantel_results <- rbind(
      mantel_results,
      data.frame(
        species = g,
        tips_used = length(species_asvs),
        pairwise_comparisons = n_pairwise,
        mantel_r = unname(mantel_out$statistic),
        mantel_p = mantel_out$signif
      )
    )
  } else {
    mantel_results <- rbind(
      mantel_results,
      data.frame(
        species = g,
        tips_used = length(species_asvs),
        pairwise_comparisons = n_pairwise,
        mantel_r = NA_real_,
        mantel_p = NA_real_
      )
    )
  }

  p <- ggplot(plot_df, aes(x = geo_distance, y = phy_distance)) +
    geom_point(
      shape = 21,
      fill = point_color,
      color = "black",
      alpha = 0.45,
      size = 1.4,
      stroke = 0.2
    ) +
    theme_classic() +
    labs(
      title = g,
      x = "Geo distance (km)",
      y = "Phylo distance"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "italic")
    )

  if (nrow(plot_df) >= 3 && length(unique(plot_df$geo_distance)) >= 2) {
    p <- p +
      geom_smooth(
        method = "lm",
        se = TRUE,
        color = line_color,
        fill = "grey80",
        linewidth = 1,
        alpha = 0.25
      )
  }

  plot_list[[g]] <- p
}

print(diagnostic_df)

mantel_results <- mantel_results %>%
  arrange(mantel_p)

print(mantel_results)

combined_plot <- wrap_plots(plot_list, ncol = 2)

combined_plot

ggsave(
  filename = "top10_species_geo_phylo_multiplot.pdf",
  plot = combined_plot,
  width = 12,
  height = 16,
  units = "in"
)

# Example: view one species plot
plot_list$Lactobacillus_reuteri
```

asv phylo and geo matrix v2: colored by region 
```{r}
# -----------------------------
# Starting object and settings
# -----------------------------

individual_col <- "labtracking"
lat_col <- "decimalLatitude"
lon_col <- "decimalLongitude"
species_col <- "Species"
region_col <- "Region"

summary_fun <- "mean"   # "mean", "min", or "median"
output_units <- "km"    # "km" or "m"

base_colors <- c(
  "within east" = "#5ab4ac",
  "within west" = "#d8b365",
  "between regions (east vs west)" = "#D85F64"
)

line_color <- "grey35"

ambiguous_pattern <- paste(
  c(
    "uncultured",
    "unidentified",
    "metagenome",
    "bacterium",
    "organism",
    "sp\\.",
    "spp\\.",
    "cf\\.",
    "aff\\."
  ),
  collapse = "|"
)

# -----------------------------
# Extract taxonomy and tree tips
# -----------------------------

tax_df <- as.data.frame(tax_table(ps_full), check.names = FALSE)
tax_df$ASV <- rownames(tax_df)

if (is.null(phy_tree(ps_full))) {
  stop("ps_full does not contain a phylogenetic tree.")
}

tree_tips <- phy_tree(ps_full)$tip.label

# -----------------------------
# Keep only named species with ASVs present in the phylogeny
# -----------------------------

tax_named_phylo <- tax_df %>%
  filter(
    !is.na(.data[[species_col]]),
    .data[[species_col]] != "",
    !grepl(ambiguous_pattern, .data[[species_col]], ignore.case = TRUE),
    ASV %in% tree_tips
  )

# -----------------------------
# Rank species by number of ASVs/tips in the phylogeny
# -----------------------------

top10_species <- tax_named_phylo %>%
  group_by(.data[[species_col]]) %>%
  summarize(
    n_tips = n_distinct(ASV),
    .groups = "drop"
  ) %>%
  arrange(desc(n_tips)) %>%
  slice_head(n = 10)

print(top10_species)

# -----------------------------
# Subset phyloseq object to ASVs from top 10 species
# -----------------------------

top10_asvs <- tax_named_phylo %>%
  filter(.data[[species_col]] %in% top10_species[[species_col]]) %>%
  pull(ASV)

ps_top10_species <- prune_taxa(top10_asvs, ps_full)

# -----------------------------
# Remake phylogenetic distance matrix
# -----------------------------

phy_matrix <- cophenetic(phy_tree(ps_top10_species))

# -----------------------------
# Remake geographic ASV distance matrix
# -----------------------------

asv_tab <- as(otu_table(ps_top10_species), "matrix")

if (!taxa_are_rows(ps_top10_species)) {
  asv_tab <- t(asv_tab)
}

meta <- data.frame(sample_data(ps_top10_species), check.names = FALSE)

needed_cols <- c(individual_col, lat_col, lon_col, region_col)
missing_cols <- setdiff(needed_cols, colnames(meta))

if (length(missing_cols) > 0) {
  stop("Missing metadata columns: ", paste(missing_cols, collapse = ", "))
}

sample_ids <- colnames(asv_tab)

if (!all(sample_ids %in% rownames(meta))) {
  missing_samples <- setdiff(sample_ids, rownames(meta))
  stop(
    "Some OTU-table sample names are missing from sample_data rownames. Examples: ",
    paste(head(missing_samples), collapse = ", ")
  )
}

meta <- meta[sample_ids, , drop = FALSE]

asv_pa_sample <- asv_tab > 0

individuals <- unique(meta[[individual_col]])

asv_by_ind <- sapply(individuals, function(ind) {
  samples_for_ind <- rownames(meta)[meta[[individual_col]] == ind]
  rowSums(asv_pa_sample[, samples_for_ind, drop = FALSE]) > 0
})

asv_by_ind <- as.matrix(asv_by_ind)
mode(asv_by_ind) <- "numeric"

rownames(asv_by_ind) <- rownames(asv_tab)
colnames(asv_by_ind) <- individuals

asv_by_ind <- asv_by_ind[rowSums(asv_by_ind) > 0, , drop = FALSE]

individual_coords <- meta[
  !duplicated(meta[[individual_col]]),
  c(individual_col, lon_col, lat_col, region_col),
  drop = FALSE
]

individual_coords <- individual_coords[
  complete.cases(individual_coords[, c(lon_col, lat_col, region_col)]),
  ,
  drop = FALSE
]

rownames(individual_coords) <- individual_coords[[individual_col]]

common_inds <- intersect(colnames(asv_by_ind), rownames(individual_coords))

asv_by_ind <- asv_by_ind[, common_inds, drop = FALSE]
individual_coords <- individual_coords[common_inds, , drop = FALSE]

ind_coords_mat <- as.matrix(individual_coords[, c(lon_col, lat_col)])

ind_geo_dist <- geosphere::distm(
  ind_coords_mat,
  ind_coords_mat,
  fun = geosphere::distHaversine
)

rownames(ind_geo_dist) <- rownames(individual_coords)
colnames(ind_geo_dist) <- rownames(individual_coords)

asvs <- rownames(asv_by_ind)

asv_geo_dist <- matrix(
  NA_real_,
  nrow = length(asvs),
  ncol = length(asvs),
  dimnames = list(asvs, asvs)
)

summarize_distance <- switch(
  summary_fun,
  mean = mean,
  min = min,
  median = median
)

for (i in seq_along(asvs)) {
  inds_i <- colnames(asv_by_ind)[asv_by_ind[i, ] > 0]

  for (j in seq(i, length(asvs))) {
    inds_j <- colnames(asv_by_ind)[asv_by_ind[j, ] > 0]

    pairwise_distances <- ind_geo_dist[inds_i, inds_j, drop = FALSE]

    d <- summarize_distance(pairwise_distances, na.rm = TRUE)

    asv_geo_dist[i, j] <- d
    asv_geo_dist[j, i] <- d
  }
}

diag(asv_geo_dist) <- 0

if (output_units == "km") {
  asv_geo_dist <- asv_geo_dist / 1000
}

# -----------------------------
# Assign ASVs to regions
# -----------------------------

individual_region <- individual_coords[, c(individual_col, region_col), drop = FALSE]
rownames(individual_region) <- individual_region[[individual_col]]

asv_to_regions <- lapply(rownames(asv_by_ind), function(asv) {
  inds_present <- colnames(asv_by_ind)[asv_by_ind[asv, ] > 0]
  regions_present <- unique(individual_region[inds_present, region_col])
  regions_present <- regions_present[!is.na(regions_present)]
  tolower(regions_present)
})

names(asv_to_regions) <- rownames(asv_by_ind)

# -----------------------------
# Plot one graph per top species and run Mantel tests
# -----------------------------

tax_top10 <- as.data.frame(tax_table(ps_top10_species), check.names = FALSE)
tax_top10$ASV <- rownames(tax_top10)

plot_list <- list()
diagnostic_df <- data.frame()
mantel_results <- data.frame()

for (g in top10_species[[species_col]]) {

  species_asvs <- tax_top10$ASV[tax_top10[[species_col]] == g]

  species_asvs <- Reduce(
    intersect,
    list(
      species_asvs,
      rownames(phy_matrix),
      rownames(asv_geo_dist)
    )
  )

  n_pairwise <- ifelse(
    length(species_asvs) >= 2,
    choose(length(species_asvs), 2),
    0
  )

  diagnostic_df <- rbind(
    diagnostic_df,
    data.frame(
      species = g,
      tips_used = length(species_asvs),
      pairwise_comparisons = n_pairwise
    )
  )

  if (length(species_asvs) < 2) {
    mantel_results <- rbind(
      mantel_results,
      data.frame(
        species = g,
        tips_used = length(species_asvs),
        pairwise_comparisons = n_pairwise,
        mantel_r = NA_real_,
        mantel_p = NA_real_
      )
    )
    next
  }

  phy_sub <- phy_matrix[species_asvs, species_asvs, drop = FALSE]
  geo_sub <- asv_geo_dist[species_asvs, species_asvs, drop = FALSE]

  lower_tri <- lower.tri(phy_sub)

  plot_df <- data.frame(
    ASV1 = rownames(phy_sub)[row(phy_sub)[lower_tri]],
    ASV2 = colnames(phy_sub)[col(phy_sub)[lower_tri]],
    phy_distance = phy_sub[lower_tri],
    geo_distance = geo_sub[lower_tri]
  ) %>%
    filter(
      !is.na(phy_distance),
      !is.na(geo_distance),
      is.finite(phy_distance),
      is.finite(geo_distance)
    )

  if (nrow(plot_df) == 0) {
    mantel_results <- rbind(
      mantel_results,
      data.frame(
        species = g,
        tips_used = length(species_asvs),
        pairwise_comparisons = n_pairwise,
        mantel_r = NA_real_,
        mantel_p = NA_real_
      )
    )
    next
  }

  plot_df$region_pair <- mapply(function(a1, a2) {
    regions_1 <- asv_to_regions[[a1]]
    regions_2 <- asv_to_regions[[a2]]

    combined_regions <- unique(c(regions_1, regions_2))
    combined_regions <- combined_regions[!is.na(combined_regions)]

    if (length(combined_regions) == 1 && combined_regions == "east") {
      "within east"
    } else if (length(combined_regions) == 1 && combined_regions == "west") {
      "within west"
    } else {
      "between regions (east vs west)"
    }
  }, plot_df$ASV1, plot_df$ASV2)

  plot_df$region_pair <- factor(
    plot_df$region_pair,
    levels = c(
      "within east",
      "within west",
      "between regions (east vs west)"
    )
  )

  if (length(species_asvs) >= 3) {
    mantel_out <- vegan::mantel(
      as.dist(phy_sub),
      as.dist(geo_sub),
      method = "spearman",
      permutations = 999
    )

    mantel_results <- rbind(
      mantel_results,
      data.frame(
        species = g,
        tips_used = length(species_asvs),
        pairwise_comparisons = n_pairwise,
        mantel_r = unname(mantel_out$statistic),
        mantel_p = mantel_out$signif
      )
    )
  } else {
    mantel_results <- rbind(
      mantel_results,
      data.frame(
        species = g,
        tips_used = length(species_asvs),
        pairwise_comparisons = n_pairwise,
        mantel_r = NA_real_,
        mantel_p = NA_real_
      )
    )
  }

  p <- ggplot(plot_df, aes(x = geo_distance, y = phy_distance)) +
    geom_point(
      aes(fill = region_pair),
      shape = 21,
      color = "black",
      alpha = 0.45,
      size = 1.4,
      stroke = 0.2
    ) +
    scale_fill_manual(
      values = base_colors,
      name = "Region"
    ) +
    theme_classic() +
    labs(
      title = g,
      x = "Geo distance (km)",
      y = "Phylo distance"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "italic"),
      legend.position = "right"
    )

  if (nrow(plot_df) >= 3 && length(unique(plot_df$geo_distance)) >= 2) {
    p <- p +
      geom_smooth(
        aes(group = 1),
        method = "lm",
        se = TRUE,
        color = line_color,
        fill = "grey80",
        linewidth = 1,
        alpha = 0.25
      )
  }

  plot_list[[g]] <- p
}

print(diagnostic_df)

mantel_results <- mantel_results %>%
  arrange(mantel_p)

print(mantel_results)

combined_plot <- wrap_plots(plot_list, ncol = 2)

combined_plot

ggsave(
  filename = "top10_species_geo_phylo_multiplot.pdf",
  plot = combined_plot,
  width = 12,
  height = 16,
  units = "in"
)

# Example: view one species plot
plot_list$Lactobacillus_reuteri
```

asv phylo and geo matrix v3: groups from the symbiont phylogeny
```{r}
base_colors <- c(
  "within east" = "#5ab4ac",
  "within west" = "#d8b365",
  "between regions (east vs west)" = "#D85F64"
)

line_colors <- darken(
  base_colors,
  0.3
)

# -----------------------------
# Get symbiont tree
# -----------------------------

symbiont_tree <- phy_tree(ps_full)

if (is.null(symbiont_tree)) {
  stop("ps_full does not contain a phylogenetic tree.")
}

# -----------------------------
# Define clades using nodes in distal 1/10th of root depth
# -----------------------------

node_depths <- ape::node.depth.edgelength(symbiont_tree)

n_tips <- length(symbiont_tree$tip.label)
n_nodes <- symbiont_tree$Nnode

tip_depths <- node_depths[1:n_tips]
max_root_depth <- max(tip_depths, na.rm = TRUE)

distal_cutoff <- 0.90 * max_root_depth

internal_nodes <- (n_tips + 1):(n_tips + n_nodes)

distal_nodes <- internal_nodes[node_depths[internal_nodes] >= distal_cutoff]

if (length(distal_nodes) == 0) {
  stop("No internal nodes found in the distal 1/10th of the tree depth.")
}

# -----------------------------
# Rank distal clades by number of tips
# -----------------------------

clade_df <- data.frame()

for (node in distal_nodes) {
  clade_tips <- ape::extract.clade(symbiont_tree, node = node)$tip.label

  clade_df <- rbind(
    clade_df,
    data.frame(
      node = node,
      n_tips = length(clade_tips)
    )
  )
}

clade_df <- clade_df %>%
  arrange(desc(n_tips))

# -----------------------------
# Select top 10 largest non-overlapping distal clades
# -----------------------------

selected_clades <- list()
selected_summary <- data.frame()
used_tips <- character()

for (node in clade_df$node) {
  clade_tips <- ape::extract.clade(symbiont_tree, node = node)$tip.label

  if (length(intersect(clade_tips, used_tips)) > 0) {
    next
  }

  clade_name <- paste0("Clade_", length(selected_clades) + 1, "_Node_", node)

  selected_clades[[clade_name]] <- clade_tips

  selected_summary <- rbind(
    selected_summary,
    data.frame(
      clade = clade_name,
      node = node,
      n_tips = length(clade_tips)
    )
  )

  used_tips <- c(used_tips, clade_tips)

  if (length(selected_clades) == 10) {
    break
  }
}

print(selected_summary)

if (length(selected_clades) == 0) {
  stop("No non-overlapping clades were selected.")
}

# -----------------------------
# Subset phyloseq object to ASVs in top 10 clades
# -----------------------------

top10_clade_asvs <- unique(unlist(selected_clades))
top10_clade_asvs <- intersect(top10_clade_asvs, taxa_names(ps_full))

ps_top10_clades <- prune_taxa(top10_clade_asvs, ps_full)

# -----------------------------
# Remake phylogenetic distance matrix
# -----------------------------

phy_matrix <- cophenetic(phy_tree(ps_top10_clades))

# -----------------------------
# Remake geographic ASV distance matrix
# -----------------------------

asv_tab <- as(otu_table(ps_top10_clades), "matrix")

if (!taxa_are_rows(ps_top10_clades)) {
  asv_tab <- t(asv_tab)
}

meta <- data.frame(sample_data(ps_top10_clades), check.names = FALSE)

needed_cols <- c(individual_col, lat_col, lon_col, region_col)
missing_cols <- setdiff(needed_cols, colnames(meta))

if (length(missing_cols) > 0) {
  stop("Missing metadata columns: ", paste(missing_cols, collapse = ", "))
}

sample_ids <- colnames(asv_tab)

if (!all(sample_ids %in% rownames(meta))) {
  missing_samples <- setdiff(sample_ids, rownames(meta))
  stop(
    "Some OTU-table sample names are missing from sample_data rownames. Examples: ",
    paste(head(missing_samples), collapse = ", ")
  )
}

meta <- meta[sample_ids, , drop = FALSE]

asv_pa_sample <- asv_tab > 0

individuals <- unique(meta[[individual_col]])

asv_by_ind <- sapply(individuals, function(ind) {
  samples_for_ind <- rownames(meta)[meta[[individual_col]] == ind]
  rowSums(asv_pa_sample[, samples_for_ind, drop = FALSE]) > 0
})

asv_by_ind <- as.matrix(asv_by_ind)
mode(asv_by_ind) <- "numeric"

rownames(asv_by_ind) <- rownames(asv_tab)
colnames(asv_by_ind) <- individuals

asv_by_ind <- asv_by_ind[rowSums(asv_by_ind) > 0, , drop = FALSE]

individual_coords <- meta[
  !duplicated(meta[[individual_col]]),
  c(individual_col, lon_col, lat_col, region_col),
  drop = FALSE
]

individual_coords <- individual_coords[
  complete.cases(individual_coords[, c(lon_col, lat_col, region_col)]),
  ,
  drop = FALSE
]

rownames(individual_coords) <- individual_coords[[individual_col]]

common_inds <- intersect(colnames(asv_by_ind), rownames(individual_coords))

asv_by_ind <- asv_by_ind[, common_inds, drop = FALSE]
individual_coords <- individual_coords[common_inds, , drop = FALSE]

ind_coords_mat <- as.matrix(individual_coords[, c(lon_col, lat_col)])

ind_geo_dist <- geosphere::distm(
  ind_coords_mat,
  ind_coords_mat,
  fun = geosphere::distHaversine
)

rownames(ind_geo_dist) <- rownames(individual_coords)
colnames(ind_geo_dist) <- rownames(individual_coords)

asvs <- rownames(asv_by_ind)

asv_geo_dist <- matrix(
  NA_real_,
  nrow = length(asvs),
  ncol = length(asvs),
  dimnames = list(asvs, asvs)
)

summarize_distance <- switch(
  summary_fun,
  mean = mean,
  min = min,
  median = median
)

for (i in seq_along(asvs)) {
  inds_i <- colnames(asv_by_ind)[asv_by_ind[i, ] > 0]

  for (j in i:length(asvs)) {
    inds_j <- colnames(asv_by_ind)[asv_by_ind[j, ] > 0]

    pairwise_distances <- ind_geo_dist[inds_i, inds_j, drop = FALSE]

    d <- summarize_distance(pairwise_distances, na.rm = TRUE)

    asv_geo_dist[i, j] <- d
    asv_geo_dist[j, i] <- d
  }
}

diag(asv_geo_dist) <- 0

if (output_units == "km") {
  asv_geo_dist <- asv_geo_dist / 1000
}

# -----------------------------
# Assign each ASV to east/west region categories
# -----------------------------

individual_region <- individual_coords[, c(individual_col, region_col), drop = FALSE]

rownames(individual_region) <- individual_region[[individual_col]]

asv_to_regions <- lapply(rownames(asv_by_ind), function(asv) {
  inds_present <- colnames(asv_by_ind)[asv_by_ind[asv, ] > 0]
  regions_present <- unique(individual_region[inds_present, region_col])
  regions_present <- regions_present[!is.na(regions_present)]
  tolower(regions_present)
})

names(asv_to_regions) <- rownames(asv_by_ind)

# -----------------------------
# Plot one graph per selected clade
# -----------------------------

plot_list <- list()
diagnostic_df <- data.frame()

for (clade_name in names(selected_clades)) {
  clade_asvs <- selected_clades[[clade_name]]

  clade_asvs <- Reduce(
    intersect,
    list(
      clade_asvs,
      rownames(phy_matrix),
      rownames(asv_geo_dist)
    )
  )

  n_pairwise <- ifelse(
    length(clade_asvs) >= 2,
    choose(length(clade_asvs), 2),
    0
  )

  diagnostic_df <- rbind(
    diagnostic_df,
    data.frame(
      clade = clade_name,
      asvs_used = length(clade_asvs),
      pairwise_comparisons = n_pairwise
    )
  )

  if (length(clade_asvs) < 2) {
    next
  }

  phy_sub <- phy_matrix[clade_asvs, clade_asvs, drop = FALSE]
  geo_sub <- asv_geo_dist[clade_asvs, clade_asvs, drop = FALSE]

  lower_tri <- lower.tri(phy_sub)

  plot_df <- data.frame(
    ASV1 = rownames(phy_sub)[row(phy_sub)[lower_tri]],
    ASV2 = colnames(phy_sub)[col(phy_sub)[lower_tri]],
    phy_distance = phy_sub[lower_tri],
    geo_distance = geo_sub[lower_tri]
  ) %>%
    filter(
      !is.na(phy_distance),
      !is.na(geo_distance),
      is.finite(phy_distance),
      is.finite(geo_distance)
    )

  if (nrow(plot_df) == 0) {
    next
  }

  plot_df$region_pair <- mapply(function(a1, a2) {
    regions_1 <- asv_to_regions[[a1]]
    regions_2 <- asv_to_regions[[a2]]

    combined_regions <- unique(c(regions_1, regions_2))
    combined_regions <- combined_regions[!is.na(combined_regions)]

    if (length(combined_regions) == 1 && combined_regions == "east") {
      "within east"
    } else if (length(combined_regions) == 1 && combined_regions == "west") {
      "within west"
    } else {
      "between regions (east vs west)"
    }
  }, plot_df$ASV1, plot_df$ASV2)

  plot_df$region_pair <- factor(
    plot_df$region_pair,
    levels = c(
      "within east",
      "within west",
      "between regions (east vs west)"
    )
  )

  p <- ggplot(plot_df, aes(x = geo_distance, y = phy_distance)) +
    geom_point(
      aes(fill = region_pair),
      shape = 21,
      color = "black",
      alpha = 0.45,
      size = 1.4,
      stroke = 0.15
    ) +
    scale_fill_manual(
      values = base_colors,
      name = "Region comparison"
    ) +
    theme_classic() +
    labs(
      title = clade_name,
      x = "Geographic distance between ASVs (km)",
      y = "Phylogenetic distance between ASVs"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5),
      legend.position = "right"
    )

  if (nrow(plot_df) >= 3 && length(unique(plot_df$geo_distance)) >= 2) {
    p <- p +
      geom_smooth(
        aes(color = region_pair, fill = region_pair),
        method = "lm",
        se = TRUE,
        linewidth = 1,
        alpha = 0.18
      ) +
      scale_color_manual(
        values = line_colors,
        name = "Region comparison"
      )
  }

  plot_list[[clade_name]] <- p
}

print(diagnostic_df)

combined_plot <- wrap_plots(plot_list, ncol = 2)

combined_plot
```

ASV phylo signal 
```{r}
library(phylosignal)
# =========================================================
# EXTRACT TREE
# =========================================================
tree <- as.phylo(asvtree)
tree$node.label <- NULL

# =========================================================
# BUILD RAW EAST / WEST COUNTS
# =========================================================
trait_df <- tree_df %>%
  group_by(OTU) %>%
  summarise(
    east = sum(Region == "east"),
    west = sum(Region == "west"),
    .groups = "drop"
  ) %>%
  mutate(
    total = east + west,
    
    prop_east = ifelse(total > 0, east / total, NA),
    prop_west = ifelse(total > 0, west / total, NA)
  )

trait_df <- trait_df %>%
  mutate(region_bias = prop_east - prop_west)

#Range:
# -1 → fully West
#  0 → mixed
# +1 → fully East

# =========================================================
# ALIGN TO TREE
# =========================================================
trait_df <- trait_df[trait_df$OTU %in% tree$tip.label, ]
trait_df <- trait_df[!duplicated(trait_df$OTU), ]

# IMPORTANT: matrix format required
trait_mat <- trait_df %>%
  dplyr::select(east, west,region_bias) %>%
  as.matrix()

rownames(trait_mat) <- trait_df$OTU

# reorder to match tree
trait_mat <- trait_mat[tree$tip.label, ]

# remove missing tips safely
keep <- !is.na(rowSums(trait_mat))

trait_mat <- trait_mat[keep, ]
tree2 <- drop.tip(tree, tree$tip.label[!keep])

# =========================================================
# MATCH TRAIT VECTOR TO TREE 
# =========================================================

# choose ONE trait for phylosig (must be numeric vector)
trait_vector <- as.numeric(trait_df$region_bias)


# assign names = ASV labels
names(trait_vector) <- trait_df$OTU

# align with tree tips
trait_vector <- trait_vector[tree$tip.label]
var(trait_vector)

# prune tree to match
tree2 <- drop.tip(tree, setdiff(tree$tip.label, names(trait_vector)))

# --- FIX TREE ---
tree2 <- drop.tip(tree, setdiff(tree$tip.label, names(trait_vector)))
tree2$edge.length <- tree2$edge.length + 1e-6

# =========================================================
# PHYLOGENETIC SIGNAL — BLOMBERG’S K
# =========================================================

#K_result <- phylosig(
#  tree2,
#  trait_vector,
#  method = "K",
#  test = TRUE,
#  nsim = 1000
#)

#print(K_result)
#plot(K_result)

# =========================================================
# PHYLOGENETIC SIGNAL — PAGEL’S LAMBDA
# =========================================================

lambda_result <- phylosig(
  tree2,
  trait_vector,
  method = "lambda",
  test = TRUE
)

library(ape)
library(nlme)

dat <- data.frame(
  species = names(trait_vector),
  trait = as.numeric(trait_vector)
)

rownames(dat) <- dat$species

dat <- dat[tree2$tip.label, ]

ll_fun <- function(lambda) {
  fit <- gls(
    trait ~ 1,
    data = dat,
    correlation = corPagel(
      value = lambda,
      phy = tree2,
      fixed = TRUE,
      form = ~ species
    ),
    method = "ML"
  )

  as.numeric(logLik(fit))
}

lambda_hat <- lambda_result$lambda
ll_max <- ll_fun(lambda_hat)

cutoff <- ll_max - qchisq(0.95, df = 1) / 2

root_fun <- function(lambda) {
  ll_fun(lambda) - cutoff
}

lower_95 <- if (root_fun(0) >= 0) {
  0
} else {
  uniroot(root_fun, interval = c(0, lambda_hat))$root
}

upper_95 <- if (root_fun(1) >= 0) {
  1
} else {
  uniroot(root_fun, interval = c(lambda_hat, 1))$root
}

lambda_ci <- c(
  lambda = lambda_hat,
  lower_95 = lower_95,
  upper_95 = upper_95
)

print(lambda_ci)

print(lambda_result)
plot(lambda_result)

#tree2$phy <- multi2di(tree2)

#dev.off()

#dat <- list()
#dat$region_bias <- trait_df$region_bias
#dat <- as.data.frame(dat)

#p4d <- phylo4d(tree2, dat)
#barplot.phylo4d(p4d, tree.type = "phylo", tree.ladderize = TRUE)

#carni.lipa <- lipaMoran(p4d)
#carni.lipa.p4d <- lipaMoran(p4d, as.p4d = TRUE)

#barplot.phylo4d(p4d, bar.col=(carni.lipa$p.value < 0.05) + 1, center = FALSE , scale = FALSE)
#barplot.phylo4d(carni.lipa.p4d, bar.col = (carni.lipa$p.value < 0.05) + 1, center = FALSE, scale = FALSE)

#phyloSignal(p4d = p4d, method = "Lambda")
#phylosim <- phyloSim(tree = tree2, method = "Lambda", nsim = 100, reps = 9)
#plot(phylosim, stacked.methods = FALSE, quantiles = c(0.05, 0.95))
#plot.phylosim(phylosim, what = "pval", stacked.methods = TRUE)
```

dist decay by gen
```{r}
### -----------------------------
### 1. Filter genera (from your top 20)
### -----------------------------
tax <- as.data.frame(tax_table(ps_top10))
region_map <- setNames(meta$Region, meta$labtracking)

top_genera <- unique(tax$Genus)

# Get state vector aligned to samples
state_vec <- meta$stateProvince
names(state_vec) <- sample_names(ps)

# Function: number of states a taxon appears in
taxa_states <- apply(ps_otu, 1, function(x) {
  present_samples <- names(x)[x > 0]
  length(unique(state_vec[present_samples]))
})

# Apply BOTH filters:
# 1. present in >= 5 samples
# 2. present in >= 2 states
keep_taxa <- (rowSums(ps_otu > 0) >= 5) & (taxa_states >= 2)

# Prune phyloseq object
ps3 <- prune_taxa(keep_taxa, ps)

### -----------------------------
### 2. Loop over genera
### -----------------------------
plot_list <- list()
results_list <- list()

for(g in top_genera){

  # Get all ASVs for this genus
  taxa_g <- taxa_names(ps3)[tax$Genus == g]
  if(length(taxa_g) == 0) next

  ps_g <- prune_taxa(taxa_g, ps3)

  # Skip empty phyloseq objects
  if(ntaxa(ps_g) == 0 || nsamples(ps_g) == 0) next

  otu_g <- as(otu_table(ps_g), "matrix")

  if(!taxa_are_rows(ps_g)){
    otu_g <- t(otu_g)
  }

  # Safety checks
  if(nrow(otu_g) == 0 || ncol(otu_g) == 0) next
  if(all(is.na(otu_g))) next
  if(sum(otu_g, na.rm = TRUE) == 0) next

  ### -----------------------------
  ### Bray-Curtis across samples
  ### -----------------------------
  bc <- vegdist(t(otu_g), method = "bray")
  bc_mat <- as.matrix(bc)

  if(all(is.na(bc_mat))) next
  if(nrow(bc_mat) == 0 || ncol(bc_mat) == 0) next

  bc_df <- melt(bc_mat,
                varnames = c("Sample1","Sample2"),
                value.name = "BC_Distance")

  geo_df <- melt(as.matrix(geo_dist),
                 varnames = c("Sample1","Sample2"),
                 value.name = "geo_Distance")

  df <- merge(bc_df, geo_df, by = c("Sample1","Sample2"))

  # Clean
  df <- df %>%
    filter(Sample1 != Sample2) %>%
    mutate(pair_min = pmin(as.character(Sample1), as.character(Sample2)),
           pair_max = pmax(as.character(Sample1), as.character(Sample2))) %>%
    distinct(pair_min, pair_max, .keep_all = TRUE)

  # Convert to similarity for plotting
  df$BC_Similarity <- 1 - df$BC_Distance

  ### -----------------------------
  ### Add region categories
  ### -----------------------------
  df$region1 <- region_map[df$Sample1]
  df$region2 <- region_map[df$Sample2]

  df <- df %>%
    mutate(
      Comparison = case_when(
        region1 == "east" & region2 == "east" ~ "within east",
        region1 == "west" & region2 == "west" ~ "within west",
        region1 != region2 ~ "between regions"))

  df$Genus <- g
  plot_list[[g]] <- df

  ### -----------------------------
  ### Fit models per category
  ### -----------------------------
  slopes <- df %>%
    group_by(Comparison) %>%
    filter(n() > 5) %>%
    summarise(
      slope = coef(lm(BC_Similarity ~ geo_Distance))[2],
      intercept = coef(lm(BC_Similarity ~ geo_Distance))[1],
      r2 = summary(lm(BC_Similarity ~ geo_Distance))$r.squared,
      n_pairs = n(),
      mean_similarity = mean(BC_Similarity, na.rm = TRUE),
      .groups = "drop"
    )

  slopes$Genus <- g
  results_list[[g]] <- slopes
}

### -----------------------------
### 3. Combine outputs
### -----------------------------
plot_df <- bind_rows(plot_list)
genus_results <- bind_rows(results_list)

### -----------------------------
### 5. Plot
### -----------------------------
line_colors <- c(
  "within east" = "#5ab4ac",
  "within west" = "#d8b365",
  "between regions" = "#D85F64"
)

line_colors2 <- darken(line_colors, 0.3)
base_colors  <- lighten(line_colors, 0.3)

ggplot(plot_df, aes(x = geo_Distance, y = BC_Similarity)) +

  geom_point(aes(color = Comparison), size = 1.5, alpha = 0.5) +
  scale_color_manual(values = base_colors, guide = "none") +

  ggnewscale::new_scale_color() +

  geom_smooth(aes(color = Comparison), method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = line_colors2) +

  facet_wrap(~Genus, scales = "free_y") +

  theme_bw() +
  labs(
    x = "Distance (km)",
    y = "Bray-Curtis Similarity",
    color = "Comparison"
  )
```

genus pheno
```{r}
### ---------------------------
### 1. Extract phyloseq data
### ---------------------------

# OTU table (taxa as rows)
otu <- otu_table(ps3)
if (!taxa_are_rows(otu)) otu <- t(otu)
otu <- as(otu, "matrix")

# Metadata
meta <- as.data.frame(sample_data(ps3))

# Taxonomy
tax <- as.data.frame(tax_table(ps3))

# Region vector
region_vec <- meta$Region
names(region_vec) <- rownames(meta)

### ---------------------------
### 2. Sum abundance by region
### ---------------------------

east_samples <- names(region_vec[region_vec == "east"])
west_samples <- names(region_vec[region_vec == "west"])

east_counts <- rowSums(otu[, east_samples, drop = FALSE])
west_counts <- rowSums(otu[, west_samples, drop = FALSE])

asv_counts <- data.frame(
  ASV = rownames(otu),
  east = east_counts,
  west = west_counts
)

# classify ASVs
asv_counts <- asv_counts %>%
  mutate(
    RegionClass = case_when(
      east > 0 & west > 0 ~ "both",
      east > 0 ~ "east",
      west > 0 ~ "west",
      TRUE ~ NA_character_
    )
  )

# join taxonomy
tax$ASV <- rownames(tax)
df <- left_join(asv_counts, tax, by = "ASV")


### ---------------------------
### 4. Load functional (GTDB) data
### ---------------------------

con <- file('gtdb_family_summary.jsonl.gz', open = 'r')
data <- stream_in(con)
close(con)

df_func <- lapply(seq_len(nrow(data)), function(i) {
  fam <- data$tax_name[i]
  s <- data$summaries[[i]]

  s$pct_true <- s$percentages$true
  s$pct_false <- s$percentages$false
  s$percentages <- NULL

  s$family <- fam
  s
}) |> bind_rows()

df_func <- df_func |> select(family, everything())

### ---------------------------
### 5. Filter for gene count trait
### ---------------------------

df_gene_count <- df_func %>%
  filter(name == "genome size") %>%
  filter(family %in% df$Family) 

### ---------------------------
### 6. Merge ecology + function
### ---------------------------

plot_df1 <- df %>%
  left_join(
    df_gene_count %>% 
      select(family, majority_label_gene_count = majority_label),
    by = c("Family" = "family")
  )

# numeric gene count
plot_df1$gene_count <- as.numeric(gsub("[^0-9.]", "", plot_df1$majority_label_gene_count))


### ---------------------------
### 7. Filter for obligate anaerobic
### ---------------------------
unique(df_func$name)

df_ob_ana <- df_func %>%
  filter(name == "oxygen preference") %>%
  filter(family %in% df$Family) %>%
  mutate(
    verdict_ob_ana = case_when(
      grepl("^true", majority_label, ignore.case = TRUE) ~ "true",
      grepl("^false", majority_label, ignore.case = TRUE) ~ "false",
      grepl("No robust majority", majority_label, ignore.case = TRUE) ~ "no_robust_majority"
    )
  )

### ---------------------------
### 8. Merge ecology + obligate anaerobic
### ---------------------------

plot_df1 <- plot_df1 %>%
  left_join(
    df_ob_ana %>% 
      select(family, majority_label_ob_ana = majority_label, verdict_ob_ana, pct_true_ob_ana = pct_true),
    by = c("Family" = "family")
  )

### ---------------------------
### 9. Filter for obligate aerobic
### ---------------------------

df_ob_aer <- df_func %>%
  filter(name == "obligate aerobic") %>%
  filter(family %in% df$Family) %>%
  mutate(
    verdict_ob_aer = case_when(
      grepl("^true", majority_label, ignore.case = TRUE) ~ "true",
      grepl("^false", majority_label, ignore.case = TRUE) ~ "false",
      grepl("No robust majority", majority_label, ignore.case = TRUE) ~ "no_robust_majority"
    )
  )

### ---------------------------
### 8. Merge ecology + obligate aerobic
### ---------------------------

plot_df1 <- plot_df1 %>%
  left_join(
    df_ob_aer %>% 
      select(family, majority_label_ob_aer = majority_label, verdict_ob_aer, pct_true_ob_aer = pct_true),
    by = c("Family" = "family")
  )

### ---------------------------
### 9. Filter for sporulation
### ---------------------------

df_sporulation <- df_func %>%
  filter(name == "sporulation") %>%
  filter(family %in% df$Family) %>%
  mutate(
    verdict_df_sporulation = case_when(
      grepl("^true", majority_label, ignore.case = TRUE) ~ "true",
      grepl("^false", majority_label, ignore.case = TRUE) ~ "false",
      grepl("No robust majority", majority_label, ignore.case = TRUE) ~ "no_robust_majority"
    )
  )

### ---------------------------
### 8. Merge ecology + sporulation
### ---------------------------

plot_df1 <- plot_df1 %>%
  left_join(
    df_sporulation %>% 
      select(family, majority_label_verdict_df_sporulation = majority_label, verdict_df_sporulation, pct_true_verdict_df_sporulation = pct_true),
    by = c("Family" = "family")
  )
```

circle plot
```{r}
# =========================
# 1. TREE + TAXONOMY
# =========================
tree <- phy_tree(ps3)

tax_df <- data.frame(
  ASV = taxa_names(ps3),
  Family = as.character(tax_table(ps3)[, "Family"]),
  stringsAsFactors = FALSE
) %>%
  mutate(
    Family = trimws(Family),
    Family = ifelse(is.na(Family) | Family == "", "Unknown", Family)
  )

# =========================
# 2. DATA CLEANING
# =========================
cap <- function(x, q = 0.98){
  lim <- quantile(x, q, na.rm = TRUE)
  pmin(x, lim)
}

ring_df <- plot_df1 %>%
  mutate(
    ASV = as.character(ASV),
    east = log1p(cap(east)),
    west = log1p(cap(west)),
    gene_count = cap(gene_count),
    obligate_anaerobic = majority_label_ob_ana,
    sporulation = verdict_df_sporulation
  ) %>%
  filter(ASV %in% tree$tip.label) %>%
  mutate(
    RegionClass = case_when(
      tolower(RegionClass) == "east" ~ "East",
      tolower(RegionClass) == "west" ~ "West",
      tolower(RegionClass) == "both" ~ "Both",
      TRUE ~ NA_character_
    )
  )

# =========================
# 3. BASE TREE (FADED)
# =========================
p <- ggtree(
  tree,
  layout = "circular",
  size = 0.15,
  color = "grey80"
) %<+% tax_df

# =========================
# 4. RING GEOMETRY
# =========================
# ring index system
ring_positions <- c(
  #east = 1,
  #west = 2,
  region = 1,
  gene = 2,
  anaerobic = 3,
  sporulation = 4
)

w <- 0.20
# =========================
# 5. DATA RINGS
# =========================

# EAST
#p <- p + new_scale_fill() +
#  geom_fruit(
#    data = ring_df,
#    geom = geom_tile,
#    aes(y = ASV, x = ring_positions["east"] * w + 1, fill = east),
#    width = w,offset = 0  
#  ) +
#  scale_fill_viridis_c(
#    option = "mako",
#    na.value = "white",
#    name = "East (log)"
#  )

# WEST
#p <- p + new_scale_fill() +
#  geom_fruit(
#    data = ring_df,
#    geom = geom_tile,
#    aes(y = ASV, x = ring_positions["west"] * w + 1, fill = west),
#    width = w,offset = 0  
#  ) +
#  scale_fill_viridis_c(
#    option = "rocket",
#    na.value = "white",
#    name = "West (log)"
#  )

# REGION
p <- p + new_scale_fill() +
  geom_fruit(
    data = ring_df,
    geom = geom_tile,
    aes(y = ASV, x = ring_positions["region"] * w + 1, fill = RegionClass),
    width = w
  ) +
  scale_fill_manual(
    values = c(
      "East" = "#1b9e77",
      "West" = "#d95f02",
      "Both" = "#7570b3"
    ),
    name = "Region"
  )

# GENE COUNT
p <- p + new_scale_fill() +
  geom_fruit(
    data = ring_df,
    geom = geom_tile,
    aes(y = ASV, x = ring_positions["gene"] * w + 1, fill = gene_count),
    width = w
  ) +
  scale_fill_viridis_c(
    option = "mako",
    na.value = "lightgray",
    name = "Genome size"
  )

# ANAEROBIC
p <- p + new_scale_fill() +
  geom_fruit(
    data = ring_df,
    geom = geom_tile,
    aes(y = ASV, x = ring_positions["anaerobic"] * w + 1, fill = obligate_anaerobic),
    width = w 
  ) +
  scale_fill_viridis_d(
    na.value = "lightgray",
    name = "Oxygen preference"
  )

# sporulation
p <- p + new_scale_fill() +
  geom_fruit(
    data = ring_df,
    geom = geom_tile,
    aes(y = ASV, x = ring_positions["sporulation"] * w + 1, fill = sporulation),
    width = w 
  ) +
  scale_fill_viridis_d(
    na.value = "lightgray",
    name = "sporulation"
  )

# =========================
# 6. GENUS LABELS (PUBLICATION STYLE)
# =========================

genus_nodes <- tax_df %>% 
  filter(Family != "Unknown") %>% 
  group_by(Family) %>% 
  filter(n() > 1) %>% 
  summarise(node = getMRCA(tree, ASV), .groups = "drop") %>%
  filter(!is.na(node)) 

label_offset <- 1 + 5*0.02

for(i in seq_len(nrow(genus_nodes))){
  
  p <- p +
    geom_cladelab(
      node = genus_nodes$node[i],
      label = genus_nodes$Family[i],
      
      offset = label_offset,     #  push OUTSIDE rings
      offset.text = 0.01,
      
      align = TRUE,              #  key for readability
      linetype = 1,
      barsize = 0.3,
      
      fontsize = 3,            # slightly larger
      fontface = "italic",
      angle = "auto"
    )
}
# =========================
# 7. FINAL THEME
# =========================
p +
  theme_void() +
  theme(
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 8),
    plot.margin = margin(80, 110, 70, 80),  # top, right, bottom, left,
  legend.position = c(1.1, 0.5),  # push legend outside
  legend.justification = c("left", "center")
  )
```

phyloseq import again. did a ton of manipulation and want to start fresh to be sure
```{r}
#---------------------------
# Import data
#---------------------------
physeq <- qza_to_phyloseq(
  features  = "data/microbiome/rarefied-table.qza",
  tree      = "data/microbiome/rooted-tree.qza",
  taxonomy  = "data/microbiome/taxonomy.qza",
  metadata  = "MicrobiotaProcess_sample_data2.tsv"
)

#---------------------------
# Sample filtering
#---------------------------
ps <- physeq %>%
  ps_filter(scientificName == "Peromyscus maniculatus") %>%
  ps_filter(stateProvince != "Arizona")

sample_names(ps) <- sample_data(ps)$labtracking

ps_otu <- otu_table(ps)
sort(rowSums(t(ps_otu)))

ps_bc_dist   <- vegdist(t(ps_otu), method = "bray") 
ps_sor_dist  <- vegdist(t(ps_otu), method = "bray", binary = TRUE)

meta <- data.frame(sample_data(ps))
meta$decimalLatitude <- as.numeric(meta$decimalLatitude)
meta$decimalLongitude <- as.numeric(meta$decimalLongitude)
```

general stats
```{r}
total_asvs <- ntaxa(ps)
cat("Total number of ASVs:", total_asvs, "\n\n")

# Agglomerate at Phylum level
ps_phylum <- tax_glom(ps, taxrank = "Species")

# Transform to relative abundance
ps_phylum_rel <- transform_sample_counts(ps_phylum, 
                                         function(x) x / sum(x) * 100)

# Melt to dataframe
phylum_df <- psmelt(ps_phylum_rel)

# Calculate mean % abundance per region per phylum
phylum_summary <- phylum_df %>%
  group_by(Species) %>%
  summarise(mean_percent = mean(Abundance), .groups = "drop")

print(phylum_summary)

```

% shared ASV
```{r}
library(ggpubr)
#easiset with binary data
ps.pa <- transform_sample_counts(ps, function(x) ifelse(x > 0, 1, 0))
otu <- as(otu_table(ps.pa), "matrix")

if (taxa_are_rows(ps.pa)) {
  otu <- t(otu)
}

# Number of ASVs (richness) per sample
asv_richness <- sample_sums(ps.pa)

# Range across all samples
range_asvs <- range(asv_richness)

cat("ASV richness range across samples:\n")
cat("Min:", range_asvs[1], "\n")
cat("Max:", range_asvs[2], "\n\n")

# Build dataframe with richness + metadata
richness_df <- data.frame(
  SampleID = sample_names(ps.pa),
  ASV_Richness = asv_richness,
  sample_data(ps.pa)
)

# Median and SD by region
region_summary <- richness_df %>%
  group_by(Region) %>%   # change 'region' if your metadata column has a different name
  summarise(
    n_samples = n(),
    median_ASV_richness = median(ASV_Richness),
    mean_ASV_richness = mean(ASV_Richness),
    sd_ASV_richness = sd(ASV_Richness),
    min_ASV_richness = min(ASV_Richness),
    max_ASV_richness = max(ASV_Richness)
  )

print(region_summary)

bray <- vegdist(otu, method = "bray", binary = TRUE)
bray_sim <- 1 - as.matrix(bray)

# 3. Long format + upper triangle only
shared_df <- as.data.frame(as.table(bray_sim)) %>%
  dplyr::filter(as.integer(Var1) < as.integer(Var2)) %>%   # safest filter
  dplyr::mutate(
    Var1 = as.character(Var1),
    Var2 = as.character(Var2),
    PercentShared = Freq * 100
  ) %>%
  dplyr::select(-Freq)

# 4. Join metadata
shared_df <- shared_df %>%
  dplyr::left_join(meta, by = c("Var1" = "labtracking")) %>%
  dplyr::rename(State1 = stateProvince, Region1 = Region) %>%
  dplyr::left_join(meta, by = c("Var2" = "labtracking")) %>%
  dplyr::rename(State2 = stateProvince, Region2 = Region)

# 5. Define comparison categories
shared_df <- shared_df %>%
  dplyr::mutate(
    Comparison = dplyr::case_when(
      Region1 == "east" & Region2 == "east" ~ "within east",
      Region1 == "west" & Region2 == "west" ~ "within west",
      Region1 != Region2 ~ "between regions (east vs west)",
      TRUE ~ NA_character_
    )
  )

summary <- shared_df %>%
  group_by(Comparison) %>%
  summarise(mean_percent = mean(PercentShared), .groups = "drop")

print(summary)

sd <- shared_df %>%
  group_by(Comparison) %>%
  summarise(mean_percent = sd(PercentShared), .groups = "drop")

print(sd)

comparisons <- list(
  c("within east", "within west"),
  c("within east", "between regions (east vs west)"),
  c("within west", "between regions (east vs west)")
)

ggplot(shared_df, aes(x = Comparison, y = PercentShared, fill = Comparison)) +
  geom_violin(trim = FALSE) +
  geom_jitter(aes(color = Comparison),
              width = 0.15, size = 1, shape = 21, color = "black") +
stat_compare_means(
  comparisons = comparisons,
  method = "wilcox.test",
  method.args = list(alternative = "two.sided"),
  label = "p.format"
)+

  theme_bw() +
  labs(
    x = NULL,
    y = "Percent shared ASVs"
  ) +
  scale_fill_manual(values = c(
    "within east" = "#5ab4ac",
    "within west" = "#d8b365",
    "between regions (east vs west)" = "#D85F64"
  )) +
  scale_color_manual(values = c(
    "within east" = "#5ab4ac",
    "within west" = "#d8b365",
    "between regions (east vs west)" = "#D85F64"
  )) +
  theme(
    axis.title.y = element_text(size = 16),
    axis.text.x  = element_blank(),  # show labels now that "all" exists
    axis.text.y  = element_text(size = 14)
  )
```

rasters and such
```{r}
#use the gps points and USGS national land coverage data to assign habitat
point <- as.data.frame(colData(mpse_samples_filtered))
point$decimalLatitude <- as.numeric(point$decimalLatitude)
point$decimalLongitude <- as.numeric(point$decimalLongitude)

p1 <- vect(point, geom = c("decimalLongitude", "decimalLatitude"), crs = "+proj=longlat +datum=WGS84")

#Convert to sf
points_sf <- st_as_sf(p1)  

#raster data from https://www.mrlc.gov/data
raster_obj <- rast("Annual_NLCD_LndCov_2020_CU_C1V1/Annual_NLCD_LndCov_2020_CU_C1V1.tif")
tree_raster <- rast("science_tcc_conus_2025_v2025-6_wgs84/science_tcc_conus_wgs84_v2025-6_20250101_20251231.tif")
#raster data from https://prism.oregonstate.edu/data/
rain_raster <- rast("prism_ppt_us_30s_2020/prism_ppt_us_30s_2020.tif")
mean_temp_raster <- rast("prism_tmean_us_30s_2020/prism_tmean_us_30s_2020.tif")

plot(tree_raster)
plot(raster_obj)
plot(rain_raster)
plot(mean_temp_raster)

pts_sf <- st_transform(points_sf, crs(raster_obj))
pst_sf_rain <- st_transform(points_sf, crs(rain_raster))
pts_sf_temp <- st_transform(points_sf, crs(mean_temp_raster))

# Convert sf → terra vector
pts_vect <- vect(pts_sf)
pts_vect_rain <- vect(pst_sf_rain)
pts_vect_temp <- vect(pts_sf_temp)

# Extract raster values
vals <- extract(raster_obj, pts_vect)
vals_rain <- extract(rain_raster, pts_vect_rain)
vals_temp <- extract(mean_temp_raster, pts_vect_temp)

# Combine with original data
results <- pts_sf |>
  bind_cols(
    land_use = vals[,2]
  )

results$land_use_location <- paste(results$land_use, results$Mississippi_River_Label, sep = " ")
#write_tsv(results, file = "metadata_edit2.tsv")

colData(mpse_samples_filtered)$land_use_location <- results$land_use_location
colData(mpse_samples_filtered)$land_use <- results$land_use

results <- pst_sf_rain |>
  bind_cols(
    prism_ppt_us_30s_2020 = vals_rain[,2]
  )
colData(mpse_samples_filtered)$rain <- results$prism_ppt_us_30s_2020

results <- pts_sf_temp |>
  bind_cols(
    prism_tmean_us_30s_2020 = vals_temp[,2]
  )
colData(mpse_samples_filtered)$temp <- results$prism_tmean_us_30s_2020

# Extract sample data from MPSE object
sample_data_df <- colData(mpse_samples_filtered)

# Export as tab-delimited text file
write.table(sample_data_df, 
            file = "MicrobiotaProcess_sample_data.txt", 
            sep = "\t", 
            quote = FALSE, 
            row.names = TRUE)

```

from gabri for alpha stats
```{r}
otu <- t(otu_table(ps))

#sort the metadata and the otu table
metadata_filtered_sorted <- meta[order(meta$labtracking),]
otu_sorted <- otu[order(rownames(otu)),]

#check the data for order 
unique(rownames(otu_sorted) == metadata_filtered_sorted$labtracking) # OK

### Calculate alpha diversity
metadata_filtered_sorted$Richness.rar<-specnumber(otu_sorted) # Counts otu number
metadata_filtered_sorted$Shannon.rar<-vegan::diversity(otu_sorted, index="shannon") ## calculate Shannon H

kruskal.test(Shannon.rar ~ Region, data = metadata_filtered_sorted)

min_richness <- min(metadata_filtered_sorted$Richness.rar)
max_richness <- max(metadata_filtered_sorted$Richness.rar)

cat("Richness range per sample (Observed ASVs):\n")
cat("Min:", min_richness, "\n")
cat("Max:", max_richness, "\n\n")

# Statistical test (Kruskal-Wallis recommended for richness)
kruskal_richness <- kruskal.test(Richness.rar ~ Region, data = metadata_filtered_sorted)
print(kruskal_richness)

summary <- metadata_filtered_sorted %>%
  group_by(Region) %>%
  summarise(mean_percent = mean(Richness.rar),
            sd_percent = sd(Richness.rar),
           .groups = "drop") 

print(summary)

a = ggplot(metadata_filtered_sorted, aes(x = Region, y = Richness.rar, fill = Region)) +
  geom_boxplot() +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), alpha = 0.7) +
  theme_classic() +
  labs(y = "Number of ASVs", x = "Region") +
  scale_fill_manual(values = east_west_colors)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
a
```

Beta diversity (nmds)
```{r}
set.seed(12)

#ps_bc_dist
#ps_sor_dist

bac.nmds<-metaMDS(ps_sor_dist, k=2, try=999)
bac.nmds$stress #0.21
stress= paste("stress = ", round(bac.nmds$stress, digits = 2))

meta$Axis01<-bac.nmds$points[,1]
meta$Axis02<-bac.nmds$points[,2]

bac_nmds <- ggplot(meta, aes(x = Axis01, y = Axis02)) +
  stat_ellipse(aes(fill = stateProvince), geom = "polygon", alpha = 0.45, level = 0.68) +
  geom_point(aes(color = stateProvince), size = 2) +
  scale_color_manual(values = state_colors, name = "State") +
  scale_fill_manual(values =  state_colors, name = "State") +
  theme_classic() +
  labs(x = "NMDS1", y = "NMDS2") 
  #ggtitle("Sor dissimilarity")
bac_nmds
```

Bayesian Generalized (Non-)Linear Multivariate Multilevel Models
```{r}
#### Create metadata -------------
# geo_dist
# ps_bc_dist
# ps_sor_dist

# =========================
# 1. Melt distance matrices
# =========================

melted_BC <- melt(
  as.matrix(ps_bc_dist),
  varnames = c("Sample1", "Sample2"),
  value.name = "BC_Dissimilarity"
)

melted_SOR <- melt(
  as.matrix(ps_sor_dist),
  varnames = c("Sample1", "Sample2"),
  value.name = "SOR_Dissimilarity"
)

melted_geo <- melt(
  as.matrix(geo_dist),
  varnames = c("Sample1", "Sample2"),
  value.name = "geo_Distance"
)


# =========================
# 2. SAFE merge
# =========================

melted_diffs <- left_join(
  melted_geo,
  melted_BC,
  by = c("Sample1", "Sample2")
) %>%
  left_join(
    melted_SOR,
    by = c("Sample1", "Sample2")
  )


# =========================
# 3. Convert to similarity
# =========================

melted_diffs <- melted_diffs %>%
  mutate(
    BC_Similarity = 1 - BC_Dissimilarity,
    SOR_Similarity = 1 - SOR_Dissimilarity
  )


# =========================
# 4. Remove duplicates + self-comparisons
# =========================

melted_diffs <- melted_diffs %>%
  filter(Sample1 != Sample2) %>%
  mutate(
    pair_min = pmin(
      as.character(Sample1),
      as.character(Sample2)
    ),
    pair_max = pmax(
      as.character(Sample1),
      as.character(Sample2)
    )
  ) %>%
  distinct(
    pair_min,
    pair_max,
    .keep_all = TRUE
  ) %>%
  select(
    Sample1,
    Sample2,
    geo_Distance,
    BC_Similarity,
    SOR_Similarity
  )


# =========================
# 5. Add region metadata
# =========================

region_map <- setNames(
  meta$Region,
  meta$labtracking
)

melted_diffs <- melted_diffs %>%
  mutate(
    region1 = region_map[Sample1],
    region2 = region_map[Sample2],

    compairson = case_when(
      region1 == "east" &
        region2 == "east" ~
        "within east",

      region1 == "west" &
        region2 == "west" ~
        "within west",

      TRUE ~
        "between regions (east vs west)"
    ),

    same_region = region1 == region2
  )


# =========================
# 6. Scale distance
# =========================

range.use <- function(
    x,
    min.use,
    max.use
) {

  (x - min(x, na.rm = TRUE)) /
    (
      max(x, na.rm = TRUE) -
        min(x, na.rm = TRUE)
    ) *
    (max.use - min.use) +
    min.use
}

melted_diffs <- melted_diffs %>%
  mutate(
    dist_scaled = range.use(
      geo_Distance,
      0,
      1
    ),

    dist_100 = geo_Distance / 100,

    dist_z = as.numeric(
      scale(dist_100)
    )
  )


# =========================
# 7. Clean factor levels
# =========================

melted_diffs$compairson <- factor(
  melted_diffs$compairson,
  levels = c(
    "within east",
    "within west",
    "between regions (east vs west)"
  )
)

# =========================
# 8. Colors
# =========================

base_colors <- c(
  "within east" = "#5ab4ac",
  "within west" = "#d8b365",
  "between regions (east vs west)" = "#D85F64"
)

line_colors <- c(
  darken(
    base_colors,
    0.3
  ),
  "all" = "grey35"
)


# =========================
# 9. Common axis limits
# =========================

xlims <- range(
  melted_diffs$geo_Distance,
  na.rm = TRUE
)

bc_ylims <- range(
  melted_diffs$BC_Similarity,
  na.rm = TRUE
)

sc_ylims <- range(
  melted_diffs$SOR_Similarity,
  na.rm = TRUE
)


# =========================
# 10. Plotting function
# =========================

plot_decay <- function(
    dat,
    yvar,
    ylab,
    ylims
) {

  ggplot(
    dat,
    aes(
      x = geo_Distance,
      y = .data[[yvar]]
    )
  ) +

    # points use base colors, no outlines
    geom_point(
      aes(color = compairson),
      size = 2,
      alpha = 0.5
    ) +

    scale_color_manual(
      values = base_colors,
      guide = "none",
      drop = FALSE
    ) +

    ggnewscale::new_scale_color() +

    # within east, within west, and between regression lines
    geom_smooth(
      aes(color = compairson),
      method = "lm",
      se = FALSE,
      linewidth = 1.2
    ) +

    # all-data grey regression line on top
    geom_smooth(
      aes(color = "all"),
      method = "lm",
      se = FALSE,
      linewidth = 1.4
    ) +

    scale_color_manual(
      values = line_colors,
      name = "Comparison",
      breaks = c(
        "within east",
        "within west",
        "between regions (east vs west)",
        "all"
      ),
      labels = c(
        "within east",
        "within west",
        "between regions (east vs west)",
        "all"
      )
    ) +

    coord_cartesian(
      xlim = xlims,
      ylim = ylims
    ) +

    labs(
      x = "distance (km)",
      y = ylab
    ) +

    theme_bw() +

    theme(
      axis.title = element_text(size = 13),
      axis.text = element_text(size = 11),
      legend.title = element_text(size = 12),
      legend.text = element_text(size = 11),
      panel.grid.minor = element_blank()
    )
}


# =========================
# 11. Make SC and BC plots
# =========================

p_sc <- plot_decay(
  dat = melted_diffs,
  yvar = "SOR_Similarity",
  ylab = "Sørensen Similarity",
  ylims = sc_ylims
)

p_bc <- plot_decay(
  dat = melted_diffs,
  yvar = "BC_Similarity",
  ylab = "Bray-Curtis Similarity",
  ylims = bc_ylims
)


# =========================
# 12. Combine in one figure
# BC = A, SC = B
# =========================

library(patchwork)

combined_plot <- p_bc + p_sc +
  plot_layout(
    ncol = 2,
    guides = "collect"
  ) +
  plot_annotation(
    tag_levels = "A"
  ) &
  theme(
    legend.position = "right"
  )

combined_plot

# =========================
# Mantel:
#    Geographic distance
#    vs Bray-Curtis
# =========================


mantel_BC<-vegan::mantel(vegan::vegdist(bc_mat), vegan::vegdist(geo_mat),method = "spearman", permutations = 999, parallel = 8)

# =========================
# Mantel:
#    Geographic distance
#    vs Sørensen
# =========================
mantel_SOR<-vegan::mantel(vegan::vegdist(sor_mat), vegan::vegdist(geo_mat),method = "spearman", permutations = 999, parallel = 8)

# =========================
# Results
# =========================

mantel_BC$signif
mantel_BC$statistic

mantel_SOR$signif
mantel_SOR$statistic

### =========================
### 10. Prepare for BRMS 
### =========================
# Avoid exact 0/1 for Beta regression
n <- nrow(melted_diffs)
melted_diffs$BC_Similarity <- (melted_diffs$BC_Similarity * (n - 1) + 0.5) / n

### =========================
### 11. Model (FIXED)
### =========================
#library(brms)

#model <- brm(
#  BC_Similarity ~ 1 + dist_scaled * compairson +
#    (1 | mm(Sample1, Sample2)),
#  data = melted_diffs,
#  family = Beta(),
#  warmup = 2000,
#  iter = 6000,
#  chains = 4,
#  cores = 4,              # safer default; increase if stable
#  threads = threading(4),
#  control = list(adapt_delta = 0.95),
#  init = 0
#)
#summary(model)

#Scale distance within region
#Use || for random slopes
#Add priors + increase adapt_delta

# || removes slope–intercept correlation
#within-region scaling removes confounding
#priors regularize weakly identified slopes
#distance is still biologically meaningful

#You’ll get:
#distance–decay slope for within-region pairs
#distance–decay slope for across-region pairs

#If:
#slopes differ → region-specific decay
#slopes similar but intercepts differ → Mississippi as a barrier
#across-region slope = 0 → distance irrelevant once region differs
```

distance decay of overlapping distances only 
```{r}
east_range <- range(
  melted_diffs$geo_Distance[
    melted_diffs$compairson == "within east"
  ],
  na.rm = TRUE
)

west_range <- range(
  melted_diffs$geo_Distance[
    melted_diffs$compairson == "within west"
  ],
  na.rm = TRUE
)

# distance range of between-region comparisons
between_range <- range(
  melted_diffs$geo_Distance[
    melted_diffs$compairson == "between regions (east vs west)"
  ],
  na.rm = TRUE
)

between_min <- between_range[1]
between_max <- east_range[2]

### =========================
### Keep only overlapping distances
### =========================

overlap_df <- melted_diffs %>%
  filter(
    geo_Distance >= between_min,
    geo_Distance <= between_max
  )

### =========================
### Pairwise comparisons
### =========================

comparisons <- list(
  c("within east", "within west"),
  c("within east", "between regions (east vs west)"),
  c("within west", "between regions (east vs west)")
)

### =========================
### Box plot
### =========================

p_overlap <- ggplot(
  overlap_df,
  aes(x = compairson, y = BC_Similarity, fill = compairson)
) +

  geom_boxplot(
    alpha = 0.7,
    outlier.alpha = 0.4,
    width = 0.7,
    color = "black"
  ) +

  stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    method.args = list(alternative = "two.sided"),
    label = "p.format"
  ) +

  scale_fill_manual(values = base_colors, name = "Comparison") +

  labs(
    x = NULL,
    y = "Sor Similarity"
  ) +

  theme_bw() +

  theme(
    axis.text.x = element_text(angle = 15, hjust = 1)
  )

print(p_overlap)
```

permanovas
```{r}
#========================================================
# ASV-LEVEL
#========================================================
ps_asv_otu <- as(otu_table(ps), "matrix")

if (!taxa_are_rows(ps)) {
  ps_asv_otu <- t(ps_asv_otu)
}

# Bray-Curtis
permanova_asv_bray <- adonis2(
  t(ps_asv_otu) ~ stateProvince + decimalLatitude + decimalLongitude,
  data = meta,
  permutations = 999,
  method = "bray",
  by = "terms"
)

# Bray-Curtis
permanova_asv_bray <- adonis2(
  t(ps_asv_otu) ~ Region + habitat + temp  + decimalLatitude + decimalLongitude,
  data = meta,
  permutations = 999,
  method = "bray",
  by = "terms"
)

# Sørensen
permanova_asv_sor <- adonis2(
  t(ps_asv_otu) ~ stateProvince + decimalLatitude + decimalLongitude,
  data = meta,
  permutations = 999,
  method = "bray",
  binary = TRUE,
  by = "terms"
)

permanova_asv_sor <- adonis2(
  t(ps_asv_otu) ~ Region + habitat + temp  + decimalLatitude + decimalLongitude,
  data = meta,
  permutations = 999,
  method = "bray",
  binary = TRUE,
  by = "terms"
)


#========================================================
# GENUS-LEVEL
#========================================================
ps_genus <- tax_glom(ps, taxrank = "Genus")

# remove missing genera
ps_genus <- subset_taxa(ps_genus, !is.na(Genus))

genus_otu <- as(otu_table(ps_genus), "matrix")

if (!taxa_are_rows(ps_genus)) {
  genus_otu <- t(genus_otu)
}

# Bray-Curtis
permanova_genus_bray <- adonis2(
  t(genus_otu) ~ Region + habitat + temp  + decimalLatitude + decimalLongitude,
  data = meta,
  permutations = 999,
  method = "bray",
  by = "terms"
)

# Sørensen
permanova_genus_sor <- adonis2(
  t(genus_otu) ~ Region + habitat + temp  + decimalLatitude + decimalLongitude,
  data = meta,
  permutations = 999,
  method = "bray",
  binary = TRUE,
  by = "terms"
)

#========================================================
# SPECIES-LEVEL
#========================================================
ps_species <- tax_glom(ps, taxrank = "Species")

# remove missing species labels
ps_species <- subset_taxa(ps_species, !is.na(Species))

species_otu <- as(otu_table(ps_species), "matrix")

if (!taxa_are_rows(ps_species)) {
  species_otu <- t(species_otu)
}

# Bray-Curtis
permanova_species_bray <- adonis2(
  t(species_otu) ~ Region + habitat + temp  + decimalLatitude + decimalLongitude,
  data = meta,
  permutations = 999,
  method = "bray",
  by = "terms"
)

# Sørensen
permanova_species_sor <- adonis2(
  t(species_otu) ~ Region + habitat + temp  + decimalLatitude + decimalLongitude,
  data = meta,
  permutations = 999,
  method = "bray",
  binary = TRUE,
  by = "terms"
)
```

mantel tests
```{r}
# =========================
# 1. Region map + groups
# =========================
region_map <- setNames(meta$Region, meta$labtracking)

east_samples <- names(region_map[region_map == "east"])
west_samples <- names(region_map[region_map == "west"])

# =========================
# 2. Mantel function
# =========================
run_mantel <- function(dist1, dist2) {
  vegan::mantel(dist1, dist2, permutations = 9999, method = "spearman")
}

mantel_bc <- run_mantel(
  as.dist(as.matrix(ps_bc_dist)), as.dist(as.matrix(geo_dist)))
  
mantel_bc$statistic
mantel_bc$signif

mantel_sor <- run_mantel(
  as.dist(as.matrix(ps_sor_dist)), as.dist(as.matrix(geo_dist)))

mantel_sor$statistic
mantel_sor$signif
  
# =========================
# 3. EAST
# =========================
mantel_bc_east <- run_mantel(
  as.dist(as.matrix(ps_bc_dist)[east_samples, east_samples]),
  as.dist(as.matrix(geo_dist)[east_samples, east_samples])
)

mantel_sor_east <- run_mantel(
  as.dist(as.matrix(ps_sor_dist)[east_samples, east_samples]),
  as.dist(as.matrix(geo_dist)[east_samples, east_samples])
)

# =========================
# 4. WEST
# =========================
mantel_bc_west <- run_mantel(
  as.dist(as.matrix(ps_bc_dist)[west_samples, west_samples]),
  as.dist(as.matrix(geo_dist)[west_samples, west_samples])
)

mantel_sor_west <- run_mantel(
  as.dist(as.matrix(ps_sor_dist)[west_samples, west_samples]),
  as.dist(as.matrix(geo_dist)[west_samples, west_samples])
)

# =========================
# 5. BETWEEN REGIONS 
# =========================

bc_mat  <- as.matrix(ps_bc_dist)
sor_mat <- as.matrix(ps_sor_dist)
geo_mat <- as.matrix(geo_dist)

samples <- rownames(bc_mat)

pairs <- expand.grid(samples, samples)

region1 <- region_map[pairs$Var1]
region2 <- region_map[pairs$Var2]

keep <- region1 != region2   # EAST vs WEST only

bc_vals <- bc_mat[cbind(pairs$Var1, pairs$Var2)][keep]
sor_vals <- sor_mat[cbind(pairs$Var1, pairs$Var2)][keep]
geo_vals <- geo_mat[cbind(pairs$Var1, pairs$Var2)][keep]

between_bc<-vegan::mantel(vegan::vegdist(bc_vals), vegan::vegdist(geo_vals),method = "spearman", permutations = 999, parallel = 8)

between_bc$statistic
between_bc$signif

between_sor<-vegan::mantel(vegdist(sor_vals), vegdist(geo_vals),method = "spearman", permutations = 999, parallel = 8)

between_sor$statistic
between_sor$signif
# =========================
# 6. RESULTS TABLE
# =========================
results <- data.frame(
  group = c("east", "west", "between"),
  
  bc_stat = c(
    mantel_bc_east$statistic,
    mantel_bc_west$statistic,
    between_bc$statistic
  ),
  
  bc_p = c(
    mantel_bc_east$signif,
    mantel_bc_west$signif,
    between_bc$p.value
  ),
  
  sor_stat = c(
    mantel_sor_east$statistic,
    mantel_sor_west$statistic,
    between_sor$statistic
  ),
  
  sor_p = c(
    mantel_sor_east$signif,
    mantel_sor_west$signif,
    between_sor$p.value
  )
)

results
```

mantel tests for overlap
```{r}
# =========================
# 1. Region map + groups
# =========================

region_map <- setNames(meta$Region, meta$labtracking)

east_samples <- names(region_map[region_map == "east"])
west_samples <- names(region_map[region_map == "west"])

# =========================
# 2. Overlap distance range
# =========================

east_range <- range(
  melted_diffs$geo_Distance[
    melted_diffs$compairson == "within east"
  ],
  na.rm = TRUE
)

west_range <- range(
  melted_diffs$geo_Distance[
    melted_diffs$compairson == "within west"
  ],
  na.rm = TRUE
)

between_range <- range(
  melted_diffs$geo_Distance[
    melted_diffs$compairson ==
      "between regions (east vs west)"
  ],
  na.rm = TRUE
)

between_min <- between_range[1]

between_max <- min(
  east_range[2],
  west_range[2]
)

# =========================
# 3. Mantel helper
# =========================

run_overlap_test <- function(
    microbe_dist,
    geo_dist,
    sample_subset = NULL,
    between = FALSE,
    region_map = NULL,
    min_dist,
    max_dist
) {

  microbe_mat <- as.matrix(microbe_dist)
  geo_mat <- as.matrix(geo_dist)

  # -------------------------
  # WITHIN REGION
  # -------------------------

  if (!between) {

    microbe_mat <- microbe_mat[
      sample_subset,
      sample_subset
    ]

    geo_mat <- geo_mat[
      sample_subset,
      sample_subset
    ]

    upper <- upper.tri(geo_mat)

    microbe_vals <- microbe_mat[upper]
    geo_vals <- geo_mat[upper]

  } else {

    # -------------------------
    # BETWEEN REGIONS
    # -------------------------

    samples <- rownames(microbe_mat)

    pairs <- expand.grid(
      sample1 = samples,
      sample2 = samples
    )

    region1 <- region_map[pairs$sample1]
    region2 <- region_map[pairs$sample2]

    keep_between <- region1 != region2

    microbe_vals <- microbe_mat[
      cbind(pairs$sample1, pairs$sample2)
    ][keep_between]

    geo_vals <- geo_mat[
      cbind(pairs$sample1, pairs$sample2)
    ][keep_between]
  }

  # -------------------------
  # overlap filter
  # -------------------------

  overlap_keep <- geo_vals >= min_dist &
                  geo_vals <= max_dist

  microbe_vals <- microbe_vals[overlap_keep]
  geo_vals <- geo_vals[overlap_keep]

  # remove NA
  complete_keep <- complete.cases(
    microbe_vals,
    geo_vals
  )

  microbe_vals <- microbe_vals[complete_keep]
  geo_vals <- geo_vals[complete_keep]

  # -------------------------
  # permutation correlation
  # -------------------------

  vegan::mantel(
    dist(microbe_vals),
    dist(geo_vals),
    method = "spearman",
    permutations = 9999
  )
}

# =========================
# 4. EAST
# =========================

mantel_bc_east <- run_overlap_test(
  ps_bc_dist,
  geo_dist,
  sample_subset = east_samples,
  min_dist = between_min,
  max_dist = between_max
)

mantel_sor_east <- run_overlap_test(
  ps_sor_dist,
  geo_dist,
  sample_subset = east_samples,
  min_dist = between_min,
  max_dist = between_max
)

# =========================
# 5. WEST
# =========================

mantel_bc_west <- run_overlap_test(
  ps_bc_dist,
  geo_dist,
  sample_subset = west_samples,
  min_dist = between_min,
  max_dist = between_max
)

mantel_sor_west <- run_overlap_test(
  ps_sor_dist,
  geo_dist,
  sample_subset = west_samples,
  min_dist = between_min,
  max_dist = between_max
)

# =========================
# 6. BETWEEN REGIONS
# =========================

between_bc <- run_overlap_test(
  ps_bc_dist,
  geo_dist,
  between = TRUE,
  region_map = region_map,
  min_dist = between_min,
  max_dist = between_max
)

between_sor <- run_overlap_test(
  ps_sor_dist,
  geo_dist,
  between = TRUE,
  region_map = region_map,
  min_dist = between_min,
  max_dist = between_max
)

# =========================
# 7. RESULTS TABLE
# =========================

results <- data.frame(
  group = c("east", "west", "between"),

  bc_stat = c(
    mantel_bc_east$statistic,
    mantel_bc_west$statistic,
    between_bc$statistic
  ),

  bc_p = c(
    mantel_bc_east$signif,
    mantel_bc_west$signif,
    between_bc$signif
  ),

  sor_stat = c(
    mantel_sor_east$statistic,
    mantel_sor_west$statistic,
    between_sor$statistic
  ),

  sor_p = c(
    mantel_sor_east$signif,
    mantel_sor_west$signif,
    between_sor$signif
  )
)

results
```

physical distance matrix using GPS
```{r}
gps <- as.data.frame(mpse_samples_filtered@colData)[, c("labtracking","stateProvince","decimalLatitude","decimalLongitude","Region")]

locations <- unique(gps$stateProvince)

####################
# Physical distance#
####################
sf_points <- st_as_sf(
  gps,
  coords = c("decimalLongitude", "decimalLatitude"),
  crs = 4326  # WGS84
)

dist_matrix <- st_distance(sf_points)

# Set row/column names
rownames(dist_matrix) <- gps$labtracking
colnames(dist_matrix) <- gps$labtracking

dist_object <- as.dist(dist_matrix)

# Perform hierarchical clustering 
hc_tree <- hclust(dist_object, method = "complete") 
hc_phylo <- as.phylo(hc_tree)

# Create the expanded tree
expanded_tree <- hc_phylo
for(stateProvince in hc_phylo$tip.label) {
  # Find all samples for this stateProvince
  samples <- gps$labtracking[gps$stateProvince == stateProvince]
  
  if(length(samples) > 0) {
    # Find the tip position for this stateProvince
    tip_position <- which(expanded_tree$tip.label == stateProvince)
    
    if(length(samples) == 1) {
      # If only one sample, just rename the tip
      expanded_tree$tip.label[tip_position] <- samples[1]
    } else {
      # If multiple samples, we need to add them as a polytomy
      # First, rename the original tip to the first sample
      expanded_tree$tip.label[tip_position] <- samples[1]
      
      # Then add the rest as sister tips (with very small branch length)
      for(i in 2:length(samples)) {
        expanded_tree <- phytools::bind.tip(expanded_tree, 
                                            tip.label = samples[i],
                                            where = tip_position,
                                            position = 0.0001)  # tiny branch length
      }
    }
  }
}

plot(expanded_tree)

# Match tip labels to rows in gps
tip_index <- match(expanded_tree$tip.label, gps$labtracking)

# Get habitats for all tips (NA if no match)
tip_habitats <- as.character(gps$Region[tip_index])

# Look up colors in habitat_colors
tip_colors <- east_west_colors[tip_habitats]

# Replace any missing colors with black
tip_colors[is.na(tip_colors)] <- "black"

#build tip color vector from your state/colors (must be same length as nrow(assoc))
gps$tip_colors <- east_west_colors[
  match(gps$land_use, names(east_west_colors))
]
names(gps$tip_colors) <- gps$labtracking

phy_tree <- as.phylo(expanded_tree)

# Plot the tree with colored tips
plot(phy_tree, 
     tip.color = tip_colors,
     cex = 0.6,
     no.margin = TRUE)
```

mantel test for whole dataset
```{r}
####################
# Physical distance#
####################
sor_mat <- as.matrix(mpse_bc_dist)[labs, labs]
sor_mat <- sor_mat[rownames(dist_object), colnames(dist_object)]

unique(rownames(dist_object) == rownames(sor_mat)) # ok 
unique(colnames(dist_object) == colnames(sor_mat)) # ok

mt = mantel(sor_mat,dist_object, method = "spearman")
mt$statistic
mt$signif

dist_object_no_zero = dist_object
mt = cultevo::mantel.test(dist(sor_mat),dist(dist_object_no_zero), omitzerodistances = TRUE, trials = 999)
mt

#### plot the mantel tests 
# Convert to long data frames for plotting
df_with_zeros <- data.frame(
  x = as.vector(dist_object),
  y = as.vector(mpse_sor_dist)
)

df_no_zeros <- data.frame(
  x = as.vector(dist_object_no_zero),
  y = as.vector(mpse_sor_dist)
)
df_no_zeros <- df_no_zeros[df_no_zeros$x > 0,]

# Plot with zeros
p1 <- ggplot(df_with_zeros, aes(x = x, y = y)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", col = "blue") +
  labs(
    x = "physical distance",
    y = "Microbiome distance")+
  theme_minimal()

# Plot without zeros
p2 <- ggplot(df_no_zeros, aes(x = x, y = y)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", col = "#8c510a") +
  labs(
    x = "Geograohic distance",
    y = "Microbiome distance"
  ) +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```


vignette used for code chunks below: https://bioconductor.org/packages/release/bioc/vignettes/MicrobiotaProcess/inst/doc//MicrobiotaProcess.html

Rare Abundance curve
```{r}
mpse_samples_filtered %<>% mp_rrarefy()

mpse_samples_filtered %<>% 
    mp_cal_rarecurve(
        .abundance = RareAbundance,
        chunks = 400
    )
# default will display the confidence interval around smooth.
# se=TRUE
p1 <- mpse_samples_filtered %>% 
      mp_plot_rarecurve(
         .rare = RareAbundanceRarecurve, 
         .alpha = Observe,
      )

p2 <- mpse_samples_filtered %>% 
      mp_plot_rarecurve(
         .rare = RareAbundanceRarecurve, 
         .alpha = Observe, 
         .group = Region)
      #) +
      #scale_color_manual(values=c("#00A087FF", "#3C5488FF")) +
      #scale_fill_manual(values=c("#00A087FF", "#3C5488FF"), guide="none")

# combine the samples belong to the same groups
p3 <- mpse_samples_filtered %>% 
      mp_plot_rarecurve(
         .rare = RareAbundanceRarecurve, 
         .alpha = "Observe", 
         .group = Region, 
         plot.group = TRUE)
      #) +
       #scale_color_manual(values=c("#00A087FF", "#3C5488FF")) +
       #scale_fill_manual(values=c("#00A087FF", "#3C5488FF"),guide="none")  

p1 
p2
p3
```

calculate alpha index and visualization
```{r}
mpse_samples_filtered %<>% 
    mp_cal_alpha(.abundance=RareAbundance)

f1 <- mpse_samples_filtered %>% 
      mp_plot_alpha(
        .group=Region, 
        .alpha=c(Observe, Chao1, ACE, Shannon, Simpson, Pielou))
      #) +
      #scale_fill_manual(values=c("#00A087FF", "#3C5488FF"), guide="none") +
      #scale_color_manual(values=c("#00A087FF", "#3C5488FF"), guide="none")

f2 <- mpse_samples_filtered %>%
      mp_plot_alpha(
        .alpha=c(Observe, Chao1, ACE, Shannon, Simpson, Pielou)
      )

f1 / f2
```

Beta diversity analysis
```{r}
mpse_samples_filtered %<>% 
    mp_decostand(.abundance=Abundance)

mpse_samples_filtered %<>% mp_cal_dist(.abundance=hellinger, distmethod="bray")

p2 <- mpse_samples_filtered %>% mp_plot_dist(.distmethod = bray, .group = Region)
p2

p3 <- mpse_samples_filtered %>% mp_plot_dist(.distmethod = bray, .group = Region, group.test=T, textsize=2) +
  scale_fill_manual(values = c(
    "#5ab4ac",
    "#8c510a",
    "#d8b365" ))+
  scale_color_manual(values = c(
    "#5ab4ac",
    "#8c510a",
    "#d8b365" ))

 p3 
```

PCoAs
```{r}
make_pcoa_df <- function(ps, binary = FALSE) {
  otu <- as(otu_table(ps), "matrix")

  if (taxa_are_rows(ps)) {
    otu <- t(otu)
  }

  dist <- vegan::vegdist(otu, method = "bray", binary = binary)
  pcoa_res <- cmdscale(dist, eig = TRUE, k = 2)

  var_exp <- round(
    100 * pcoa_res$eig / sum(pcoa_res$eig[pcoa_res$eig > 0]),
    2
  )

  pcoa_df <- data.frame(
    SampleID = rownames(pcoa_res$points),
    PCo1 = -1 * pcoa_res$points[, 1],
    PCo2 = pcoa_res$points[, 2]
  )

  meta <- data.frame(sample_data(ps))
  meta$SampleID <- rownames(meta)

  pcoa_df <- dplyr::left_join(pcoa_df, meta, by = "SampleID")

  list(
    data = pcoa_df,
    var_exp = var_exp
  )
}

make_temp_group <- function(x) {
  x <- as.numeric(as.character(x))
  breaks <- quantile(x, probs = c(0, 1 / 3, 2 / 3, 1), na.rm = TRUE)

  cut(
    x,
    breaks = unique(breaks),
    labels = c("Low", "Medium", "High")[seq_len(length(unique(breaks)) - 1)],
    include.lowest = TRUE
  )
}

plot_pcoa_group <- function(pcoa_obj,
                            group_var,
                            title,
                            colors,
                            legend_title = group_var) {
  pcoa_df <- pcoa_obj$data
  var_exp <- pcoa_obj$var_exp

  pcoa_df$PlotGroup <- pcoa_df[[group_var]]
  pcoa_df <- pcoa_df %>% dplyr::filter(!is.na(PlotGroup))

  ggplot(
    pcoa_df,
    aes(PCo1, PCo2, fill = PlotGroup, color = PlotGroup)
  ) +
    geom_hline(
      yintercept = 0,
      linetype = "dashed",
      color = "grey45",
      linewidth = 0.4
    ) +
    geom_vline(
      xintercept = 0,
      linetype = "dashed",
      color = "grey45",
      linewidth = 0.4
    ) +
    stat_ellipse(
      aes(group = PlotGroup),
      geom = "polygon",
      linewidth = 0.4,
      level = 0.80,
      alpha = 0.20
    ) +
    geom_point(
      size = 3,
      alpha = 0.95,
      stroke = 0.4,
      shape = 21
    ) +
    scale_fill_manual(values = colors, name = legend_title) +
    scale_color_manual(values = colors, name = legend_title) +
    labs(
      x = paste0("PCo1 (", var_exp[1], "%)"),
      y = paste0("PCo2 (", var_exp[2], "%)"),
      title = title
    ) +
    theme_classic(base_size = 12) +
    theme(
      panel.border = element_rect(
        color = "grey45",
        fill = NA,
        linewidth = 0.8
      ),
      axis.line = element_blank(),
      legend.position = "right",
      legend.title = element_text(size = 11),
      legend.text = element_text(size = 10),
      axis.title = element_text(size = 11),
      axis.text = element_text(size = 9),
      panel.grid = element_blank(),
      plot.margin = margin(10, 10, 10, 10)
    ) +
    guides(
      color = guide_legend(
        override.aes = list(shape = 21, size = 4, alpha = 1)
      ),
      fill = guide_legend(
        override.aes = list(shape = 21, size = 4, alpha = 1)
      )
    )
}

# -----------------------------
# Build PCoA objects
# -----------------------------

bc_pcoa <- make_pcoa_df(ps, binary = FALSE)
sc_pcoa <- make_pcoa_df(ps, binary = TRUE)

# -----------------------------
# Prepare grouping variables
# -----------------------------

bc_pcoa$data$Region <- factor(bc_pcoa$data$Region, levels = c("east", "west"))
sc_pcoa$data$Region <- factor(sc_pcoa$data$Region, levels = c("east", "west"))

bc_pcoa$data$TempGroup <- make_temp_group(bc_pcoa$data$temp)
sc_pcoa$data$TempGroup <- make_temp_group(sc_pcoa$data$temp)

bc_pcoa$data$TempGroup <- factor(
  bc_pcoa$data$TempGroup,
  levels = c("Low", "Medium", "High")
)
sc_pcoa$data$TempGroup <- factor(
  sc_pcoa$data$TempGroup,
  levels = c("Low", "Medium", "High")
)

temp_colors <- c(
  "Low" = "#4575B4",
  "Medium" = "orange",
  "High" = "#D73027"
)

# -----------------------------
# Bray-Curtis PCoA plots
# -----------------------------

bc_habitat <- plot_pcoa_group(
  bc_pcoa,
  group_var = "habitat",
  title = NULL,
  colors = habitat_colors,
  legend_title = "Habitat"
)

bc_temp <- plot_pcoa_group(
  bc_pcoa,
  group_var = "TempGroup",
  title = NULL,
  colors = temp_colors,
  legend_title = "Mean Annual Temperature"
)

bc_region <- plot_pcoa_group(
  bc_pcoa,
  group_var = "Region",
  title = NULL,
  colors = east_west_colors,
  legend_title = "Region"
)

# -----------------------------
# Sorensen PCoA plots
# -----------------------------

sc_habitat <- plot_pcoa_group(
  sc_pcoa,
  group_var = "habitat",
  title = NULL,
  colors = habitat_colors,
  legend_title = "Habitat"
)

sc_temp <- plot_pcoa_group(
  sc_pcoa,
  group_var = "TempGroup",
  title = NULL,
  colors = temp_colors,
  legend_title = "Mean Annual Temperature"
)

sc_region <- plot_pcoa_group(
  sc_pcoa,
  group_var = "Region",
  title = NULL,
  colors = east_west_colors,
  legend_title = "Region"
)

# -----------------------------
# One combined figure
# -----------------------------

blank_header <- ggplot() + theme_void()

bc_header <- ggplot() +
  labs(title = "Bray-Curtis") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 13))

sc_header <- ggplot() +
  labs(title = "Sorensen") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 13))

row_label <- function(label) {
  ggplot() +
    annotate(
      "text",
      x = 0.5,
      y = 0.5,
      label = label,
      angle = 90,
      fontface = "bold",
      size = 4
    ) +
    xlim(0, 1) +
    ylim(0, 1) +
    theme_void()
}

habitat_label <- row_label("Habitat")
temp_label <- row_label("Temperature")
region_label <- row_label("Region")

pcoa_figure <- (
  blank_header + bc_header + sc_header +
    habitat_label + (bc_habitat + labs(tag = "A")) + (sc_habitat + labs(tag = "B")) +
    temp_label + (bc_temp + labs(tag = "C")) + (sc_temp + labs(tag = "D")) +
    region_label + (bc_region + labs(tag = "E")) + (sc_region + labs(tag = "F"))
) +
  patchwork::plot_layout(
    ncol = 3,
    widths = c(0.08, 1, 1),
    heights = c(0.08, 1, 1, 1),
    guides = "collect"
  ) &
  theme(
    legend.position = "right",
    plot.tag = element_text(face = "bold", size = 14)
  )

pcoa_figure

# Optional save:
 ggsave(
   "pcoa_bc_sc_figure.pdf",
   pcoa_figure,
   width = 11,
   height = 12,
   dpi = 500
)
```

#### Plot the tanglegram with the expanded tree --------------------------------
note: this phylosymbosis
```{r}
# Convert hclust to phylo
hc <- hclust(mpse_sor_dist, method = "average")
tree <- as.phylo(hc)

plot(tree,
     tip.color = tip_colors,
     cex = 0.6,
     no.margin = TRUE,
     use.edge.length = FALSE)


assoc <- as.matrix(data.frame(
  V1 = as.character(gps$labtracking),   # labels for tree_exp (tree1)
  V2 = as.character(gps$labtracking),   # labels for tree   (tree2)
  stringsAsFactors = FALSE
))

# 3) build link color vector from your state/colors (must be same length as nrow(assoc))
gps$link_cols <- east_west_colors[ match(gps$Region, names(east_west_colors)) ]
link_cols <- as.character(gps$link_cols[ match(assoc[,1], gps$labtracking) ])

tree2 <- tree
tree2$edge.length <- NULL

### reate cophylo - object for tanglegram --------------------------------------
cophylo_obj <- cophylo(tree2, phy_tree,assoc)

plot(cophylo_obj,
     type = "phylogram",
     link.type = "curved",
     link.lty = "solid",
     link.lwd = 1.5,
     link.col = link_cols,
     ftype = "off",
     #cex = 0.6,
     mar = c(0.1, 0.1, 0.1, 0.1))

# assuming state_colors is defined as you pasted
df_legend <- data.frame(Region = names(east_west_colors), color = unname(east_west_colors))
df_legend$Region <- factor(df_legend$Region, levels = names(east_west_colors))

ggplot(df_legend, aes(x = 1, y = Region, fill = Region)) +
  geom_tile(color = "white", height = 0.9, width = 0.6) +
  scale_fill_manual(values = east_west_colors) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  theme_minimal(base_size = 10) +
  theme(axis.title = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none") +
  labs(title = "Region")
```

purty map with bac tree distance 
```{r}
library(mapdata)

my_colors <- c(
  "#5ab4ac",
  "#d8b365")
# Prepare map data
smoll_map <- gps[, c("decimalLatitude", "decimalLongitude","link_cols", "labtracking")]
smoll_map$decimalLatitude <- as.numeric(smoll_map$decimalLatitude)
smoll_map$decimalLongitude <- as.numeric(smoll_map$decimalLongitude)
smoll_map$link_cols <- as.factor(gps$link_cols)

# Create a color palette for tips
palette <- colorRampPalette(smoll_map$link_cols)(length(levels(smoll_map$link_cols)))

tip_colors <- palette[as.numeric(smoll_map$link_cols[match(tree$tip.label, gps$labtracking)])]

# Set edge lengths to 1 for plotting
tree_clado <- tree
tree_clado$edge.length <- rep(1, length(tree$edge.length))

# Convert phylo tree to map object
phymap <- phylo.to.map(
  tree_clado,
  smoll_map,
  plot = FALSE,
  database = "worldHires",
  regions = "USA"
)

# Plot the map
plot(phymap,
     ftype = "off",
     fsize = 0.8,
     lty = "solid",
     colors=palette,
     map.bg = "grey",
     #colors = "#8c510a",
     xlim = c(-130, -65),
     ylim = c(24, 50))
```

codiversity attempt 
```{r}
#---------------------------
# Taxonomic aggregation
#---------------------------
ps_Family <- tax_glom(ps, taxrank = "Genus", NArm = TRUE)

#---------------------------
# Clean taxa & samples
#---------------------------
ps_clean <- ps_Family %>%
  prune_taxa(taxa_sums(.) > 0, .) %>%
  prune_samples(sample_sums(.) > 0, .)

taxa_names(ps_clean) <- make.unique(taxa_names(ps_clean))


#---------------------------
# Prevalence filtering (≥5%)
#---------------------------
asv_mat <- as(otu_table(ps_clean), "matrix")
prev <- rowSums(asv_mat > 0)
keep <- prev >= 0.05 * ncol(asv_mat)
ps_clean <- prune_taxa(keep, ps_clean)

#---------------------------
# Split asvs by Phylum
#---------------------------
tax_df <- as(tax_table(ps_clean), "matrix") %>%
  as.data.frame() %>%
  rownames_to_column("asv") %>%
  select(asv, Phylum)

asv_by_phylum <- split(tax_df$asv, tax_df$Phylum)

#---------------------------
# Split asvs by Family
#---------------------------
#tax_df <- as(tax_table(ps_clean), "matrix") %>%
#  as.data.frame() %>%
#  rownames_to_column("asv") %>%
#  select(asv, Family)

#asv_by_Family <- split(tax_df$asv, tax_df$Family)

#---------------------------
# Build Left Trees (per phylum)
#---------------------------
scale_left <- 1.5

left_tree_plot_df <- map_dfr(names(asv_by_phylum), function(ph) {
  asvs <- asv_by_phylum[[ph]]
  ps_sub <- prune_taxa(asvs, ps_clean)
  
  if (ntaxa(ps_sub) < 2) return(NULL) # skip empty / singleton trees
  
  p <- ggtree(phy_tree(ps_sub)) + theme_tree2()
  
  p$data %>%
    mutate(
      Phylum = ph,
      x = x * scale_left  # scale left tree
    )
})


# Only keep phyla with trees
valid_phyla <- unique(left_tree_plot_df$Phylum)

left_tree_plot_df <- left_tree_plot_df %>%
  filter(Phylum %in% valid_phyla)

left_tip_df <- left_tree_plot_df %>%
  filter(isTip) %>%
  transmute(
    asv = label,
    x_left = x,
    y_left = y,
    Phylum
  )


#---------------------------
# Build Left Trees (per Family)
#---------------------------
#scale_left <- 1.5

#left_tree_plot_df <- purrr::map_dfr(names(asv_by_Family), function(gen) {
#  asvs <- asv_by_Family[[gen]]
#  ps_sub <- prune_taxa(asvs, ps_clean)

#  if (ntaxa(ps_sub) < 2) return(NULL)

#  p <- ggtree(phy_tree(ps_sub)) + theme_tree2()

#  p$data %>%
#    mutate(
#      Family = gen,
#      x = x * scale_left
#    )
#})

#valid_genera <- unique(left_tree_plot_df$Family)

#left_tree_plot_df <- left_tree_plot_df %>%
#  filter(Family %in% valid_genera)

#left_tip_df <- left_tree_plot_df %>%
#  filter(isTip) %>%
#  transmute(
#    asv = label,
#    x_left = x,
#    y_left = y,
#    Family
#  )


#---------------------------
# Build Right Tree (all samples)
#---------------------------
scale_right <- 1.0 # right tree width

p_right <- ggtree(phy_tree) +
  scale_x_reverse() +
  theme_tree2()

right_df <- p_right$data %>%
  filter(isTip) %>%
  transmute(
    labtracking = label,
    x_right = x,
    y_right = y
  )

right_tree_df <- p_right$data
max_x <- max(right_tree_df$x, na.rm = TRUE)
# flip tips to face left
right_tree_df <- right_tree_df %>%
  mutate(
    x = max(x, na.rm = TRUE) - x # mirror
  )

# normalize and scale
right_tree_df <- right_tree_df %>%
  mutate(
    x = (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE)),
    x = x * scale_right
  )

right_tip_df <- right_tree_df %>%
  filter(isTip) %>%
  transmute(
    labtracking = label,
    x_base = x,
    y_base = y
  )

#---------------------------
# Shift right trees per phylum
#---------------------------
max_left_x_df <- left_tree_plot_df %>%
  group_by(Phylum) %>%
  summarize(max_left_x = max(x, na.rm = TRUE), .groups = "drop")

# replicate right tree for each valid phylum
right_tree_df <- right_tree_df %>%
  crossing(Phylum = valid_phyla) %>%
  left_join(max_left_x_df, by = "Phylum") %>%
  mutate(x = x + max_left_x + 1) # shift right tree after left tree

right_tip_df <- right_tip_df %>%
  crossing(Phylum = valid_phyla) %>%
  left_join(max_left_x_df, by = "Phylum") %>%
  mutate(x_right = x_base + max_left_x + 1,
         y_right = y_base) %>%
  select(labtracking, x_right, y_right, Phylum)

#---------------------------
# Shift right trees per Family
#---------------------------
#max_left_x_df <- left_tree_plot_df %>%
#  group_by(Family) %>%
#  summarize(max_left_x = max(x, na.rm = TRUE), .groups = "drop")

#right_tree_df <- right_tree_df %>%
#  crossing(Family = valid_genera) %>%
#  left_join(max_left_x_df, by = "Family") %>%
#  mutate(x = x + max_left_x + 1)

#right_tip_df <- right_tip_df %>%
#  crossing(Family = valid_genera) %>%
#  left_join(max_left_x_df, by = "Family") %>%
#  mutate(
#    x_right = x_base + max_left_x + 1,
#    y_right = y_base
#  ) %>%
#  select(labtracking, x_right, y_right, Family)

#---------------------------
# Build links
#---------------------------
asv_mat <- as(otu_table(ps_clean), "matrix")

links <- asv_mat %>%
  as.data.frame() %>%
  rownames_to_column("asv") %>%
  pivot_longer(
    -asv,
    names_to = "labtracking",
    values_to = "Abundance"
  ) %>%
  filter(Abundance > 0) %>%
  left_join(tax_df, by = "asv") %>%
  #filter(Phylum %in% valid_genera) %>% # only valid phyla
  left_join(as(sample_data(ps_clean), "data.frame"), by = "labtracking") %>%
  mutate(
    w = scales::rescale(log10(Abundance + 1), c(0.2, 1))
  )

links_plot <- links %>%
  left_join(left_tip_df,  by = c("asv", "Phylum")) %>%
  left_join(right_tip_df, by = c("labtracking", "Phylum")) %>%
  filter(
    !is.na(x_left) & !is.na(y_left) &
    !is.na(x_right) & !is.na(y_right)
  ) %>%
  mutate(
    x_right = ifelse(x_left == x_right, x_right + 1e-6, x_right),
    y_right = ifelse(y_left == y_right, y_right + 1e-6, y_right)
  )



#---------------------------
# Plot
#---------------------------
ggplot() +
  # LEFT trees
  geom_tree(
    data = left_tree_plot_df,
    aes(x = x, y = y),
    inherit.aes = FALSE
  ) +
  # RIGHT trees
  geom_tree(
    data = right_tree_df,
    aes(x = x, y = y),
    inherit.aes = FALSE
  ) +
  # LINKS
  geom_curve(
    data = links_plot,
    aes(
      x = x_left,
      y = y_left,
      xend = x_right,
      yend = y_right,
      color = Region,
      linewidth = w
    ),
    curvature = 0.1,
    alpha = 0.3
  ) +
  scale_color_brewer(palette = "Set2", name = "Region") +
  scale_linewidth(range = c(0.2, 1), guide = "none") +
  facet_wrap(~Phylum, scales = "free_y") +
  theme_void() +
  coord_cartesian(clip = "off")

```

top 20 PCoA with individuals as points
```{r}
# Step 1: Get all Families 
families <- unique(as.vector(tax_table(ps_top20)[, "Genus"]))
families <- families[!is.na(families)]


# Step 2: Prepare containers
pcoa_list   <- list()
axis_labels <- list()

for (fam in families) {

  # Subset by Family
  ps_sub <- subset_taxa(ps, Genus == fam)

  # Remove empty samples
  ps_sub <- prune_samples(sample_sums(ps_sub) > 0, ps_sub)

  # Distance matrix
  dist_mat <- tryCatch(
    phyloseq::distance(ps_sub, method = "bray"),
    error = function(e) NULL
  )
  if (is.null(dist_mat)) next

  # PCoA
  pcoa_res <- tryCatch(
    ape::pcoa(dist_mat),
    error = function(e) NULL
  )
  if (is.null(pcoa_res)) next

  # Variance explained (guard against NA)
  var_exp <- round(
    100 * pcoa_res$values$Relative_eig[1:2],
    1
  )
  if (any(is.na(var_exp))) next

  # Sample metadata
  sample_df <- data.frame(sample_data(ps_sub))

  # PCoA dataframe
  pcoa_df <- data.frame(
    Sample = rownames(sample_df),
    PC1    = pcoa_res$vectors[, 1],
    PC2    = pcoa_res$vectors[, 2],
    land_use = sample_df$land_use,
    Region = sample_df$Region,
    Genus = fam
  )

  # Store facet title
  axis_labels[[fam]] <- data.frame(
    Genus = fam,
    title  = paste0(
      fam,
      "\nPC1 (", var_exp[1], "%) | PC2 (", var_exp[2], "%)"
    )
  )

  pcoa_list[[fam]] <- pcoa_df
}

# Combine results
combined_pcoa <- bind_rows(pcoa_list)
axis_labels   <- bind_rows(axis_labels)

combined_pcoa <- combined_pcoa %>%
  left_join(axis_labels, by = "Genus") %>%
  mutate(title = factor(title, levels = unique(title)))

# Step 3: Plot
ggplot(
  combined_pcoa,
  aes(PC1, PC2, color = Region)
) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(
    aes(group = Region),
    level = 0.95,
    linetype = 2,
    na.rm = TRUE
  ) +
  facet_wrap(~ title, scales = "free") +
  scale_color_manual(values = east_west_colors) +
  theme_bw()

```

with abundance 
```{r}
# Step 1: Get all Families 
families <- unique(as.vector(tax_table(ps_top20)[, "Species"]))
families <- families[!is.na(families)]

# Step 2: Prepare containers
pcoa_list   <- list()
axis_labels <- list()

for (fam in families) {

  # Subset by Species
  ps_sub <- subset_taxa(ps, Species == fam)

  # Remove empty samples
  ps_sub <- prune_samples(sample_sums(ps_sub) > 0, ps_sub)

  # Skip if too few samples
 if (length(sample_names(ps_sub)) < 3) next

  # -------------------------
  # Get abundance per sample
  # -------------------------
  abundance <- sample_sums(ps_sub)

  # Distance matrix
  dist_mat <- tryCatch(
    phyloseq::distance(ps_sub, method = "bray"),
    error = function(e) NULL
  )
  if (is.null(dist_mat)) next

  # PCoA
  pcoa_res <- tryCatch(
    ape::pcoa(dist_mat),
    error = function(e) NULL
  )
  if (is.null(pcoa_res)) next

  # Variance explained
  var_exp <- round(
    100 * pcoa_res$values$Relative_eig[1:2],
    1
  )
  if (any(is.na(var_exp))) next

  # Sample metadata
  sample_df <- data.frame(sample_data(ps_sub))

  # PCoA dataframe (ADD abundance)
  pcoa_df <- data.frame(
    Sample    = rownames(sample_df),
    PC1       = pcoa_res$vectors[, 1],
    PC2       = pcoa_res$vectors[, 2],
    Region    = sample_df$Region,
    Abundance = abundance[rownames(sample_df)],
    Family    = fam
  )

  # Store facet title
  axis_labels[[fam]] <- data.frame(
    Family = fam,
    title  = paste0(
      fam,
      "\nPC1 (", var_exp[1], "%) | PC2 (", var_exp[2], "%)"
    )
  )

  pcoa_list[[fam]] <- pcoa_df
}

# Combine results
combined_pcoa <- bind_rows(pcoa_list)
axis_labels   <- bind_rows(axis_labels)

combined_pcoa <- combined_pcoa %>%
  left_join(axis_labels, by = "Family") %>%
  mutate(title = factor(title, levels = unique(title))) 

combined_pcoa <- combined_pcoa %>%
  mutate(Abundance_log = log10(Abundance + 1))

# Step 3: Plot
ggplot(
  combined_pcoa,
  aes(PC1, PC2,
      color = Region,
      shape = Region,
      size  = Abundance)
) +
  geom_point(alpha = 0.75) +
  stat_ellipse(
    aes(group = Region),
    level = 0.95,
    linetype = 2,
    linewidth = 0.6,
    na.rm = TRUE
  ) +
  facet_wrap(~ title, scales = "free") +
  scale_color_manual(values = east_west_colors) +
  
  # Cleaner size scaling
  scale_size_continuous(
    range = c(2, 7),
    breaks = pretty(combined_pcoa$Abundance, n = 4),
    name = "Abundance"
  ) +
  
  guides(
    size = guide_legend(
      override.aes = list(alpha = 1)
    )
  ) +
  
  theme_bw(base_size = 12) +
  theme(
    legend.position = "right",
    strip.text = element_text(face = "bold")
  )

```
