---
title: "microbiome by habitat"
author: "Dani Blumstein"
date: "2025-12-30"
output: html_document
---

mostly packages 
```{r}
#if (!requireNamespace("BiocManager", quietly=TRUE))
#    install.packages("BiocManager")
## BiocManager::install("BiocUpgrade") ## you may need this
#BiocManager::install("MicrobiotaProcess",force = TRUE)
#BiocManager::install("phyloseq",force = TRUE)
#BiocManager::install("ggtree", force = TRUE)
#BiocManager::install("tidytree", force = TRUE, type = "source")

#install.packages("remotes")
#remotes::install_github("YuLab-SMU/ggtree")
#remotes::install_github("YuLab-SMU/MicrobiotaProcess")



#install.packages(
#  "microViz",
#  repos = c(davidbarnett = "https://david-barnett.r-universe.dev", getOption("repos"))
#)

#if (!requireNamespace("devtools", quietly = TRUE)){install.packages("devtools")}
#devtools::install_github("jbisanz/qiime2R")
#devtools::install_github("zpneal/backbone", ref = "devel", build_vignettes = TRUE)

setwd("~/Library/CloudStorage/OneDrive-ArizonaStateUniversity/pdoc_ASU/projects/NEON/Neon_Rproject")

suppressPackageStartupMessages({
    library(MicrobiotaProcess) 
    library(brms)
    library(phyloseq)
    library(ggplot2)
    library(tidyverse) 
    library(vegan) 
    library(coin) 
    library(reshape2) 
    library(ggnewscale)
    library(qiime2R)
    library(readxl)
    library(dplyr)
    library(microViz)
    library(UpSetR)
    library(patchwork)
    library(ggtree)
    library(tidytree)
    library(phylobase)
    library(ggtreeExtra)
    library(sf)
    library(phytools)
    library(SummarizedExperiment)
    library(remotes)
    library(terra)
    library(ggspatial)
    library(stringr)
    library(purrr)
    library(gridExtra)
    library(geosphere)
    library(ape)
    library(phangorn)
})
```

colors
```{r}
habitat_colors <- c(
  # East of Mississippi (cool / forested)
  "Deciduous Forest"     = "#0072B2",  # true blue
  "Mixed Forest"         = "#2CA25F",  # teal-green
  
  # West of Mississippi (warm / dry / managed)
  "Cultivated Crops"     = "#E69F00",  # orange (CVD-safe, high salience)
  "Evergreen Forest"     = "#A0522D",  # 
  "Grassland/Herbaceous" = "#F0E442",  # yellow (excellent contrast on maps)
  "Shrub/Scrub"          = "#D55E00"   # vermillion (distinct from crops)
)
  
east_west_colors <- c(
  "east" = "#5ab4ac",  # muted steel blue
  "west" = "#d8b365"   # muted warm tan
)

```

MicrobiotaProcess import
```{r}
mpse_samples <- mp_import_qiime2(otuqza="data/microbiome/rarefied-table.qza", 
                          taxaqza="data/microbiome/taxonomy.qza", 
                          mapfilename="data/microbiome/metadata_edit.tsv")
mpse_samples

#Set labtracking column as the sample names
new_names <- colData(mpse_samples)$labtracking

# Make sure they are unique
stopifnot(!anyDuplicated(new_names))

# Assign as sample names
colnames(mpse_samples) <- new_names
rownames(colData(mpse_samples)) <- new_names

#filter to only pema and drop AZ
mpse_samples_filtered <- mpse_samples %>%
  filter(specificEpithet == "maniculatus") %>%
  filter(stateProvince != "Arizona")

mpse_asv_table <- assay(mpse_samples_filtered, "Abundance")

mpse_bc_dist   <- vegdist(t(mpse_asv_table), method = "bray") 
mpse_sor_dist  <- vegdist(t(mpse_asv_table), method = "bray", binary = TRUE) 

labs <- attr(mpse_bc_dist, "Labels")
```

phyloseq import 
```{r}
#---------------------------
# Import data
#---------------------------
physeq <- qza_to_phyloseq(
  features  = "data/microbiome/rarefied-table.qza",
  tree      = "data/microbiome/rooted-tree.qza",
  taxonomy  = "data/microbiome/taxonomy.qza",
  metadata  = "MicrobiotaProcess_sample_data.tsv"
)

#---------------------------
# Sample filtering
#---------------------------
ps <- physeq %>%
  ps_filter(scientificName == "Peromyscus maniculatus") %>%
  ps_filter(stateProvince != "Arizona")

sample_names(ps) <- sample_data(ps)$labtracking

ps_otu <- otu_table(ps)
sort(rowSums(t(ps_otu)))

ps_bc_dist   <- vegdist(t(ps_otu), method = "bray") 
ps_sor_dist  <- vegdist(t(ps_otu), method = "bray", binary = TRUE)

meta <- data.frame(sample_data(ps))
meta$decimalLatitude <- as.numeric(meta$decimalLatitude)
meta$decimalLongitude <- as.numeric(meta$decimalLongitude)
```

geographic distance
```{r}
# -----------------------------
# Geographic distance
# -----------------------------
#distance from a centroid
coords <- as.matrix(meta[, c("decimalLongitude", "decimalLatitude")])
centroid <- colMeans(coords, na.rm = TRUE)
meta$geo_km <- distVincentyEllipsoid(coords, centroid) / 1000

#or 

#distance in a matrix
coords2 <- meta[, c("decimalLongitude", "decimalLatitude")]
geo_dist <- distm(coords2, fun = distHaversine) / 1000 
rownames(geo_dist) <- rownames(meta) 
colnames(geo_dist) <- rownames(meta)
```

ANCOMBC
```{r}
if (!require("ANCOMBC", quietly = TRUE)){
       BiocManager::install("ANCOMBC")}

if (!require("microbiome", quietly = TRUE)){
       BiocManager::install("microbiome")}

library(ANCOMBC)
library(microbiome)

# Remove uncultured / unclassified / unidentified taxa
ps2 <- subset_taxa(
  ps,
  !grepl(
    "uncultured|unclassified|unidentified|unknown",
    Species,
    ignore.case = TRUE
  )
)

out = ancombc(data = ps2, formula = "Region", p_adj_method = "fdr")

results_ancom_bc = data.frame(ASV = out$res$lfc$taxon,
                              lcf = out$res$lfc$Regionwest, 
                              W = out$res$W$Regionwest, 
                              p_val = out$res$p_val$Regionwest, 
                              q_value = out$res$q_val$Regionwest, 
                              Diff_ab =  out$res$diff_abn$Regionwest)
results_ancom_bc$lcf = results_ancom_bc$lcf * -1

results_ancom_bc$association = ifelse(results_ancom_bc$p_val < 0.05 & results_ancom_bc$lcf > 0, "east", "Not singificant")
results_ancom_bc$association = ifelse(results_ancom_bc$p_val < 0.05 & results_ancom_bc$lcf < 0, "west",results_ancom_bc$association)

ggplot(results_ancom_bc, aes(x = as.numeric(lcf), y = -log10(as.numeric(p_val)), color = association)) +
  geom_point() +
  labs(x = "Log2 Fold Change", y = "-log10(p-value)",
       title = "Volcano Plot") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  theme_bw()


TAXtab =as.data.frame(tax_table(ps2))

name_tmp = sprintf("ASV%s",seq(1:nrow(TAXtab))) 
#now let's add some information to this. For example, we want to add the genus, and if not available, the last taxonomic information available.
TAXtab$name =  TAXtab$Species
TAXtab$name = ifelse(is.na(TAXtab$name), TAXtab$Genus, TAXtab$name)
TAXtab$name = ifelse(is.na(TAXtab$name), TAXtab$Family, TAXtab$name)
TAXtab$name = ifelse(is.na(TAXtab$name), TAXtab$Order, TAXtab$name)
TAXtab$name = ifelse(is.na(TAXtab$name), TAXtab$Class, TAXtab$name)
TAXtab$name = ifelse(is.na(TAXtab$name), TAXtab$Phylum, TAXtab$name)
unique(is.na(TAXtab$name)) # Ok, we have all names

TAXtab$name = paste(name_tmp, TAXtab$name, sep = "-")
TAXtab$name[1:10] # ok it looks good.

results_ancom_bc$name = TAXtab$name[match(results_ancom_bc$ASV ,rownames(TAXtab))]

results_sig = results_ancom_bc[results_ancom_bc$p_val < 0.05,]
table(results_sig$association)

Sum_abundance= colSums(decostand(t(ps_otu), method = "total"))
results_sig$sum_abundance = Sum_abundance[match(results_sig$ASV, names(Sum_abundance))]
results_sig = results_sig[order(results_sig$sum_abundance, decreasing = TRUE),]
results_sig = results_sig[1:14,]
results_sig$name = factor(results_sig$name, levels = rev(results_sig$name))

a = ggplot(results_sig, aes(lcf, name, color = association ))+
  geom_point() +theme_bw() +
  geom_segment( aes(x=0, xend=lcf, y=name, yend=name, color = association)) +
  geom_vline(xintercept = 0, size=0.3) +xlab("log2FoldChange") +ylab(NULL) +
  theme_bw() +
  scale_color_manual(values = c("east" = "#5ab4ac","west" = "#d8b365"))
  
a

ASVtab_rel = decostand(t(ps_otu), method = "total")
AASVtab_sig = data.frame(ASVtab_rel[,as.numeric(rownames(results_sig))])
colnames(AASVtab_sig) = results_sig$name
AASVtab_sig$Region = meta$Region
melted = reshape2::melt(AASVtab_sig, values = "Region")
melted$variable = factor(melted$variable, levels = rev(results_sig$name))

b = ggplot(melted, aes(value, variable, fill = Region)) +
  geom_boxplot(outlier.size = 0.5) + theme_bw() + xlab("Relative abundance") +
    theme( axis.text.y=element_blank()) + ylab(NULL)

b 
```

dist decay by gen
```{r}
ps2 <- subset_taxa(
  ps,
  !grepl(
    "uncultured|unclassified|unidentified|unknown",
    Species,
    ignore.case = TRUE
  )
)
TAXtab <- as.data.frame(tax_table(ps2))

name_tmp <- sprintf("ASV%s", seq_len(nrow(TAXtab)))

# choose lowest available taxonomy
TAXtab$name <- TAXtab$Species
TAXtab$name <- ifelse(is.na(TAXtab$name), TAXtab$Genus, TAXtab$name)
TAXtab$name <- ifelse(is.na(TAXtab$name), TAXtab$Family, TAXtab$name)
TAXtab$name <- ifelse(is.na(TAXtab$name), TAXtab$Order, TAXtab$name)
TAXtab$name <- ifelse(is.na(TAXtab$name), TAXtab$Class, TAXtab$name)
TAXtab$name <- ifelse(is.na(TAXtab$name), TAXtab$Phylum, TAXtab$name)

unique(is.na(TAXtab$name))  # should be FALSE

# add ASV id
TAXtab$name <- paste(name_tmp, TAXtab$name, sep = "-")

taxa_names(ps2) <- TAXtab$name

ps_genus <- tax_glom(ps, taxrank = "Genus")
ps3 <- filter_taxa(ps3, function(x) sum(x > 0) >= 10, TRUE)
#ps4 <- filter_taxa(ps3, function(x) sum(x) > 500, TRUE)

otu <- as(otu_table(ps3), "matrix")
tax <- as.data.frame(tax_table(ps3))

genus_results <- data.frame()

for(i in 1:nrow(otu)){

  #g <- rownames(tax)[i]
  g <- tax$Genus[i]

  counts <- otu[i, , drop = FALSE]

  if(sum(counts) == 0) next

  bc <- vegdist(t(counts), method = "bray")#, binary = TRUE)

  bc_df <- melt(as.matrix(bc),
                varnames = c("Sample1","Sample2"),
                value.name = "BC_Dissimilarity")

  geo_df <- melt(as.matrix(geo_dist),
                 varnames = c("Sample1","Sample2"),
                 value.name = "geo_Distance")

  df <- geo_df
  df$BC_Similarity <- 1 - bc_df$BC_Dissimilarity

  df <- df %>%
    filter(Sample1 != Sample2) %>%
    mutate(pair_min = pmin(as.character(Sample1), as.character(Sample2)),
           pair_max = pmax(as.character(Sample1), as.character(Sample2))) %>%
    distinct(pair_min, pair_max, .keep_all = TRUE)

  model <- lm(BC_Similarity ~ geo_Distance, data = df)

  genus_results <- rbind(
    genus_results,
    data.frame(
      genus = g,
      slope = coef(model)[2],
      r2 = summary(model)$r.squared
    )
  )
}

head(genus_results)

top_genera <- genus_results %>%
  arrange(slope) %>%
  dplyr::slice(1:20)

top_names <- top_genera$genus

plot_df <- data.frame()

for(g in top_names){

  #ps_g <- prune_taxa(taxa_names(ps_genus) == g, ps_genus)
  ps_g <- subset_taxa(ps3, Genus == g)
  otu_g <- as(otu_table(ps_g), "matrix")

  bc <- vegdist(t(otu_g), method = "bray")#, binary = TRUE)

  melted_BC <- melt(as.matrix(bc),
                    varnames = c("Sample1","Sample2"),
                    value.name = "BC_Dissimilarity")

  melted_geo <- melt(as.matrix(geo_dist),
                     varnames = c("Sample1","Sample2"),
                     value.name = "geo_Distance")

  melted_diffs <- melted_geo
  melted_diffs$BC_Similarity <- 1 - melted_BC$BC_Dissimilarity

  #### remove half of the dataset
  melted_diffs <- melted_diffs %>%
    filter(Sample1 != Sample2) %>%
    mutate(pair_min = pmin(as.character(Sample1), as.character(Sample2)),
           pair_max = pmax(as.character(Sample1), as.character(Sample2))) %>%
    distinct(pair_min, pair_max, .keep_all = TRUE) %>%
    dplyr::select(Sample1, Sample2, geo_Distance, BC_Similarity)

  # Create sample -> region map
  region_map <- setNames(meta$Region, meta$labtracking)

  # Assign regions
  melted_diffs$region1 <- region_map[melted_diffs$Sample1]
  melted_diffs$region2 <- region_map[melted_diffs$Sample2]

  # Same region?
  melted_diffs$same_region <- melted_diffs$region1 == melted_diffs$region2

  # Scale geographic distance
  range.use <- function(x,min.use,max.use){
    (x - min(x,na.rm=T)) /
    (max(x,na.rm=T)-min(x,na.rm=T)) *
    (max.use - min.use) + min.use
  }

  melted_diffs$dist_scaled <- range.use(melted_diffs$geo_Distance, 0, 1)

  # Optional summary
  print(summary(melted_diffs$geo_Distance,
                by = melted_diffs$same_region))

  # Add genus label
  melted_diffs$genus <- g

  plot_df <- rbind(plot_df, melted_diffs)
}

line_colors <- c("TRUE" = "#762a83",
                 "FALSE" = "#b2182b")

base_colors <- colorspace::lighten(line_colors, 0.3)

ggplot(plot_df,
       aes(x = geo_Distance,
           y = BC_Similarity)) +

  geom_point(aes(color = same_region),
             size = 1.5,
             alpha = 0.5) +

  scale_color_manual(values = base_colors, guide = "none") +

  ggnewscale::new_scale_color() +

  geom_smooth(aes(color = same_region),
              method = "lm",
              se = FALSE,
              linewidth = 1) +

  scale_color_manual(values = line_colors) +

  facet_wrap(~genus, scales = "free_y") +

  theme_bw() +

  labs(
    x = "Distance (km)",
    y = "BC Similarity",
    color = "Same Region"
  )

ggplot(plot_df,
       aes(x = BC_Similarity, fill = same_region)) +

  geom_bar(
    bins = 30,
    alpha = 0.6,
    #position = "identity"
  ) +
  scale_fill_manual(values = line_colors) +

  facet_wrap(~genus, scales = "free_y") +

  theme_bw() +

  labs(
    x = "Sor Similarity",
    y = "Count",
    fill = "Same region"
  )
```

```{r}
# Extract abundance matrix
otu <- as(otu_table(ps2), "matrix")

# Ensure taxa are rows
if(!taxa_are_rows(ps2)) {
  otu <- t(otu)
}

# Add the new taxa names from TAXtab
rownames(otu) <- TAXtab$name

meta <- data.frame(sample_data(ps2))
region <- meta$Region
names(region) <- rownames(meta)

rownames(annotation_col) <- rownames(meta)

# Match sample order
annotation_col <- annotation_col[colnames(otu), , drop = FALSE]

top_taxa <- names(sort(rowSums(otu), decreasing = TRUE))[1:15]
otu_top <- otu[top_taxa, ]

otu_rel <- sweep(otu_top, 2, colSums(otu_top), "/")
otu_log <- log10(otu_rel + 1e-6)

otu_region <- aggregate(t(otu_rel), by = list(meta$Region), FUN = mean)
rownames(otu_region) <- otu_region$Group.1
otu_region <- t(otu_region[, -1])

ha <- HeatmapAnnotation(
  Region = region[colnames(otu_top)],
  col = list(
    Region = c(
  "east" = "#5ab4ac",  # muted steel blue
  "west" = "#d8b365"   # muted warm tan
    )
  )
)
Heatmap(
  otu_top,
  name = "log10 abundance",
  top_annotation = ha,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = FALSE,
  row_names_gp = grid::gpar(fontsize = 7),
  column_split = region[colnames(otu_top)]
)

# Extract abundance matrix
otu <- as(otu_table(ps2), "matrix")

# Ensure taxa are rows
if(!taxa_are_rows(ps2)) {
  otu <- t(otu)
}

# Get taxonomy table
tax <- as.data.frame(tax_table(ps2))

# Add genus names (replace NA with "Unknown")
tax$Genus[is.na(tax$Genus)] <- "Unknown"

# Collapse ASVs to genus
otu_genus <- rowsum(otu, group = tax$Genus)

meta <- data.frame(sample_data(ps2))
region <- meta$Region
names(region) <- rownames(meta)

rownames(annotation_col) <- rownames(meta)

# Match sample order
annotation_col <- annotation_col[colnames(otu_genus), , drop = FALSE]

# Select top 15 genera
top_taxa <- names(sort(rowSums(otu_genus), decreasing = TRUE))[1:15]
otu_top <- otu_genus[top_taxa, ]

# Convert to relative abundance
otu_rel <- sweep(otu_top, 2, colSums(otu_top), "/")

# Log transform
otu_log <- log10(otu_rel + 1e-6)

# Average by region
otu_region <- aggregate(t(otu_rel), by = list(meta$Region), FUN = mean)
rownames(otu_region) <- otu_region$Group.1
otu_region <- t(otu_region[, -1])


# Plot heatmap
Heatmap(
  otu_log,
  name = "log10 abundance",
  top_annotation = ha,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = FALSE,
  row_names_gp = grid::gpar(fontsize = 8),
  column_split = region[colnames(otu_top)]
)
```

phyloseq import again. did a ton of maniputaion and want to start fresh to be sure
```{r}
#---------------------------
# Import data
#---------------------------
physeq <- qza_to_phyloseq(
  features  = "data/microbiome/rarefied-table.qza",
  tree      = "data/microbiome/rooted-tree.qza",
  taxonomy  = "data/microbiome/taxonomy.qza",
  metadata  = "MicrobiotaProcess_sample_data.tsv"
)

#---------------------------
# Sample filtering
#---------------------------
ps <- physeq %>%
  ps_filter(scientificName == "Peromyscus maniculatus") %>%
  ps_filter(stateProvince != "Arizona")

sample_names(ps) <- sample_data(ps)$labtracking

ps_otu <- otu_table(ps)
sort(rowSums(t(ps_otu)))

ps_bc_dist   <- vegdist(t(ps_otu), method = "bray") 
ps_sor_dist  <- vegdist(t(ps_otu), method = "bray", binary = TRUE)

meta <- data.frame(sample_data(ps))
meta$decimalLatitude <- as.numeric(meta$decimalLatitude)
meta$decimalLongitude <- as.numeric(meta$decimalLongitude)
```

general stats
```{r}
total_asvs <- ntaxa(ps)
cat("Total number of ASVs:", total_asvs, "\n\n")

# Agglomerate at Phylum level
ps_phylum <- tax_glom(ps, taxrank = "Phylum")

# Transform to relative abundance
ps_phylum_rel <- transform_sample_counts(ps_phylum, 
                                         function(x) x / sum(x) * 100)

# Melt to dataframe
phylum_df <- psmelt(ps_phylum_rel)

# Calculate mean % abundance per region per phylum
phylum_summary <- phylum_df %>%
  group_by(Phylum) %>%
  summarise(mean_percent = mean(Abundance), .groups = "drop")

print(phylum_summary)

```

% shared ASV
```{r}
#easiset with binary data
ps.pa <- transform_sample_counts(ps, function(x) ifelse(x > 0, 1, 0))
otu <- as(otu_table(ps.pa), "matrix")

if (taxa_are_rows(ps.pa)) {
  otu <- t(otu)
}

bray <- vegdist(otu, method = "bray", binary = TRUE)
bray_sim <- 1 - as.matrix(bray)

# 3. Long format + upper triangle only
shared_df <- as.data.frame(as.table(bray_sim)) %>%
  dplyr::filter(as.integer(Var1) < as.integer(Var2)) %>%   # safest filter
  dplyr::mutate(
    Var1 = as.character(Var1),
    Var2 = as.character(Var2),
    PercentShared = Freq * 100
  ) %>%
  dplyr::select(-Freq)

# 4. Join metadata
shared_df <- shared_df %>%
  dplyr::left_join(meta, by = c("Var1" = "labtracking")) %>%
  dplyr::rename(State1 = stateProvince, Region1 = Region) %>%
  dplyr::left_join(meta, by = c("Var2" = "labtracking")) %>%
  dplyr::rename(State2 = stateProvince, Region2 = Region)

# 5. Define comparison categories
shared_df <- shared_df %>%
  dplyr::mutate(
    Comparison = dplyr::case_when(
      State1 == State2 ~ "Within state",
      State1 != State2 & Region1 == Region2 ~ "Between states (same region)",
      Region1 != Region2 ~ "Between regions (East vs West)"
    )
  )

shared_df$Comparison <- factor(
  shared_df$Comparison,
  levels = c(
    "Within state",
    "Between states (same region)",
    "Between regions (East vs West)"
  )
)

summary <- shared_df %>%
  group_by(Comparison) %>%
  summarise(mean_percent = mean(PercentShared), .groups = "drop")

print(summary)

# 6. Plot
ggplot(shared_df, aes(x = Comparison, y = PercentShared, fill = Comparison)) +
  geom_violin(trim = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
  theme_bw() +
  labs(
    x = NULL,
    y = "Percent shared ASVs"
  ) +
  scale_fill_manual(values = c(
    "#bababa",
    "#9C68AE",
    "#D85F64"
  )) +
  guides(fill = "none") +
  theme(
    axis.title.y = element_text(size = 16),
    axis.text.x  = element_text(size = 14),
    axis.text.y  = element_text(size = 14)
  )
```

rasters and such
```{r}
#use the gps points and USGS national land coverage data to assign habitat
point <- as.data.frame(colData(mpse_samples_filtered))
point$decimalLatitude <- as.numeric(point$decimalLatitude)
point$decimalLongitude <- as.numeric(point$decimalLongitude)

p1 <- vect(point, geom = c("decimalLongitude", "decimalLatitude"), crs = "+proj=longlat +datum=WGS84")

#Convert to sf
points_sf <- st_as_sf(p1)  

#raster data from https://www.mrlc.gov/data
raster_obj <- rast("Annual_NLCD_LndCov_2020_CU_C1V1/Annual_NLCD_LndCov_2020_CU_C1V1.tif")

#raster data from https://prism.oregonstate.edu/data/
rain_raster <- rast("prism_ppt_us_30s_2020/prism_ppt_us_30s_2020.tif")
mean_temp_raster <- rast("prism_tmean_us_30s_2020/prism_tmean_us_30s_2020.tif")

plot(raster_obj)
plot(rain_raster)
plot(mean_temp_raster)

pts_sf <- st_transform(points_sf, crs(raster_obj))
pst_sf_rain <- st_transform(points_sf, crs(rain_raster))
pts_sf_temp <- st_transform(points_sf, crs(mean_temp_raster))

# Convert sf → terra vector
pts_vect <- vect(pts_sf)
pts_vect_rain <- vect(pst_sf_rain)
pts_vect_temp <- vect(pts_sf_temp)

# Extract raster values
vals <- extract(raster_obj, pts_vect)
vals_rain <- extract(rain_raster, pts_vect_rain)
vals_temp <- extract(mean_temp_raster, pts_vect_temp)

# Combine with original data
results <- pts_sf |>
  bind_cols(
    land_use = vals[,2]
  )

results$land_use_location <- paste(results$land_use, results$Mississippi_River_Label, sep = " ")
#write_tsv(results, file = "metadata_edit2.tsv")

colData(mpse_samples_filtered)$land_use_location <- results$land_use_location
colData(mpse_samples_filtered)$land_use <- results$land_use

results <- pst_sf_rain |>
  bind_cols(
    prism_ppt_us_30s_2020 = vals_rain[,2]
  )
colData(mpse_samples_filtered)$rain <- results$prism_ppt_us_30s_2020

results <- pts_sf_temp |>
  bind_cols(
    prism_tmean_us_30s_2020 = vals_temp[,2]
  )
colData(mpse_samples_filtered)$temp <- results$prism_tmean_us_30s_2020

# Extract sample data from MPSE object
sample_data_df <- colData(mpse_samples_filtered)

# Export as tab-delimited text file
write.table(sample_data_df, 
            file = "MicrobiotaProcess_sample_data.txt", 
            sep = "\t", 
            quote = FALSE, 
            row.names = TRUE)

```

from gabri for alpha stats
```{r}
otu <- t(otu_table(ps))

#sort the metadata and the otu table
metadata_filtered_sorted <- meta[order(meta$labtracking),]
otu_sorted <- otu[order(rownames(otu)),]

#check the data for order 
unique(rownames(otu_sorted) == metadata_filtered_sorted$labtracking) # OK

### Calculate alpha diversity
metadata_filtered_sorted$Richness.rar<-specnumber(otu_sorted) # Counts otu number
metadata_filtered_sorted$Shannon.rar<-vegan::diversity(otu_sorted, index="shannon") ## calculate Shannon H

wilcox.test(Shannon.rar ~ Region, data = metadata_filtered_sorted)

min_richness <- min(metadata_filtered_sorted$Richness.rar)
max_richness <- max(metadata_filtered_sorted$Richness.rar)

cat("Richness range per sample (Observed ASVs):\n")
cat("Min:", min_richness, "\n")
cat("Max:", max_richness, "\n\n")

# Statistical test (Kruskal-Wallis recommended for richness)
kruskal_richness <- kruskal.test(Richness.rar ~ Region, data = metadata_filtered_sorted)
print(kruskal_richness)

summary <- metadata_filtered_sorted %>%
  group_by(Region) %>%
  summarise(mean_percent = mean(Richness.rar), .groups = "drop")

print(summary)

a = ggplot(metadata_filtered_sorted, aes(x = Region, y = Richness.rar, fill = Region)) +
  geom_boxplot() +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), alpha = 0.7) +
  theme_classic() +
  labs(y = "Number of ASVs", x = "Region") +
  scale_fill_manual(values = east_west_colors)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
a
```

Beta diversity 
```{r}
set.seed(12)
bac.dist<-vegdist(otu_sorted, method="bray") ### this calculates BC dissimilaries

set.seed(12)
bac.nmds<-metaMDS(bac.dist, k=2, try=999)
bac.nmds$stress #0.21
stress= paste("stress = ", round(bac.nmds$stress, digits = 2))

meta$Axis01<-bac.nmds$points[,1]
meta$Axis02<-bac.nmds$points[,2]

# install.packages("viridis") if needed
bac_nmds <- ggplot(meta, aes(x = Axis01, y = Axis02)) +
  geom_point(aes(color = Region, shape = Region), size = 2, alpha = 0.8) +
  stat_ellipse(aes(fill = Region), geom = "polygon", alpha = 0.45, level = 0.68, show.legend = FALSE) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  scale_fill_viridis_d(option  = "turbo", begin = 0.05, end = 0.95) +
  theme_classic() +
  labs(x = "NMDS1", y = "NMDS2", shape = "Region") +
  ggtitle("Bray Curtis dissimilarity")
bac_nmds
```

dbmem
#there is spatial autocorrelation 
```{r}
dbrda1<- dbrda(ps_bc_dist ~ decimalLatitude + decimalLongitude + Region, data = meta, distance = "bray")


anova1<-anova(dbrda1, by="margin", permutation = 9999)
anova1
```

Bayesian Generalized (Non-)Linear Multivariate Multilevel Models
```{r}
#### Create metadata -------------
#geo_dist
#ps_bc_dist
#ps_sor_dist

melted_BC <- reshape2::melt(as.matrix(ps_bc_dist), varnames = c("Sample1", "Sample2"), value.name = "BC_Dissimilarity")

melted_geo <- reshape2::melt(as.matrix(geo_dist), varnames = c("Sample1", "Sample2"), value.name = "geo_Distance")

unique(melted_BC$Sample1 == melted_geo$Sample1)
unique(melted_BC$Sample2 == melted_geo$Sample2) # Ok

#### Reshape
melted_diffs = melted_geo
melted_diffs$BC_Dissimilarity = melted_BC$BC_Dissimilarity

### convert into similarity
#melted_diffs$geo_Similarity = 1 - melted_diffs$geo_Dissimilarity
melted_diffs$BC_Similarity = 1 - melted_diffs$BC_Dissimilarity

#### remove half of the dataset
melted_diffs <- melted_diffs %>%
  filter(Sample1 != Sample2) %>%
  mutate(pair_min = pmin(as.character(Sample1), as.character(Sample2)),
         pair_max = pmax(as.character(Sample1), as.character(Sample2))) %>%
  distinct(pair_min, pair_max, .keep_all = TRUE) %>%
  dplyr::select(Sample1, Sample2, geo_Distance,BC_Similarity )

# Create a vector mapping sample -> region
region_map <- setNames(meta$Region, meta$labtracking)

# Use match (or the vector) to get the regions
melted_diffs$region1 <- region_map[melted_diffs$Sample1]
melted_diffs$region2 <- region_map[melted_diffs$Sample2]

# Check if they are in the same region
melted_diffs$same_region <- melted_diffs$region1 == melted_diffs$region2

range.use <- function(x,min.use,max.use){ (x - min(x,na.rm=T))/(max(x,na.rm=T)-min(x,na.rm=T)) * (max.use - min.use) + min.use }

melted_diffs$dist_scaled <- range.use(melted_diffs$geo_Distance, 0, 1)

summary(melted_diffs$geo_Distance, by = same_region)

#slope = change in similarity per 1 SD of 100 km distances

#With z-scores:
#mean ≈ 0 in East and West
#sd ≈ 1 in each region
#which is much closer to a biological decay rate.
melted_diffs <- melted_diffs %>%
  #group_by(same_region) %>%
  mutate(
    dist_100 = geo_Distance / 100,
    dist_z = as.numeric(scale(dist_100))
  ) %>%
  ungroup()

#install.packages("ggnewscale")  # once
library(ggnewscale)

#install.packages("colorspace")  # once
library(colorspace)
line_colors <- c("TRUE" = "#762a83",
                 "FALSE"  = "#b2182b")

#line_colors <- darken(base_colors, 0.3)
base_colors <- lighten(line_colors, 0.3 )
ggplot(melted_diffs, aes(x = geo_Distance, y = BC_Similarity)) +

  # Points (light colors)
  geom_point(aes(color = same_region), size = 2) +
  scale_color_manual(values = base_colors, guide = "none") +

  ggnewscale::new_scale_color() +

  # Lines (darker colors)
  geom_smooth(aes(color = same_region),
              method = "lm",
              se = FALSE,
              linewidth = 1.2) +
  scale_color_manual(values = line_colors) +

  theme_bw() +
    labs(
    x = "Distance (km)",
    y = "Bray– Curtis Similarity",
    color = "Same Regions"
  ) +

  theme(
    axis.title = element_text(size = 13),
    axis.text  = element_text(size = 11),
    legend.title = element_text(size = 12),
    legend.text  = element_text(size = 11)
  )

#library(brms)

#model <- brm(BC_Similarity~1+dist_scaled * same_region +
#            (1|mm(Sample1,Sample2)),
#              data = melted_diffs,
#              #family= "Beta",
#              warmup = 2000, iter = 6000,
#              threads = threading(6),
#              cores = 24,
#              chains = 4,
#              control=list(adapt_delta=0.95),
#              init=0)

#cor(melted_diffs$dist_scaled, as.numeric(melted_diffs$same_region))

#Scale distance within region
#Use || for random slopes
#Add priors + increase adapt_delta

# || removes slope–intercept correlation
#within-region scaling removes confounding
#priors regularize weakly identified slopes
#distance is still biologically meaningful

#model <- brm(
#  BC_Similarity ~ 1 + dist_z +
#    (1 + dist_z || same_region) +
#    (1 | mm(Sample1, Sample2)),
#  data = melted_diffs,
  
#  prior = c(
#    prior(normal(0, 1), class = "b"),
#    prior(normal(0, 1), class = "sd")
#  ),

#  warmup = 2000, iter = 6000,
#  chains = 4,
#  cores = 24,
#  threads = threading(6),
#  control = list(
#    adapt_delta = 0.99,
#    max_treedepth = 15
#  )
#)

#You’ll get:
#distance–decay slope for within-region pairs
#distance–decay slope for across-region pairs

#If:
#slopes differ → region-specific decay
#slopes similar but intercepts differ → Mississippi as a barrier
#across-region slope = 0 → distance irrelevant once region differs


#model <- brm(sor_Similarity ~ dist_z + same_region + dist_z:same_region + (1 | mm(Sample1, Sample2)),
#             data = melted_diffs, 
             #prior = c(prior(normal(0, 1), class = "b"), prior(normal(0, 1), class = "sd")),

#             warmup = 2000, iter = 6000,
#             chains = 4,
#             cores = 24,
#             threads = threading(6),
#             control = list(adapt_delta = 0.99,max_treedepth = 15))

#summary(model)
```


vignette used for code chunks below: https://bioconductor.org/packages/release/bioc/vignettes/MicrobiotaProcess/inst/doc//MicrobiotaProcess.html

Rare Abundance curve
```{r}
mpse_samples_filtered %<>% mp_rrarefy()

mpse_samples_filtered %<>% 
    mp_cal_rarecurve(
        .abundance = RareAbundance,
        chunks = 400
    )
# default will display the confidence interval around smooth.
# se=TRUE
p1 <- mpse_samples_filtered %>% 
      mp_plot_rarecurve(
         .rare = RareAbundanceRarecurve, 
         .alpha = Observe,
      )

p2 <- mpse_samples_filtered %>% 
      mp_plot_rarecurve(
         .rare = RareAbundanceRarecurve, 
         .alpha = Observe, 
         .group = Region)
      #) +
      #scale_color_manual(values=c("#00A087FF", "#3C5488FF")) +
      #scale_fill_manual(values=c("#00A087FF", "#3C5488FF"), guide="none")

# combine the samples belong to the same groups
p3 <- mpse_samples_filtered %>% 
      mp_plot_rarecurve(
         .rare = RareAbundanceRarecurve, 
         .alpha = "Observe", 
         .group = Region, 
         plot.group = TRUE)
      #) +
       #scale_color_manual(values=c("#00A087FF", "#3C5488FF")) +
       #scale_fill_manual(values=c("#00A087FF", "#3C5488FF"),guide="none")  

p1 
p2
p3
```

calculate alpha index and visualization
```{r}
mpse_samples_filtered %<>% 
    mp_cal_alpha(.abundance=RareAbundance)

f1 <- mpse_samples_filtered %>% 
      mp_plot_alpha(
        .group=Region, 
        .alpha=c(Observe, Chao1, ACE, Shannon, Simpson, Pielou))
      #) +
      #scale_fill_manual(values=c("#00A087FF", "#3C5488FF"), guide="none") +
      #scale_color_manual(values=c("#00A087FF", "#3C5488FF"), guide="none")

f2 <- mpse_samples_filtered %>%
      mp_plot_alpha(
        .alpha=c(Observe, Chao1, ACE, Shannon, Simpson, Pielou)
      )

f1 / f2
```

Beta diversity analysis
```{r}
mpse_samples_filtered %<>% 
    mp_decostand(.abundance=Abundance)

mpse_samples_filtered %<>% mp_cal_dist(.abundance=hellinger, distmethod="bray")

p2 <- mpse_samples_filtered %>% mp_plot_dist(.distmethod = bray, .group = Region)
p2

p3 <- mpse_samples_filtered %>% mp_plot_dist(.distmethod = bray, .group = Region, group.test=T, textsize=2) +
  scale_fill_manual(values = c(
    "#5ab4ac",
    "#8c510a",
    "#d8b365" ))+
  scale_color_manual(values = c(
    "#5ab4ac",
    "#8c510a",
    "#d8b365" ))

 p3 
```

pcoa
```{r}
mpse_samples_filtered %<>% 
    mp_cal_pcoa(.abundance=hellinger, distmethod="bray")

#perform adonis or anosim to check whether it is significant to the dissimilarities of groups.
mpse_samples_filtered %<>%
     mp_adonis(.abundance=hellinger, .formula=~Region, distmethod="bray", permutations=9999, action="add")

mpse_samples_filtered$Region <- 
  factor(
    mpse_samples_filtered$Region,
    levels = c("east", "west")
  )

mpse_samples_filtered %>% mp_extract_internal_attr(name=adonis)

p1 <- mpse_samples_filtered %>%
        mp_plot_ord(
          .ord = pcoa, 
          .group = Region, 
          .color = Region, 
          .starshape = Region,
          .size = 3,
          .alpha = 1,
          ellipse = TRUE,
          show.legend = F # don't display the legend of stat_ellipse
        )+   # land-use colors (primary)
  scale_color_manual(
    name   = "Region",
    values = east_west_colors
  ) +
  scale_fill_manual(
    name   = "Region",
    values = east_west_colors
  ) 

p1

# The size of point also can be mapped to other variables such as Observe, or Shannon 
# Then the alpha diversity and beta diversity will be displayed simultaneously.
p2 <- mpse_samples_filtered %>% 
        mp_plot_ord(
          .ord = pcoa, 
          .group = Region, 
          .color = Region, 
          .size = Observe, 
          .alpha = Shannon,
          .starshape = Region,
          ellipse = TRUE,
          show.legend = FALSE # don't display the legend of stat_ellipse 
        ) +
        scale_color_manual(
    name   = "Region",
    values = east_west_colors
  ) +
  scale_fill_manual(
    name   = "Region",
    values = east_west_colors
  ) +
        scale_size_continuous(
          range=c(0.5, 3),
          guide = guide_legend(keywidth=0.6, keyheight=0.6, label.theme=element_text(size=6.5))
        )
p1 
p2
```

below this is whole microbiome community
___________________________________________________________________________
physical distance matrix using GPS
```{r}
gps <- as.data.frame(mpse_samples_filtered@colData)[, c("labtracking","stateProvince","land_use","decimalLatitude","decimalLongitude","rain","temp","Region")]

locations <- unique(gps$stateProvince)

####################
# Physical distance#
####################
sf_points <- st_as_sf(
  gps,
  coords = c("decimalLongitude", "decimalLatitude"),
  crs = 4326  # WGS84
)

dist_matrix <- st_distance(sf_points)

# Set row/column names
rownames(dist_matrix) <- gps$labtracking
colnames(dist_matrix) <- gps$labtracking

dist_object <- as.dist(dist_matrix)

# Perform hierarchical clustering 
hc_tree <- hclust(dist_object, method = "complete") 
hc_phylo <- as.phylo(hc_tree)

# Create the expanded tree
expanded_tree <- hc_phylo
for(stateProvince in hc_phylo$tip.label) {
  # Find all samples for this stateProvince
  samples <- gps$labtracking[gps$stateProvince == stateProvince]
  
  if(length(samples) > 0) {
    # Find the tip position for this stateProvince
    tip_position <- which(expanded_tree$tip.label == stateProvince)
    
    if(length(samples) == 1) {
      # If only one sample, just rename the tip
      expanded_tree$tip.label[tip_position] <- samples[1]
    } else {
      # If multiple samples, we need to add them as a polytomy
      # First, rename the original tip to the first sample
      expanded_tree$tip.label[tip_position] <- samples[1]
      
      # Then add the rest as sister tips (with very small branch length)
      for(i in 2:length(samples)) {
        expanded_tree <- phytools::bind.tip(expanded_tree, 
                                            tip.label = samples[i],
                                            where = tip_position,
                                            position = 0.0001)  # tiny branch length
      }
    }
  }
}

plot(expanded_tree)

# Match tip labels to rows in gps
tip_index <- match(expanded_tree$tip.label, gps$labtracking)

# Get habitats for all tips (NA if no match)
tip_habitats <- as.character(gps$Region[tip_index])

# Look up colors in habitat_colors
tip_colors <- east_west_colors[tip_habitats]

# Replace any missing colors with black
tip_colors[is.na(tip_colors)] <- "black"

#build tip color vector from your state/colors (must be same length as nrow(assoc))
gps$tip_colors <- east_west_colors[
  match(gps$land_use, names(east_west_colors))
]
names(gps$tip_colors) <- gps$labtracking

phy_tree <- as.phylo(expanded_tree)

# Plot the tree with colored tips
plot(phy_tree, 
     tip.color = tip_colors,
     cex = 0.6,
     no.margin = TRUE)


```

mantel test
DB need to readjust the distance decay to include the east west divide not just distance
```{r}
####################
# Physical distance#
####################
sor_mat <- as.matrix(mpse_sor_dist)[labs, labs]
sor_mat <- sor_mat[rownames(dist_object), colnames(dist_object)]

unique(rownames(dist_object) == rownames(sor_mat)) # ok 
unique(colnames(dist_object) == colnames(sor_mat)) # ok

mt = mantel(sor_mat,dist_object, method = "spearman")
mt$statistic
mt$signif

dist_object_no_zero = dist_object
mt = cultevo::mantel.test(dist(sor_mat),dist(dist_object_no_zero), omitzerodistances = TRUE, trials = 999)
mt

#### plot the mantel tests 
# Convert to long data frames for plotting
df_with_zeros <- data.frame(
  x = as.vector(dist_object),
  y = as.vector(mpse_sor_dist)
)

df_no_zeros <- data.frame(
  x = as.vector(dist_object_no_zero),
  y = as.vector(mpse_sor_dist)
)
df_no_zeros <- df_no_zeros[df_no_zeros$x > 0,]

# Plot with zeros
p1 <- ggplot(df_with_zeros, aes(x = x, y = y)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", col = "blue") +
  labs(
    x = "physical distance",
    y = "Microbiome distance")+
  theme_minimal()

# Plot without zeros
p2 <- ggplot(df_no_zeros, aes(x = x, y = y)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", col = "#8c510a") +
  labs(
    x = "Geograohic distance",
    y = "Microbiome distance"
  ) +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

#### Plot the tanglegram with the expanded tree --------------------------------
note: this phylosymbosis
```{r}
# Convert hclust to phylo
hc <- hclust(mpse_sor_dist, method = "average")
tree <- as.phylo(hc)

plot(tree,
     tip.color = tip_colors,
     cex = 0.6,
     no.margin = TRUE,
     use.edge.length = FALSE)


assoc <- as.matrix(data.frame(
  V1 = as.character(gps$labtracking),   # labels for tree_exp (tree1)
  V2 = as.character(gps$labtracking),   # labels for tree   (tree2)
  stringsAsFactors = FALSE
))

# 3) build link color vector from your state/colors (must be same length as nrow(assoc))
gps$link_cols <- east_west_colors[ match(gps$Region, names(east_west_colors)) ]
link_cols <- as.character(gps$link_cols[ match(assoc[,1], gps$labtracking) ])

tree2 <- tree
tree2$edge.length <- NULL

### reate cophylo - object for tanglegram --------------------------------------
cophylo_obj <- cophylo(tree2, phy_tree,assoc)

plot(cophylo_obj,
     type = "phylogram",
     link.type = "curved",
     link.lty = "solid",
     link.lwd = 1.5,
     link.col = link_cols,
     ftype = "off",
     #cex = 0.6,
     mar = c(0.1, 0.1, 0.1, 0.1))

# assuming state_colors is defined as you pasted
df_legend <- data.frame(Region = names(east_west_colors), color = unname(east_west_colors))
df_legend$Region <- factor(df_legend$Region, levels = names(east_west_colors))

ggplot(df_legend, aes(x = 1, y = Region, fill = Region)) +
  geom_tile(color = "white", height = 0.9, width = 0.6) +
  scale_fill_manual(values = east_west_colors) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  theme_minimal(base_size = 10) +
  theme(axis.title = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none") +
  labs(title = "Region")
```

purty map with bac tree distance 
```{r}
library(mapdata)

# Prepare map data
smoll_map <- gps[, c("decimalLatitude", "decimalLongitude")]
smoll_map$decimalLatitude <- as.numeric(smoll_map$decimalLatitude)
smoll_map$decimalLongitude <- as.numeric(smoll_map$decimalLongitude)
smoll_map$link_cols <- as.factor(gps$link_cols)

# Create a color palette for tips
palette <- rainbow(length(levels(smoll_map$link_cols)))
tip_colors <- palette[as.numeric(smoll_map$link_cols[match(tree$tip.label, gps$SampleID)])]

# Set edge lengths to 1 for plotting
tree_clado <- tree
tree_clado$edge.length <- rep(1, length(tree$edge.length))



# Convert phylo tree to map object
phymap <- phylo.to.map(
  tree_clado,
  smoll_map,
  plot = FALSE,
  database = "worldHires",
  regions = "USA"
)

cols<-setNames(sample(rainbow(n=Ntip(tree_clado))),
    tree_clado$tip.label)

# Plot the map
plot(phymap,
     ftype = "off",
     fsize = 0.8,
     lty = "solid",
     colors=cols,
     map.bg = "grey",
     #colors = "#8c510a",
     xlim = c(-130, -65),
     ylim = c(24, 50))
```

codiversity attempt 
```{r}
#---------------------------
# Taxonomic aggregation
#---------------------------
ps_Family <- tax_glom(ps, taxrank = "Genus", NArm = TRUE)

#---------------------------
# Clean taxa & samples
#---------------------------
ps_clean <- ps_Family %>%
  prune_taxa(taxa_sums(.) > 0, .) %>%
  prune_samples(sample_sums(.) > 0, .)

taxa_names(ps_clean) <- make.unique(taxa_names(ps_clean))


#---------------------------
# Prevalence filtering (≥5%)
#---------------------------
asv_mat <- as(otu_table(ps_clean), "matrix")
prev <- rowSums(asv_mat > 0)
keep <- prev >= 0.05 * ncol(asv_mat)
ps_clean <- prune_taxa(keep, ps_clean)

#---------------------------
# Split asvs by Phylum
#---------------------------
tax_df <- as(tax_table(ps_clean), "matrix") %>%
  as.data.frame() %>%
  rownames_to_column("asv") %>%
  select(asv, Phylum)

asv_by_phylum <- split(tax_df$asv, tax_df$Phylum)

#---------------------------
# Split asvs by Family
#---------------------------
#tax_df <- as(tax_table(ps_clean), "matrix") %>%
#  as.data.frame() %>%
#  rownames_to_column("asv") %>%
#  select(asv, Family)

#asv_by_Family <- split(tax_df$asv, tax_df$Family)

#---------------------------
# Build Left Trees (per phylum)
#---------------------------
scale_left <- 1.5

left_tree_plot_df <- map_dfr(names(asv_by_phylum), function(ph) {
  asvs <- asv_by_phylum[[ph]]
  ps_sub <- prune_taxa(asvs, ps_clean)
  
  if (ntaxa(ps_sub) < 2) return(NULL) # skip empty / singleton trees
  
  p <- ggtree(phy_tree(ps_sub)) + theme_tree2()
  
  p$data %>%
    mutate(
      Phylum = ph,
      x = x * scale_left  # scale left tree
    )
})


# Only keep phyla with trees
valid_phyla <- unique(left_tree_plot_df$Phylum)

left_tree_plot_df <- left_tree_plot_df %>%
  filter(Phylum %in% valid_phyla)

left_tip_df <- left_tree_plot_df %>%
  filter(isTip) %>%
  transmute(
    asv = label,
    x_left = x,
    y_left = y,
    Phylum
  )


#---------------------------
# Build Left Trees (per Family)
#---------------------------
#scale_left <- 1.5

#left_tree_plot_df <- purrr::map_dfr(names(asv_by_Family), function(gen) {
#  asvs <- asv_by_Family[[gen]]
#  ps_sub <- prune_taxa(asvs, ps_clean)

#  if (ntaxa(ps_sub) < 2) return(NULL)

#  p <- ggtree(phy_tree(ps_sub)) + theme_tree2()

#  p$data %>%
#    mutate(
#      Family = gen,
#      x = x * scale_left
#    )
#})

#valid_genera <- unique(left_tree_plot_df$Family)

#left_tree_plot_df <- left_tree_plot_df %>%
#  filter(Family %in% valid_genera)

#left_tip_df <- left_tree_plot_df %>%
#  filter(isTip) %>%
#  transmute(
#    asv = label,
#    x_left = x,
#    y_left = y,
#    Family
#  )


#---------------------------
# Build Right Tree (all samples)
#---------------------------
scale_right <- 1.0 # right tree width

p_right <- ggtree(phy_tree) +
  scale_x_reverse() +
  theme_tree2()

right_df <- p_right$data %>%
  filter(isTip) %>%
  transmute(
    labtracking = label,
    x_right = x,
    y_right = y
  )

right_tree_df <- p_right$data
max_x <- max(right_tree_df$x, na.rm = TRUE)
# flip tips to face left
right_tree_df <- right_tree_df %>%
  mutate(
    x = max(x, na.rm = TRUE) - x # mirror
  )

# normalize and scale
right_tree_df <- right_tree_df %>%
  mutate(
    x = (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE)),
    x = x * scale_right
  )

right_tip_df <- right_tree_df %>%
  filter(isTip) %>%
  transmute(
    labtracking = label,
    x_base = x,
    y_base = y
  )

#---------------------------
# Shift right trees per phylum
#---------------------------
max_left_x_df <- left_tree_plot_df %>%
  group_by(Phylum) %>%
  summarize(max_left_x = max(x, na.rm = TRUE), .groups = "drop")

# replicate right tree for each valid phylum
right_tree_df <- right_tree_df %>%
  crossing(Phylum = valid_phyla) %>%
  left_join(max_left_x_df, by = "Phylum") %>%
  mutate(x = x + max_left_x + 1) # shift right tree after left tree

right_tip_df <- right_tip_df %>%
  crossing(Phylum = valid_phyla) %>%
  left_join(max_left_x_df, by = "Phylum") %>%
  mutate(x_right = x_base + max_left_x + 1,
         y_right = y_base) %>%
  select(labtracking, x_right, y_right, Phylum)

#---------------------------
# Shift right trees per Family
#---------------------------
#max_left_x_df <- left_tree_plot_df %>%
#  group_by(Family) %>%
#  summarize(max_left_x = max(x, na.rm = TRUE), .groups = "drop")

#right_tree_df <- right_tree_df %>%
#  crossing(Family = valid_genera) %>%
#  left_join(max_left_x_df, by = "Family") %>%
#  mutate(x = x + max_left_x + 1)

#right_tip_df <- right_tip_df %>%
#  crossing(Family = valid_genera) %>%
#  left_join(max_left_x_df, by = "Family") %>%
#  mutate(
#    x_right = x_base + max_left_x + 1,
#    y_right = y_base
#  ) %>%
#  select(labtracking, x_right, y_right, Family)

#---------------------------
# Build links
#---------------------------
asv_mat <- as(otu_table(ps_clean), "matrix")

links <- asv_mat %>%
  as.data.frame() %>%
  rownames_to_column("asv") %>%
  pivot_longer(
    -asv,
    names_to = "labtracking",
    values_to = "Abundance"
  ) %>%
  filter(Abundance > 0) %>%
  left_join(tax_df, by = "asv") %>%
  #filter(Phylum %in% valid_genera) %>% # only valid phyla
  left_join(as(sample_data(ps_clean), "data.frame"), by = "labtracking") %>%
  mutate(
    w = scales::rescale(log10(Abundance + 1), c(0.2, 1))
  )

links_plot <- links %>%
  left_join(left_tip_df,  by = c("asv", "Phylum")) %>%
  left_join(right_tip_df, by = c("labtracking", "Phylum")) %>%
  filter(
    !is.na(x_left) & !is.na(y_left) &
    !is.na(x_right) & !is.na(y_right)
  ) %>%
  mutate(
    x_right = ifelse(x_left == x_right, x_right + 1e-6, x_right),
    y_right = ifelse(y_left == y_right, y_right + 1e-6, y_right)
  )



#---------------------------
# Plot
#---------------------------
ggplot() +
  # LEFT trees
  geom_tree(
    data = left_tree_plot_df,
    aes(x = x, y = y),
    inherit.aes = FALSE
  ) +
  # RIGHT trees
  geom_tree(
    data = right_tree_df,
    aes(x = x, y = y),
    inherit.aes = FALSE
  ) +
  # LINKS
  geom_curve(
    data = links_plot,
    aes(
      x = x_left,
      y = y_left,
      xend = x_right,
      yend = y_right,
      color = Region,
      linewidth = w
    ),
    curvature = 0.1,
    alpha = 0.3
  ) +
  scale_color_brewer(palette = "Set2", name = "Region") +
  scale_linewidth(range = c(0.2, 1), guide = "none") +
  facet_wrap(~Phylum, scales = "free_y") +
  theme_void() +
  coord_cartesian(clip = "off")

```

top 20 
```{r}
colnames(tax_table(ps))
ps_genus <- tax_glom(ps, taxrank = "Genus")

# Remove uncultured / unclassified / unidentified taxa
ps_genus <- subset_taxa(
  ps_genus,
  !grepl(
    "uncultured|unclassified|unidentified|unknown",
    Genus,
    ignore.case = TRUE
  )
)

ps_genus_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))
top20 <- names(sort(taxa_sums(ps_genus_rel), decreasing = TRUE))[1:30]
ps_top20 <- prune_taxa(top20, ps_genus_rel)
```

top 20 PCoA with individuals as points
```{r}
# Step 1: Get all Families 
families <- unique(as.vector(tax_table(ps_top20)[, "Genus"]))
families <- families[!is.na(families)]


# Step 2: Prepare containers
pcoa_list   <- list()
axis_labels <- list()

for (fam in families) {

  # Subset by Family
  ps_sub <- subset_taxa(ps, Genus == fam)

  # Remove empty samples
  ps_sub <- prune_samples(sample_sums(ps_sub) > 0, ps_sub)

  # Distance matrix
  dist_mat <- tryCatch(
    phyloseq::distance(ps_sub, method = "bray"),
    error = function(e) NULL
  )
  if (is.null(dist_mat)) next

  # PCoA
  pcoa_res <- tryCatch(
    ape::pcoa(dist_mat),
    error = function(e) NULL
  )
  if (is.null(pcoa_res)) next

  # Variance explained (guard against NA)
  var_exp <- round(
    100 * pcoa_res$values$Relative_eig[1:2],
    1
  )
  if (any(is.na(var_exp))) next

  # Sample metadata
  sample_df <- data.frame(sample_data(ps_sub))

  # PCoA dataframe
  pcoa_df <- data.frame(
    Sample = rownames(sample_df),
    PC1    = pcoa_res$vectors[, 1],
    PC2    = pcoa_res$vectors[, 2],
    land_use = sample_df$land_use,
    Region = sample_df$Region,
    Genus = fam
  )

  # Store facet title
  axis_labels[[fam]] <- data.frame(
    Genus = fam,
    title  = paste0(
      fam,
      "\nPC1 (", var_exp[1], "%) | PC2 (", var_exp[2], "%)"
    )
  )

  pcoa_list[[fam]] <- pcoa_df
}

# Combine results
combined_pcoa <- bind_rows(pcoa_list)
axis_labels   <- bind_rows(axis_labels)

combined_pcoa <- combined_pcoa %>%
  left_join(axis_labels, by = "Genus") %>%
  mutate(title = factor(title, levels = unique(title)))

# Step 3: Plot
ggplot(
  combined_pcoa,
  aes(PC1, PC2, color = Region)
) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(
    aes(group = Region),
    level = 0.95,
    linetype = 2,
    na.rm = TRUE
  ) +
  facet_wrap(~ title, scales = "free") +
  scale_color_manual(values = east_west_colors) +
  theme_bw()

```

with abundance 
```{r}
# Step 1: Get all Families 
families <- unique(as.vector(tax_table(ps_top20)[, "Species"]))
families <- families[!is.na(families)]

# Step 2: Prepare containers
pcoa_list   <- list()
axis_labels <- list()

for (fam in families) {

  # Subset by Species
  ps_sub <- subset_taxa(ps, Species == fam)

  # Remove empty samples
  ps_sub <- prune_samples(sample_sums(ps_sub) > 0, ps_sub)

  # Skip if too few samples
 if (length(sample_names(ps_sub)) < 3) next

  # -------------------------
  # Get abundance per sample
  # -------------------------
  abundance <- sample_sums(ps_sub)

  # Distance matrix
  dist_mat <- tryCatch(
    phyloseq::distance(ps_sub, method = "bray"),
    error = function(e) NULL
  )
  if (is.null(dist_mat)) next

  # PCoA
  pcoa_res <- tryCatch(
    ape::pcoa(dist_mat),
    error = function(e) NULL
  )
  if (is.null(pcoa_res)) next

  # Variance explained
  var_exp <- round(
    100 * pcoa_res$values$Relative_eig[1:2],
    1
  )
  if (any(is.na(var_exp))) next

  # Sample metadata
  sample_df <- data.frame(sample_data(ps_sub))

  # PCoA dataframe (ADD abundance)
  pcoa_df <- data.frame(
    Sample    = rownames(sample_df),
    PC1       = pcoa_res$vectors[, 1],
    PC2       = pcoa_res$vectors[, 2],
    Region    = sample_df$Region,
    Abundance = abundance[rownames(sample_df)],
    Family    = fam
  )

  # Store facet title
  axis_labels[[fam]] <- data.frame(
    Family = fam,
    title  = paste0(
      fam,
      "\nPC1 (", var_exp[1], "%) | PC2 (", var_exp[2], "%)"
    )
  )

  pcoa_list[[fam]] <- pcoa_df
}

# Combine results
combined_pcoa <- bind_rows(pcoa_list)
axis_labels   <- bind_rows(axis_labels)

combined_pcoa <- combined_pcoa %>%
  left_join(axis_labels, by = "Family") %>%
  mutate(title = factor(title, levels = unique(title))) 

combined_pcoa <- combined_pcoa %>%
  mutate(Abundance_log = log10(Abundance + 1))

# Step 3: Plot
ggplot(
  combined_pcoa,
  aes(PC1, PC2,
      color = Region,
      shape = Region,
      size  = Abundance)
) +
  geom_point(alpha = 0.75) +
  stat_ellipse(
    aes(group = Region),
    level = 0.95,
    linetype = 2,
    linewidth = 0.6,
    na.rm = TRUE
  ) +
  facet_wrap(~ title, scales = "free") +
  scale_color_manual(values = east_west_colors) +
  
  # Cleaner size scaling
  scale_size_continuous(
    range = c(2, 7),
    breaks = pretty(combined_pcoa$Abundance, n = 4),
    name = "Abundance"
  ) +
  
  guides(
    size = guide_legend(
      override.aes = list(alpha = 1)
    )
  ) +
  
  theme_bw(base_size = 12) +
  theme(
    legend.position = "right",
    strip.text = element_text(face = "bold")
  )

```


top 20 families distance decay plots
```{r}
# -----------------------------
# Extract OTU table
# -----------------------------
otu_mat <- as(otu_table(ps_top20), "matrix")
if (taxa_are_rows(ps_top20)) {
  otu_mat <- t(otu_mat)
}

# -----------------------------
# Bray–Curtis similarity (%)
# -----------------------------
bc_dist <- vegdist(otu_mat, method = "bray",binary = T)
bc_sim  <- (1 - as.matrix(bc_dist)) * 100

# -----------------------------
# Convert distances to long format
# -----------------------------
bc_long <- as.data.frame(as.table(bc_sim))
colnames(bc_long) <- c("sample1", "sample2", "percent_shared")

geo_long <- as.data.frame(as.table(geo_dist))
colnames(geo_long) <- c("sample1", "sample2", "geo_km")

df <- bc_long %>%
  left_join(geo_long, by = c("sample1", "sample2")) %>%
  filter(sample1 != sample2)

# -----------------------------
# Add sample metadata
# -----------------------------
#meta_lookup <- meta %>%
#  rownames_to_column("sample1") %>%
 # select(sample1, Region)

df <- df %>%
  left_join(meta_lookup, by = "sample1")

# -----------------------------
# Add phylum info for ALL ASVs
# -----------------------------
tax <- as.data.frame(tax_table(ps_top20))

otu_pa <- as.data.frame(otu_mat)
otu_pa$sample1 <- rownames(otu_pa)

# Expand OTUs to long format
otu_long <- otu_pa %>%
  pivot_longer(-sample1, names_to = "ASV", values_to = "present") %>%
  filter(present == 1) %>%
  left_join(
    tax %>% rownames_to_column("ASV") %>% select(ASV, Species),
    by = "ASV"
  )

# Join phylum info to df: now each row can repeat for multiple phyla
df <- df %>%
  left_join(
    otu_long %>% select(sample1, Species) %>% distinct(),
    by = "sample1"
  )

# -----------------------------
# Plot distance–decay faceted by phylum
# -----------------------------
ggplot(df, aes(x = geo_km, y = percent_shared,
               color = Region,
               shape = Region)) +
  geom_point(alpha = 0.1, size = 2) +
  scale_color_manual(values = east_west_colors) +
  geom_smooth(aes(group = Region),
              method = "loess",
              se = FALSE) +
  facet_wrap(~ Species, scales = "free_y") +
  labs(
    x = "Geographic distance (km)",
    y = "Bray–Curtis similarity (%)",
    color = "Region",
    shape = "Region",
    title = "Distance–decay of ASV similarity by Species"
  ) +
  theme_bw()
```


MaAsLin (Microbiome Multivariable Associations with Linear Models)
https://github.com/biobakery/biobakery/wiki/maaslin3

do this at the family and or spices level
```{r}
install_github("biobakery/Maaslin2")
library(Maaslin2)

ps_genus <- tax_glom(ps, taxrank = "Genus")
otu <-as(otu_table(ps_genus), "matrix")


# Factor the categorical variables to test
meta$Region <- factor(meta$Region, levels = c('west', 'east'))
meta$sex <- factor(meta$sex, levels = c('M', 'F'))
meta$land_use <- factor(meta$land_use, levels = c('Grassland/Herbaceous',
                                                  'Evergreen Forest',
                                                  "Shrub/Scrub",
                                                  "Cultivated Crops",
                                                  "Deciduous Forest",
                                                  "Mixed Forest"))

set.seed(1)
fit_out <- maaslin3(input_data = t(otu),
                    input_metadata = meta,
                    output = 'm2_output',
                    formula = '~ Region',
                    normalization = 'TSS',
                    transform = 'LOG',
                    augment = TRUE,
                    standardize = TRUE,
                    plot_summary_plot = TRUE,
                    plot_associations = TRUE,
                    max_significance = 0.1,
                    median_comparison_abundance = TRUE,
                    median_comparison_prevalence = F,
                    max_pngs = 250,
                    cores = 1)


```